---
title: "Migrating the posterior workflow from brms"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Migrating the posterior workflow from brms}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
# the calls need tmbstan and a NUTS run, so the page shows code and not
# output
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`vignette("brms-migration", package = "frmtmb")` ports the model
GRAMMAR: what `bf()`, the addition terms, the group-level structures
and the families do here, and how a `brm()` call becomes a `frm()`
call. None of that changes when you sample. `frm_sample()` takes the
same `bf()` object, assembles it through the same code path, and
refuses what `frm()` refuses.

This page is the other half: what happens to a brms POSTERIOR workflow.
The draws come from Stan's NUTS sampler, but through tmbstan on an RTMB
objective rather than a generated Stan program, and that one difference
explains every entry below.

## The call

```{r}
library(frmtmb)
library(frmtmb.sample)

# brms
brm(y ~ x + (1 | g), data = dd, family = gaussian(),
    chains = 4, iter = 2000, cores = 4)

# here
frm_sample(bf(y ~ x + (1 | g)) + gaussian(), data = dd,
           chains = 4, iter = 2000, cores = 4)
```

| brms | here |
|---|---|
| `brm(formula, data, family)` | `frm_sample(formula, data =, family =)`, or `frm_sample(fit)` on a fitted model |
| `chains`, `iter`, `warmup`, `thin`, `cores`, `seed`, `refresh` | the same names, passed straight through to `tmbstan::tmbstan()` |
| `prior = prior(...)` | `prior = set_prior(...)`; a `brmsprior` that brms itself built is translated row by row |
| the default prior set | applied on both routes, read off `brms::default_prior()`; `prior = "flat"` opts out |
| `init =`, `init_r =` | `init =`, `init_jitter =`; from a fit the maximum-likelihood mode anchors the chains |
| `control = list(adapt_delta = )` | not through `frm_sample()`, whose `control` is `frmtmb_control()`; see below |
| `backend =`, `threads =`, `stan_model_args =` | nothing to configure: there is no Stan program to compile |
| `sample_prior = "only"` | `frmtmb::frm_simulate()`, which draws a parameter vector per simulation and returns it with the responses |

The one argument with no route is rstan's `control` list. `frm_sample()`
already spells `control` as `frmtmb_control()`, the model-assembly
options, so `adapt_delta` and `max_treedepth` cannot travel past it.
`as_tmbstan()` passes everything to tmbstan and takes them:

```{r}
sf <- as_tmbstan(fit, chains = 4, control = list(adapt_delta = 0.99))
```

What that costs is the return value. `as_tmbstan()` gives back a
`stanfit`, whose parameters carry Stan's names and none of the method
surface below.

## Two routes, and which one to take

`frm_sample()` takes a fit or a formula, and both sample the same
posterior under the same priors. Coming from brms, the formula route is
the familiar one: it is `brm()` with a different function name. The fit
route is the one brms has no counterpart to.

```{r}
fit <- frm(bf(y ~ x + (1 | g)) + gaussian(), data = dd)
ds <- frm_sample(fit, chains = 4)
```

Fitting first is worth the seconds it takes. The optimizer either finds
the mode or reports why it could not, `diagnose(fit)` says whether the
model is identified before any chain runs, and the mode then anchors
the chains and shortens warmup. A model that will not converge under
`frm()` is a model whose posterior geometry is about to be difficult,
and finding that out in a second is better than finding it out in an
hour. `vignette("sampling")` has the detail on both routes, the default
priors and the non-centered parameterization.

## The names change

Parameter names drop parentheses on the draws side: `Intercept`, not
`(Intercept)`. That is the vocabulary posterior, bayesplot and
`hypothesis()` already speak, so a ported `hypothesis()` string usually
needs no edit.

```{r}
variables(ds)
hypothesis(ds, "sd_g__Intercept^2 / (sd_g__Intercept^2 + sigma^2)")
```

`rhat()`, `neff_ratio()` and `nuts_params()` are the exception. They
read the `stanfit` directly and report Stan's own names.

## The method surface ports

The post-processing generics are brms's own, on a `frmtmb_draws`
object, so most ported code runs unchanged: `fixef()`, `ranef()`,
`VarCorr()`, `posterior_epred()`, `posterior_predict()`,
`posterior_linpred()`, `log_lik()`, `loo()`, `waic()`, `bayes_R2()`,
`hypothesis()`, `conditional_effects()`, `pp_check()`, `mcmc_plot()`,
`prior_summary()`, `as_draws()` and the `posterior` and `coda`
conversions.

`loo()` here is the real leave-one-out, not the `AIC()` substitute core
offers, and it returns the loo package's own object.

## What refuses, and what to write instead

Every method below is defined, and every one stops with its reason and
its replacement, so a ported script fails where it went wrong rather
than at "could not find function".

| brms call | why it stops | write instead |
|---|---|---|
| `stancode(ds)` | there is no Stan program; the R closure is the source | `ds$fit$obj$fn`, `ds$fit$frame` |
| `standata(ds)` | nothing is exported to a Stan data list | `ds$fit$frame`, `model.matrix()`, `getME()` |
| `expose_functions(ds)` | a custom family is already plain R | call the `lpdf` you passed to `custom_family()` |
| `plot(ds)` | brms's default panel is trace and density | `mcmc_plot(ds)`, `mcmc_plot(ds, type = "trace")` |
| `update(ds)` | draws carry no formula to revise | `update()` the fit, then `frm_sample()` it |
| `restructure(ds)` | no older-object upgrade path exists | re-run `frm_sample()` |
| `posterior_samples(ds)` | deprecated brms spelling | `as_draws(ds)`, `as.matrix(ds)` |
| `nsamples(ds)` | deprecated brms spelling | `ndraws(ds)`, `niterations(ds)` |
| `parnames(ds)` | deprecated brms spelling | `variables(ds)` |

Two groups refuse for reasons that are not spelling.

**The refit-based loo remedies.** `reloo()`, `kfold()`,
`loo_moment_match()` and `loo_subsample()` each need to refit the model
on modified data, and that machinery is not here. When `loo()` reports
high Pareto k values, sample under the default priors rather than
`prior = "flat"` and read the number again; if the warning survives,
compare the maximum-likelihood fits by `AIC()`, or measure the
influence with `frmtmb::frm_bootstrap()`.

**The marginal-likelihood family.** `bridge_sampler()`,
`bayes_factor()` and `post_prob()` are integrals against the prior. They
are undefined under `prior = "flat"`, and even under the default priors
bridge sampling needs a normalized log-posterior evaluator that the
RTMB tape does not expose. This is the one brms capability with no
route here at all.

## When you still want brms

- A Bayes factor or any other marginal-likelihood quantity.
- `reloo()` or `kfold()` on a model whose LOO approximation fails.
- A family, a term or an addition term frmtmb does not implement.
  `frmtmb::frm_compat()` answers that per feature, and the answer is
  the same for `frm()` and `frm_sample()`.
- Stan code you intend to read, extend or hand to somebody else.

A practical division: screen models with `frm()`, which costs seconds,
sample the survivors with `frm_sample()`, and keep brms for the
questions in that list.
