---
title: "Hidden Markov models and latent class analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hidden Markov models and latent class analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb.latent)
has_tinyplot <- requireNamespace("tinyplot", quietly = TRUE)
has_depmix <- requireNamespace("depmixS4", quietly = TRUE)
has_hmmtmb <- requireNamespace("hmmTMB", quietly = TRUE)
has_polca <- requireNamespace("poLCA", quietly = TRUE)
```

This package adds the two frmtmb families whose latent variable is a
discrete state. `hmm(K, family)` puts a Markov chain over the rows of a
sequence and sums the state path out exactly. `lca(K)` puts one class
on each subject and reads it through conditionally independent items.
Both are written against the structured-family protocol that frmtmb
exports, so the rest of the grammar (priors, monotonic effects,
`hypothesis()`, and the addition terms each family accepts) applies to
them unchanged.

The animal-track study below came from frmtmb's own case-studies
vignette when `hmm()` moved out here. It checks the same numbers
against the same two reference packages, and the checks fail the build
if a number moves.

## A hidden Markov model for an animal track

An animal alternates between behavioral modes. While it forages it
takes short, undirected steps; while it travels it takes long ones. The
mode is not observed, but it persists, so the movement package
literature models it as a Markov chain over the steps and the step
length as a state-dependent distribution (Michelot, Langrock and
Patterson 2016, *Methods in Ecology and Evolution* 7, the moveHMM
paper; Michelot 2022, the hmmTMB paper; Zucchini, MacDonald and
Langrock 2016). `hmm(K, family)` writes that model as a family, so the
state sequence is summed out exactly by the forward algorithm and every
other part of the grammar still applies.

The simulated track below has 12 animals, 60 steps each, and two
states with well separated log step lengths.

```{r hmm-data}
set.seed(2001)
G <- matrix(c(0.92, 0.08, 0.18, 0.82), 2, 2, byrow = TRUE)
one_track <- function(id, nt = 60L) {
  s <- integer(nt); s[1] <- 1L
  for (t in 2:nt) s[t] <- sample.int(2L, 1L, prob = G[s[t - 1L], ])
  data.frame(track = id, step = seq_len(nt), state = s,
             logstep = rnorm(nt, c(1.4, 3.1)[s], c(0.45, 0.55)[s]))
}
mv <- do.call(rbind, lapply(1:12, one_track))
```

`time =` says which column orders the chain and `group =` says which
column cuts it into sequences. Both take a bare name or a one-sided
formula. `init = "stationary"` gives the chain the stationary
distribution of its own transition matrix, which costs no parameters.

```{r hmm-fit}
fhmm <- frm(bf(logstep ~ 1),
            family = hmm(K = 2, gaussian(), time = step, group = track,
                         init = "stationary"),
            data = mv)
e <- unlist(fixef(fhmm))
rbind(estimated = c(exp(e[["sigma1.(Intercept)"]]), e[["mu1.(Intercept)"]],
                    e[["mu2.(Intercept)"]], exp(e[["sigma2.(Intercept)"]])),
      simulated = c(0.45, 1.4, 3.1, 0.55))[, c(2, 1, 3, 4)] |>
  `colnames<-`(c("mean1", "sd1", "mean2", "sd2")) |> round(4)
```

The transition matrix is a row-wise multinomial logit with state 1 as
the reference cell of every row, which is depmixS4's convention. Turn
the two free logits `tr12` and `tr22` back into probabilities by hand.

```{r hmm-tpm}
tpm_row <- function(z) { p <- exp(c(0, z)); p / sum(p) }
Ghat <- rbind(tpm_row(e[["tr12.(Intercept)"]]),
              tpm_row(e[["tr22.(Intercept)"]]))
dimnames(Ghat) <- list(paste0("from", 1:2), paste0("to", 1:2))
round(Ghat, 4)
round(G, 4)
```

`hmm_probs()` gives the smoothed state probability of every row from a
forward-backward pass, and `hmm_viterbi()` gives the most probable
whole path. The two answer different questions: the first is marginal
per row, the second is joint over the sequence.

```{r hmm-decode}
c(occupancy_state1 = mean(hmm_probs(fhmm)[, 1]),
  viterbi_accuracy = mean(hmm_viterbi(fhmm) == mv$state))
```

### Cross-check against hmmTMB and depmixS4

The same model has two reference implementations. hmmTMB fits the
stationary-start version, and depmixS4 fits the free-start version by
expectation maximization. Fit each on the same track and compare the
maximized log likelihood.

```{r hmm-hmmtmb, eval = has_hmmtmb}
dh <- data.frame(ID = mv$track, step = mv$step, logstep = mv$logstep)
hid <- hmmTMB::MarkovChain$new(data = dh, n_states = 2,
                               initial_state = "stationary")
hid$update_tpm(matrix(c(0.9, 0.1, 0.1, 0.9), 2, 2, byrow = TRUE))
obs <- hmmTMB::Observation$new(
  data = dh, n_states = 2, dists = list(logstep = "norm"),
  par = list(logstep = list(mean = c(1.5, 3.0), sd = c(0.5, 0.5))))
hm <- hmmTMB::HMM$new(obs = obs, hid = hid)
hm$fit(silent = TRUE)
c(frmtmb = as.numeric(logLik(fhmm)), hmmTMB = hm$llk(),
  difference = as.numeric(logLik(fhmm)) - hm$llk())
```

```{r hmm-hmmtmb-assert, eval = has_hmmtmb}
stopifnot(abs(as.numeric(logLik(fhmm)) - hm$llk()) < 1e-6,
          max(abs(sort(as.numeric(hm$par()$obspar["logstep.mean", , 1])) -
                    sort(c(e[["mu1.(Intercept)"]],
                           e[["mu2.(Intercept)"]])))) < 1e-4)
```

hmmTMB starts from generic values rather than from the frmtmb answer,
so this is a check of the optimum and not only of the likelihood.

depmixS4 leaves the initial distribution free, so compare it with
`init = "estimated"`. Its expectation-maximization run is
multimodal, so take the best of several random starts.

```{r hmm-depmix, eval = has_depmix}
fest <- frm(bf(logstep ~ 1),
            family = hmm(K = 2, gaussian(), time = step, group = track,
                         init = "estimated"), data = mv)
dm <- depmixS4::depmix(logstep ~ 1, data = mv, nstates = 2,
                       ntimes = as.integer(table(mv$track)))
set.seed(3)
best <- -Inf
sink(nullfile())          # depmixS4 reports every EM run on stdout
for (i in 1:5) {
  ff <- try(suppressMessages(depmixS4::fit(
    dm, verbose = FALSE, emcontrol = depmixS4::em.control(
      random.start = TRUE, tol = 1e-12, maxit = 5000))), silent = TRUE)
  if (!inherits(ff, "try-error")) {
    best <- max(best, as.numeric(depmixS4::logLik(ff)))
  }
}
sink()
c(frmtmb = as.numeric(logLik(fest)), depmixS4 = best,
  difference = as.numeric(logLik(fest)) - best)
```

```{r hmm-depmix-assert, eval = has_depmix}
stopifnot(abs(as.numeric(logLik(fest)) - best) < 1e-6)
```

The three packages agree on the maximum to better than 1e-6. The free
start costs one parameter and buys about 3.7 log-likelihood points on
12 sequences, which is what a free start usually buys when the chain is
sticky.

One track, drawn against time and colored by its decoded state, is the
picture the model is for.

```{r hmm-fig, eval = has_tinyplot, fig.width = 7, fig.height = 3.8, fig.alt = "Plot of log step length against step number for one 60-step track. The points are colored by the decoded state: low points near 1.4 in one color and high points near 3.1 in the other. Two horizontal dashed lines mark the two fitted state means. The colors come in runs rather than alternating, which is the persistence of the Markov chain."}
one <- mv[mv$track == 1, ]
one$decoded <- factor(hmm_viterbi(fhmm)[mv$track == 1],
                      labels = c("state 1", "state 2"))
tinyplot::tinyplot(logstep ~ step | decoded, data = one, type = "p",
                   pch = 16, theme = "clean2", xlab = "step",
                   ylab = "log step length")
abline(h = c(e[["mu1.(Intercept)"]], e[["mu2.(Intercept)"]]), lty = 2,
       col = "gray40")
```

The colors come in runs. That persistence is the whole difference
between an HMM and the two-component growth mixture of
`vignette("case-studies", package = "frmtmb")`, and it is what the
transition matrix measures.

### What the HMM surface refuses

The refusals below are documented rather than worked around. Each one
has a reason that a workaround would hide.

```{r hmm-refusals, error = TRUE}
conditional_effects(fhmm)
residuals(fhmm, type = "deviance")
```

`conditional_effects()` needs an expected response on a synthetic
covariate grid. Under an HMM the expected response weights the state
means by the posterior occupancies, and those depend on the observed
responses of a whole sequence, so they do not exist on a grid. Plot one
state's own predictor with `predict(dpar = "mu2")`, or the occupancies
from `hmm_probs()`.

`residuals(type = "deviance")`, and frmtmb.sample's `log_lik()` and
`loo()`, all need a
likelihood that factors into one term per observation. An HMM's
smallest independent unit is a sequence, so a per-observation column
would be a group and leaving one out would drop a whole track. Compare
HMM fits with `AIC()` or `frm_bootstrap()`.

This section also does not cover the parts of the surface that do work:
transition covariates (`trans = ~x`, or `bf(y ~ 1, tr12 ~ x)` for one
cell), non-gaussian emissions, and random effects inside a state's
linear predictor. The last one is approximate: the Laplace
approximation integrates the random effect outside the exact state sum,
and the integrand is a mixture over state sequences rather than a
gaussian. `?hmm` gives the measured size of that bias.


## Latent class analysis of a rater panel

Seven raters each score the same slides. The scores disagree, and the
question is whether the disagreement has structure: is there a latent
grouping of slides that the raters are noisily reading? That is the
latent class model, and it is `poLCA`'s `carcinoma` example in
miniature.

`lca(K)` takes a matrix response whose columns are the items. Each
subject gets one class; within a class the items are independent, and
each item has its own class-conditional profile over its categories.

```{r lca-data}
set.seed(404)
ns <- 250L
truth <- rbinom(ns, 1L, 0.45) + 1L
# seven raters, each with its own sensitivity and specificity
sens <- c(0.93, 0.88, 0.80, 0.95, 0.72, 0.90, 0.85)
Y <- sapply(sens, function(p)
  ifelse(truth == 2L, rbinom(ns, 1L, p), rbinom(ns, 1L, 1 - p)) + 1L)
colnames(Y) <- LETTERS[1:7]
dc <- data.frame(Y = I(Y))
head(Y, 3)
```

```{r lca-fit}
flca <- frm(bf(Y ~ 1), family = lca(K = 2), data = dc)
flca
```

The item profiles are what the classes mean. Each block is one rater:
the rows are the classes and the columns are that rater's categories,
so the diagonal reads off the rater's accuracy.

```{r lca-profiles}
lca_profiles(flca)
```

`lca_probs()` gives each slide's posterior class membership, and its
`entropy` attribute is the usual separation statistic: 1 is a perfectly
resolved partition, 0 is no information about class at all.

```{r lca-probs}
pp <- lca_probs(flca)
head(round(pp, 3))
attr(pp, "entropy")
# the classes are exchangeable, so score both labelings
max(mean(max.col(pp) == truth), mean(max.col(pp) == 3L - truth))
```

### Cross-check against poLCA

poLCA fits the same model by expectation maximization. `lca()`
maximizes the same likelihood by gradient ascent on the same
parameterization, so the two should reach the same optimum.

```{r lca-polca, eval = has_polca}
dp <- as.data.frame(Y)
set.seed(11)
pl <- poLCA::poLCA(cbind(A, B, C, D, E, F, G) ~ 1, dp, nclass = 2,
                   nrep = 20, verbose = FALSE, maxiter = 20000,
                   tol = 1e-12)
c(frmtmb = as.numeric(logLik(flca)), poLCA = pl$llik,
  difference = as.numeric(logLik(flca)) - pl$llik)
```

```{r lca-polca-assert, eval = has_polca}
stopifnot(abs(as.numeric(logLik(flca)) - pl$llik) < 1e-6)
```

The class labels may come out in either order, so match them before
comparing the item profiles and the posterior memberships.

```{r lca-polca-profiles, eval = has_polca}
ours <- do.call(cbind, lca_profiles(flca))
theirs <- do.call(cbind, pl$probs)
perm <- if (sum(abs(ours - theirs)) <= sum(abs(ours[2:1, ] - theirs))) {
  1:2
} else {
  2:1
}
c(profiles = max(abs(ours[perm, ] - theirs)),
  posterior = max(abs(lca_probs(flca)[, perm] - pl$posterior)))
```

```{r lca-polca-profiles-assert, eval = has_polca}
stopifnot(max(abs(ours[perm, ] - theirs)) < 1e-4,
          max(abs(lca_probs(flca)[, perm] - pl$posterior)) < 1e-4)
```

### What the LCA surface refuses

The response is a matrix of nominal item codes, so there is no fitted
mean and therefore no residual, and the gating predictor takes no
random effects.

```{r lca-refusals, error = TRUE}
fitted(flca)
residuals(flca)
```

Latent class REGRESSION needs no extra machinery. The class-membership
weights are ordinary distributional parameters, so a covariate in the
formula puts a multinomial logit on class membership, and its
coefficients get standard errors, priors and `hypothesis()` like any
other fixed effect.
