---
title: "Bayesian Cognitive Modeling: the model list"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian Cognitive Modeling: the model list}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE, purl = FALSE}
# purl = FALSE: knitr's tangle step extracts a read_chunk() call with a
# regex that stops at the first closing parenthesis, so the nested
# bcm_file() call fails to parse and pkgdown refuses the whole article
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(frmtmb)
bcm_file <- function(f) system.file("bcm", f, package = "frmtmb",
                                    mustWork = TRUE)
for (f in c("binomial-extras.R", "process-trees.R", "similarity.R",
            "marginal.R")) {
  source(bcm_file(f))
}
knitr::read_chunk(bcm_file("marginal.R"))
```

This page is a reference. It lists every model in Lee and Wagenmakers,
*Bayesian Cognitive Modeling: A Practical Course* (2013), gives the
frmtmb spelling of each, says how that spelling was checked, and names
what the check does not cover. It is not a tutorial: read
`vignette("frmtmb")` for the grammar and
`vignette("reinforcement-learning")` for a worked example of writing a
family from scratch.

The reference implementations are the Stan ports at
[stan-dev/example-models](https://github.com/stan-dev/example-models/tree/master/Bayesian_Cognitive_Modeling),
which are BSD-3 licensed. The models themselves are the book's
published equations. Every Stan program in the test suite is adapted
from those ports, with the adaptations stated at the head of the file
that uses it.

The families the book needs and frmtmb does not ship are in
`inst/bcm/`, the way `vignette("reinforcement-learning")`'s family is
in `inst/rl/`. Load them with

```r
source(system.file("bcm", "binomial-extras.R", package = "frmtmb"))
source(system.file("bcm", "process-trees.R", package = "frmtmb"))
source(system.file("bcm", "similarity.R", package = "frmtmb"))
source(system.file("bcm", "marginal.R", package = "frmtmb"))
```

## How a row of the table was checked

Three kinds of evidence appear in the Validation column.

**identity.** Stan's `log_prob` at frmtmb's estimate equals frmtmb's
own log density there, plus a stated constant, and the gradient
vanishes where it should. `tests/testthat/helper-stan.R` is the
harness; `tests/testthat/test-bcm-*.R` are the callers. The constant is
a property of the pair of programs and never a number read off a
failing run:

* the Stan program carries a prior frmtmb does not, and the constant is
  that prior's log density at the estimate;
* frmtmb carries the same prior through `set_prior()`, and the constant
  is zero;
* neither carries a prior, and the constant is zero.

Every reference program is adapted in one way throughout: a sampling
statement `y ~ dist(...)` becomes `target += dist_lpdf(y | ...)`,
because Stan drops the normalizing constant of a `~` statement and
frmtmb's families do not.

**recovery.** The model is fitted to data generated from it and the
estimates are compared with the values that generated them. Used where
an identity at the estimate is not meaningful, and always with a stated
reason.

**structure.** The fit reproduces a quantity the chapter reports
directly: Cohen's kappa, a sample correlation, a classification.

## Chapter 3, inferring a rate

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Rate_1 | `frm(k | trials(n) ~ 1, family = binomial())` | identity | |
| Rate_2 | `k | trials(n) ~ 0 + g` | identity | |
| Rate_3 | `k | trials(n) ~ 1`, two rows | identity | |
| Rate_4 | Rate_1's spelling; the predictive is `simulate()` | identity | the program's `thetaprior` is a prior draw with no data, dropped |
| Rate_5 | Rate_3's spelling | identity | |
| Survey | `frm(k ~ 1, family = bcm_survey(nmax = 500))` | identity | the number sent is summed out; `latent_probs()` reports it |

`Survey` is the first model whose likelihood is not a product over
rows. The number of surveys sent is a whole number, a maximum over a
whole number is not an estimate, and so it is summed out of the whole
response at once through the structured protocol's `loglik` slot:

```{r bcm-survey, eval = FALSE}
```

## Chapter 4, means and standard deviations

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Gaussian | `frm(x ~ 1)` with the book's prior through `set_prior()` | identity, constant `-log(10)` | |
| SevenScientists | `bf(x ~ 1, sigma ~ 0 + obs)` with a prior on the log scales | identity | **the likelihood is unbounded**; see below |
| IQ | `frm(y ~ 0 + person)` | identity, constant `-3 log(300) - log(100)` | |

Seven Scientists is the first model that maximum likelihood cannot fit.
Profiling out the seven standard deviations leaves
`-sum(log|x_i - mu|) - 7/2`, which grows without bound as `mu`
approaches any observation, so there is no interior maximum and no
local one either. The book's `gamma(.001, .001)` prior on the
precisions is what makes the Bayesian model proper; the fit here
carries a `normal(0, 1)` prior on the log standard deviations instead,
which is a MAP fit and is stated as such, and the Stan program carries
the same penalty so the constant is zero.

## Chapter 5, data analysis

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Correlation_1 | `mvbf(bf(x1 ~ 1), bf(x2 ~ 1)) + set_rescor(TRUE)` | identity | |
| Correlation_2 | `se(sd, sigma = FALSE)` on each response plus `(1 | p | id)` | identity, inner gradient | `mi(sd)` cannot reach it; see below |
| Kappa | `cbind(y1, y2, y3, y4) ~ 1` with `bcm_kappa()` | identity, structure | the tree is saturated, so the estimate IS Cohen's kappa |
| ChangeDetection | `frm(c ~ 1, family = bcm_changepoint(t))` | identity, recovery | the changepoint is summed out; the port's continuous `tau` is a different model |
| Planes | **refused**: the model has no free parameter | identity of the arithmetic | `bcm_planes_posterior()` computes it |
| ChaSaSoon | `z | trials(n) + vint(hi) + weights(w) ~ 1` with `bcm_binomial_band()` | identity | a band of counts, not a marginalization |

`mi(sd)` is frmtmb's measurement-error spelling and it does not reach
Correlation_2: `mi()` refuses `rescor = TRUE`, and with a correlated
random intercept instead there would be two variances per response, the
latent one and the response's own sigma, which one observation per
subject cannot separate. That is a property of the design and not a gap
in the grammar. `se(sd, sigma = FALSE)` replaces the residual scale
rather than adding to it, which is exactly the book's model.

`Planes` is refused by name. Ten planes were marked, a later sample of
five held four marked ones, and the fleet size has a discrete uniform
prior; there is no parameter to estimate, which is also why its own
Stan port runs with `algorithm = "Fixed_param"`. `frm()` estimates
parameters, so there is no formula to write. The arithmetic is checked
against the reference program directly, and the posterior mode falls
where Lincoln and Petersen put it, at `floor(x n / k)`.

`ChaSaSoon` is interval censoring, and `cens("interval")` does not
reach it. Two gates stand in the way and only the first is a seam:
`cens()` needs a family with a CDF, which `frmtmb_family(lcdf = )`
supplies, and then `cens()` refuses every family whose `type` is
`"discrete"` whatever it supplies. So the band is written into the
family: the response is the smallest count consistent with the row and
`vint(hi)` the largest, `log(F(hi) - F(y - 1))` is the exact density
when the two agree and the band probability otherwise, and `weights()`
carries the 949 repeats. The test asserts both refusal messages, so the
day either gate moves it says so.

## Chapter 6, latent mixtures

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Exams_1 | `mixture(binomial, binomial)` with `mu2` and `theta1` pinned by an offset | identity | |
| Exams_2 | Exams_1 plus `(1 | id)` on `mu1` | identity, inner gradient | **Laplace**; see below |
| Malingering_1 | `mixture(binomial, binomial)`, both rates free | identity | label switching: the order restriction is a prior |
| Malingering_2 | `mixture(beta_binomial, beta_binomial)` | identity | the per-person rate integrates out exactly, and **the second precision's estimate is infinity**; see below |
| Cheating | `mixture(beta_binomial, beta_binomial)` | structure | classifies better than chance |
| TwentyQuestions | `bf(k ~ plogis(lp) * plogis(lq), ..., nl = TRUE)` with `bernoulli("identity")` | identity, value only | one rate is driven to a boundary, so the gradient is not asserted |
| TwoCountryQuiz | `frm(k ~ 1, family = bcm_two_country(person, question))` | recovery | the Stan port does not exist |

Two of these are worth reading twice.

**Malingering_2 and Cheating.** The book gives every person a rate of
its own with a beta prior whose mean and precision belong to their
group. That rate appears nowhere else, and a binomial rate with a beta
prior integrates in closed form to a beta-binomial, so the model with
the per-person rates summed out is exactly
`mixture(beta_binomial, beta_binomial)` in frmtmb's own mean-precision
parameterization. Doing the integral is not a convenience: left free,
one rate per person with one observation each is a saturated fit.

On Malingering_2's data the integral makes the model estimable and
still leaves one direction unbounded. The second component's precision
has a profile that rises monotonically to the binomial limit, so its
maximum likelihood estimate is infinity and the fit stops at
`phi2 = 7.19e8` only because the surface is flat to machine precision
there. What that says is that the bona fide group has no person-level
overdispersion: it is a plain binomial. The same mixture written with
component 2 at the limit, `mixture(beta_binomial, binomial)`, reaches
the same optimum with one parameter fewer, and the 2.2e-5 between the
two log-likelihoods is the identity residual in this chapter's table.
It is precision loss in the `lgamma` differences at divergent shapes,
not a difference of convention: at `phi2 = 7.19e8` the beta-binomial's
density sums to 1.0000022 rather than to one (the longhand formula gives 0.999999998), while at component 1's
ordinary `phi1 = 7.5` it sums to one and agrees with a longhand
`lbeta` calculation to 1e-14. Cheating, the same model on different
data, has both precisions moderate (8.9 and 5.8) and is not affected.

**TwoCountryQuiz.** Its Stan port is a file that says "this model is
not implemented". The obstacle is real: a cell's rate depends on the
person's country AND the question's, so the two sets of indicators do
not factorize and there is no per-row `log_mix` to write. They do
factorize CONDITIONALLY, which is the whole trick, and the cost is
`2^questions` passes over the data rather than
`2^(people + questions)`. The family refuses by name above sixteen
questions rather than enumerating a set it cannot hold.

Which country is called 1 is not identified: every question pattern has
a complement under which every person's country flips, and the two
carry the same likelihood. So `latent_probs()` reports exactly one half
for every person, and that is the answer rather than a failure. The
quantity that survives the flip is whether two people came from the
same country, and `bcm_two_country_agreement()` computes it from the
same sum.

**Every mixture whose components are both free needs a start.**
`mixture()` initializes each component's `mu` from a quantile of the
response, which for a binomial or a beta-binomial is a COUNT and falls
outside the logit link's range; both components then start at the
link's origin, where they are the same distribution.
`mixture(beta_binomial, beta_binomial)` stays there and returns a fit
with identical components and every latent probability at one half. The
tests start every free mixture from two separated rates.

## Chapters 8 and 9, model selection

Every model in these two chapters fits. Their **Bayes factors are not
here**: each one is a Savage-Dickey density ratio read off the
posterior of one parameter at one point, and a maximum likelihood fit
has no posterior to read. `frmtmb.sample` is gaining the ratio; the
tests fit the models and pin the quantity the ratio is taken over, so
the evidence test can be written the moment it lands.

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Pledgers_1 | `s | trials(n) ~ 0 + g` | identity | |
| Pledgers_2 | the same fit | structure | the order restriction does not bind at the estimate |
| Geurts | `k | trials(n) ~ group + (1 | id)` with `binomial(link = "probit")` | identity, inner gradient | |
| GeurtsOrderRestricted | the same fit | structure | whether the restriction binds is decided by the sign of the unrestricted delta |
| Zeelenberg | `s | trials(n) ~ both + (1 + both || id)`, probit | structure | |
| OneSample | `frm(x ~ 1)`; delta is `mu / sigma` | structure | the book's parameterization is a reparameterization |
| OneSampleOrderRestricted | the same fit | structure | the restriction does not bind |
| TwoSample | `frm(v ~ g)`; delta is the standardized difference | identity | |

## Chapter 10, memory retention

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Retention_1 | `bf(k | trials(n) ~ bcm_cap1(exp(-plogis(la) * t) + plogis(lb)), la ~ 1, lb ~ 1, nl = TRUE)` | identity | |
| Retention_2 | the same with `la ~ 0 + id, lb ~ 0 + id` | identity | |
| Retention_3 | the same with `la ~ 1 + (1 | id)` and `lb ~ 1 + (1 | id)` | identity, inner gradient | the group distribution is logit-normal, not the book's truncated normal |

The `min(1, .)` in the retention curve is carried, through
`bcm_cap1()`, and it is not decoration: one subject recalled every item
at the shortest lag, so an uncapped `exp(-alpha t) + beta` runs past
one and that cell's likelihood grows without bound. RTMB has no `pmin`
over advectors and a branch on a parameter cannot be taped, but
`min(1, u) = u - (|u - 1| + (u - 1)) / 2` is exact and `abs()` is
taped.

The group distribution is the one adaptation of substance in the
chapter. frmtmb's random effects are normal on the LINEAR PREDICTOR,
which here is a logit, so a truncated-normal random effect on `(0, 1)`
is not in this grammar and the Stan program says logit-normal too.

## Chapter 11, signal detection theory

Equal-variance Gaussian SDT says the hit rate is `Phi(d/2 - c)` and the
false-alarm rate is `Phi(-d/2 - c)`, so with one row per
(case, trial type) and

* `half` = +1/2 on signal trials and -1/2 on noise trials,
* `bias` = -1 everywhere,

the linear predictor is `d * half + c * bias` and the two regression
coefficients ARE discriminability and criterion. The alternative
coding, an intercept plus a signal indicator, is the more familiar GLM
one and puts the random effects on the wrong pair: it makes the
criterion and `d'` correlated by construction, and the book's
hierarchical model gives them independent group distributions.

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| SDT_1 | `y | trials(N) ~ 0 + case:half + case:bias`, `binomial(link = "probit")` | identity | the demo data separate one case, so the book's own first data set has no finite estimate |
| SDT_2 | `y | trials(N) ~ 0 + half + bias + (0 + half + bias || id)` | identity, inner gradient | **Laplace** |
| SDT_3 | the SDT_2 fit | identity against SDT_3's program at `xi = 1` | parameter expansion is a ridge, not a model |

Signal detection theory is defined on the normal scale, so these need
a probit. The core registry supplies it, and `binomial(link =
"probit")` is the whole spelling. `stats::binomial()` accepts the name
and frmtmb resolves it against its own AD-safe registry rather than
`stats::make.link()`, so the tape sees a probit written over plain
arithmetic. The registry entry also carries `logit_eta`, the exact log
odds behind the linear predictor, which puts a saturated cell on
`dbinom_robust()` instead of a plain round trip.

## Chapter 12, psychophysical functions

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| PsychophysicalFunction1 | `frm(r | trials(n) ~ xc + (xc || subj))` | identity, inner gradient | **Laplace** |
| PsychophysicalFunction2 | the same plus `phi ~ 1 + (1 | subj)` with `bcm_contaminant()` | identity, inner gradient | **Laplace**; the contaminant rate is integrated out |
| PsychometricFunction1_Answers | PsychophysicalFunction1's fit | structure | the exercise answers derive JND and PSE from the same fit |

The book's contaminant process has a rate of its own per cell with a
`beta(1, 1)` prior. That rate is integrated out rather than estimated,
because a binomial rate with a `beta(1, 1)` prior integrates to the
discrete uniform on `0..n`:

$$\int_0^1 \binom{n}{r} p^r (1-p)^{n-r}\,dp = \frac{1}{n+1}.$$

The integral has to be done. Left as a free parameter per cell, the
contaminant fits every cell exactly, the mixture likelihood is
maximized by a contamination rate of one, and there is no maximum
likelihood estimate of the psychometric function at all. The book's
sampler never meets this, because averaging over the same parameter is
the same integral.

## Chapter 14, multinomial processing trees

**Four of the five programs are one model.** MPT_2 through MPT_5 differ
only in how the multivariate normal over the individual differences is
written: parameter expansion with a Wishart prior, an LKJ prior on a
correlation matrix, its Cholesky factor, and a non-centered
reparameterization. The README that ships with them says as much. Their
likelihood is identical, so frmtmb fits them once.

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| MPT_1 | `cbind(k1, k2, k3, k4) ~ 1` with `bcm_mpt_pairs()` | identity, structure | |
| MPT_2 to MPT_5 | `1 + (1 | p | id)` on each of the three probits | identity, inner gradient | **Laplace**; the correlations are the latent-trait ones and their standard errors do not exist; see below |

The hierarchical fit is the one in the port whose `diagnose()` is not
clean, and the reason is the model rather than the search. Four
categories with a fixed total is three free counts per subject, the
model gives each subject three latent traits, and an unstructured three
by three covariance over those traits is six more parameters for
21 subjects to carry.

The estimates are stable: five optimizer settings return the same
log-likelihood (-102.478) and the same standard deviations (0.517,
0.464, 0.296) and correlations (-0.529, 0.539, -0.450), which are the
case study's own numbers, and the identity holds with an inner gradient
of 2.1e-10. The curvature is not: the Hessian is not positive definite
and `theta_4` and `theta_6` have no finite standard error. The fit is
not saturated either, which was the first thing to rule out; the
largest gap between a fitted and an observed cell proportion is 0.177,
so the random effects really are pooled.

The case study meets the same wall from the other side. Its README
records divergent transitions and correlation matrices that are not
positive definite during estimation, and MPT_2's Wishart prior and
MPT_3's LKJ prior are what hold this block down. Maximum likelihood has
no such prior, so it reports the degeneracy instead, and the test
asserts every line of what `diagnose()` says.

## Chapter 15, SIMPLE

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| SIMPLE_1 | `bf(k | trials(n) ~ 0 + set, s ~ 0 + set, t ~ 0 + set)` with `bcm_simple(set, pos, m)` | identity | needs a start above the cap; see below |
| SIMPLE_2 | `bf(k | trials(n) ~ 1, s ~ 1, t ~ ll)` with `link_t = "identity"` | structure | the threshold is a linear function of list length, so its link is the identity |

An item's discriminability is its similarity relative to every other
item in the SAME list, so one row's contribution reads the whole list:
another `loglik` slot, and a second use of the padded block that
`inst/rl/rw-delta.R` introduced.

`min(1, sum(resp))` is carried here too, and it has the same edge that
retention's does: at a recall probability of one, a serial position
where somebody failed has a log density of minus infinity and a
gradient of NaN, so an infeasible START is fatal rather than merely
slow. The family's own default threshold is 0.6 for that reason, and
every start at or above it reaches the same optimum.

## Chapter 16, extrasensory perception

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| Ability | Correlation_1's spelling | identity, at Correlation_1 | the hundred pairs of proportions are not re-embedded |
| OptionalStopping | Correlation_1's spelling | identity | |
| Extraversion | `mvbf()` of a probit binomial and a probit gaussian sharing `(1 | p | id)` | identity, inner gradient | **Laplace** |
| OptionalStopping_Answer_1 | not ported | | the chapter's exercise answers, on the model OptionalStopping already ports |

Extraversion is a `gaussian(link = "probit")`. The book writes the
extraversion score as normal around `100 * Phi(theta_2)`; dividing the
response by 100 makes the mean `Phi(theta_2)` exactly, so the whole
nonlinearity is a link and the latent pair is an ordinary correlated
random intercept shared by two responses.

The known measurement standard deviation rides on `se(sd)`, which maps
the free `sigma` out and leaves the per-row standard deviation as the
whole residual scale. This model used to need a custom family and
`vreal(sd)` instead, because the core gates `se()` on the family name
and a custom family could not pass that gate. Once the probit came to
the registry there was no custom family left to gate.

## Chapter 17, the generalized context model

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| GCM_1 | `y | trials(t) ~ 1` with `bcm_gcm(stim, d1, d2, a)` | identity | |
| GCM_2 | `bf(y | trials(t) ~ 0 + subj, w ~ 0 + subj)` | identity | ten of the forty subjects |
| GCM_3, GCM_3_optimized | not ported | | the case study's own contaminant extension; the same family with a mixture on top |

## Chapter 19, the Balloon Analogue Risk Task

The model looks like a cognitive process model and IS a Bernoulli
logistic regression. The probability of stopping at pump `k` is
`inv_logit(beta * k - beta * omega)`, so `stop ~ k` recovers both
coefficients and the book's parameters come back by arithmetic:
`omega = -intercept / slope` and `gamma+ = -omega log(1 - p)`.

| model | frmtmb spelling | validated by | caveat |
|---|---|---|---|
| BART_1 | `frm(stop ~ k, family = bernoulli())` | identity | |
| BART_2 | `stop ~ 0 + cond + cond:k` | structure | the group distribution over three conditions is a prior, so the fit is unpooled |

## Refused, by name

One model, and the reason is the model rather than the grammar.

| model | why |
|---|---|
| Planes | The model has NO free parameter: its own Stan port runs with `algorithm = "Fixed_param"`. `frm()` estimates parameters, so there is no formula to write. `bcm_planes_posterior()` computes the posterior over the fleet size directly and its arithmetic is checked against the reference program. |

## Not ported

These are unfinished work, not obstacles, and none is blocked by a
missing seam: each is a `loglik` slot of the shape
`inst/bcm/marginal.R` already uses, or a fit of a shape already in the
port. They are listed separately from the refusal above because a
reader should be able to tell the two apart.

| model | count | note |
|---|---|---|
| NumberConcept_1 to _3 | 3 | the knower-level models are a finite sum, but over a per-child level that indexes a different response distribution in each of two tasks |
| Search, Stop, SearchStop, TTB | 4 | heuristic decision making |
| GCM_3, GCM_3_optimized | 2 | the case study's contaminant extension of GCM_2: the same family with a mixture over it |
| OptionalStopping_Answer_1 | 1 | the chapter's exercise answers, on the model OptionalStopping already ports |

Ten models, against the 52 ported and the one refused: 63 in all, which
is what the reference repository holds once the `GettingStarted` copy of
`Rate_1` and the `! other` scratch directory are excluded.

## The Laplace caveat

Every row marked **Laplace** has a CONTINUOUS random effect underneath
something that is not Gaussian in it: a marginalized discrete latent
class, a probit binomial with four trials, a contaminant mixture. The
identity is still exact, because it compares the JOINT density at the
conditional modes and both programs hold the same joint. What is
approximate is the MARGINAL that frmtmb maximizes, and the identity
cannot see that at all.

`frm(importance = )` measures it. A positive draw count replaces the
Laplace approximation with an importance-sampling correction of it,
which is unbiased for the exact integral, and `fit$importance` records
the Monte Carlo standard error of the corrected log-likelihood. Where
it is admitted the correction is worth running:

```{r importance, eval = FALSE}
d <- data.frame(k = c(21, 17, 21, 18, 22, 31, 31, 34, 34, 35, 35, 36,
                      39, 36, 35),
                n = 40L, zero = 0, id = factor(1:15))
fo <- bf(k | trials(n) ~ 1 + (1 | id), mu2 ~ 0 + offset(zero),
         theta1 ~ 0 + offset(zero))
lap <- frm(fo, family = mixture(binomial, binomial), data = d)
imp <- frm(fo, family = mixture(binomial, binomial), data = d,
           importance = 512)
c(laplace = logLik(lap), importance = logLik(imp), imp$importance$mcse)
```

On Exams_2 the Laplace log-likelihood is -44.095 and the corrected one
is -44.109, with a Monte Carlo standard error of 0.0018. So the
approximation is off by about 0.013 nats over fifteen subjects: small,
but larger than the noise in the number that measures it, which is what
makes it worth measuring rather than assuming.

Its scope is narrower than the caveat, and the refusals are by name:
one response, one grouping factor, a family with a ROWWISE density, and
no nonlinear predictor. So

* Exams_2, SDT_2, PsychophysicalFunction1 and 2, MPT_2 to _5, and
  Geurts are **admitted**: rowwise density, one grouping factor;
* Extraversion and Correlation_2 are **refused**, because they have two
  responses;
* Retention_3 is **refused**, because its predictor is nonlinear;
* GCM, SIMPLE, Survey, ChangeDetection and TwoCountryQuiz are
  **refused**, because a structured family has no rowwise density to
  stack, which is the same fact that put them on the `loglik` slot.

For the last group the honest statement is that the Laplace
approximation is not involved at all: they have no random effects, and
their marginalization over a discrete latent is an exact finite sum.
The caveat is about the models that mix the two, and those are exactly
the rows marked **Laplace** above.

`check_laplace()` in `frmtmb.sample` is the other route: it samples the
joint with tmbstan and compares.

## Where the code is

| what | where |
|---|---|
| the identity harness | `tests/testthat/helper-stan.R` |
| the chapter tests | `tests/testthat/test-bcm-*.R` |
| the extra binomials | `inst/bcm/binomial-extras.R` |
| the processing trees | `inst/bcm/process-trees.R` |
| the similarity models | `inst/bcm/similarity.R` |
| the finite-sum families | `inst/bcm/marginal.R` |
| the port's own notes | `dev/bcm-findings.md` |

The Stan tier is opt-in, exactly like the brms one:

```r
Sys.setenv(FRMTMB_BRMS_FIT_TESTS = "true", NOT_CRAN = "true")
```
