---
title: "frmtmb for brms users"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{frmtmb for brms users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

frmtmb reimplements a documented subset of the brms grammar with
identical spelling, so brms model code ports mechanically: change
`brm()` to `frm()`, keep or drop the priors, and get a fit in
milliseconds-to-seconds instead of an MCMC run. Estimation is ML (or
REML) with the Laplace approximation for latent effects; inference is
frequentist (Wald and profile intervals, likelihood-ratio tests, AIC).

## What ports directly

| brms | frmtmb | notes |
|---|---|---|
| `bf(y ~ x + (x \| g))` | same | plus `us()`, `diag()`, `homdiag()`, `cs()`, `ar1()` wrappers |
| `bf(y ~ x, sigma ~ z + (1 \| g))` | same | any dpar, full grammar |
| `bf(y ~ x, sigma = 1)` | same | fixed via TMB `map` |
| `s(x)`, `t2(x, z)` | same | mgcv smooths in any dpar; `te()`/`ti()` unsupported |
| `(1 \| p \| g)` | same | cross-formula RE correlation |
| `mvbind(y1, y2) ~ x`, `mvbf()`, `set_rescor()` | same | per-response families; `rescor` gaussian-only |
| `y \| trials(n)`, `weights(w)`, `cens(c)`, `trunc(lb=, ub=)` | same | `cens`/`trunc` need a CDF-carrying family |
| `nl = TRUE` | same | a located prior places the start, else provide `start` (`par_template()` names it); se.fit on the nonlinear mu not yet |
| `lf()`, `nlf()` | same | `nlf()` on any dpar, bodies chain to any depth (below) |
| `custom_family()` | same idea | the lpdf is plain R over RTMB advectors, not Stan code |
| `cumulative()`, `multinomial()` | same | multinomial takes `K` explicitly |
| `car(M, gr = g, type =)` | same | all four types |
| `(1 \| mm(g1, g2))`, `mmc(x1, x2)` | same | multi-membership, `weights =` and `scale =` included (below) |
| `ar(t, g, cov = TRUE)`, `ma()`, `arma()`, `cosy()`, `unstr()` | same | gaussian/student only; `cov = TRUE` required (below) |
| `data2 = list(W = W)` | same | also resolves compound expressions (below) |

## What changes

- `brm(...)` becomes `frm(...)`; there are no chains or warmup, and a
  plain call is maximum likelihood rather than a posterior.
  `REML = TRUE` replaces priors as the small-sample correction for
  variance components.
- A `prior = c(prior(...), prior(...))` argument is kept as it stands.
  `frm()` spells the argument `prior`, as brms does, and `prior()`,
  `prior_()` and `prior_string()` build the same specification
  `set_prior()` does; a prior object that brms itself built is
  translated row by row, so the argument works whichever package's
  `prior()` was in scope.

  What the fit MEANS changes. `frm(prior = )` is MAP: the density is a
  PENALTY on the likelihood and the answer is one mode, not a
  posterior, so the reported log likelihood, AIC and `anova()` are
  penalized quantities. The estimates land close where the priors are
  weak: a kidney model with brms's own priors gives `sd(patient)` 0.38
  against brms's posterior mean of 0.40. But "close" is a measurement,
  not a guarantee, and the two numbers are not the same quantity. Read
  a prior here as regularization.

  A prior class frmtmb cannot mean the same way brms does (`sigma`,
  brms's mixture `theta`) is refused by name rather than translated
  into something else.
- A prior with a location DOES place a nonlinear start. brms uses its
  priors to place its sampler, and `frm()` reads them the same way for
  nonlinear parameters: a `normal()`, `student_t()` or `cauchy()` prior
  on a nonlinear coefficient starts it at the prior location, and a
  message names what was placed. The insurance-loss growth curve of the
  brms nonlinear vignette fits from its brms priors alone.

  Without such a prior a nonlinear model still needs `start`, because
  `frm()` optimizes rather than samples and evaluates the objective AT
  the starting values, where a nonlinear body is rarely defined.
  `par_template(formula, data)` lists the parameter names before there
  is a fit; edit the result and hand it back as `start =`. brms has no
  counterpart to it, and needs none: `brm(init =)` takes Stan program
  names and brms initializes at random in a bounded range instead.
- Posterior summaries become ML estimates: `fixef()` returns point
  estimates, `confint()` gives Wald or profile intervals, `anova()`
  gives likelihood-ratio tests, and `AIC()`/`BIC()` replace `loo()`.
- `posterior_predict()` becomes `simulate()`; `posterior_epred()` over
  `newdata` becomes `predict(newdata, type = "response")`. On a
  `cens()` response both draw the LATENT uncensored value, as brms
  does; `simulate(censored = TRUE)` applies the censoring mechanism
  instead.
- `sample_prior = "only"` plus `posterior_predict()` becomes
  `frm_simulate(formula, data, prior = set_prior(...))`, which draws
  a parameter vector per simulation and returns it alongside the
  responses.
- `mo()` monotonic effects, `mi()` one-step imputation of continuous
  predictors, `mi(sdx)` measurement error (the `me()` replacement,
  as in current brms), `cs()` category-specific ordinal effects,
  `gp(x)` / `gp(x, k =)` Gaussian processes, and `mixture()`
  families (including group-level latent classes via `groups = ~g`)
  all work. `mo()`/`mi()` support two-way `:`/`*` interactions with
  numeric terms; discrete predictors cannot be imputed in-model, the
  same restriction Stan has. Multiply imputed data can instead go
  through `frm_multiple()` (Rubin's rules), the `brm_multiple()`
  analog. Being frequentist, that fit also needs pooled tests, which
  brms gets for free from the posterior: `hypothesis()` pools any
  function of the parameters, and `anova()` on the result compares two
  nested `frm_multiple()` fits with the D1, D2 and D3 rules of `mice`
  (D3 by default). Mixture fits are ML: expect multimodality, compare
  starts (`frm_allfit()`).
- `binomial()` without `trials()` is accepted and means Bernoulli, the
  `stats::glm()` convention; brms rejects it and asks for `trials()` or
  `bernoulli()`. The divergence is permissive, so brms code ports
  unchanged, but frmtmb code written this way does not port back.
- `ar()`, `ma()`, `arma()`, `cosy()` and `unstr()` are supported in
  their covariance form only; see the next section.
- glmer's proportion-response idiom (`weights = size`) becomes
  `y | trials(size)` with either proportions or counts; `weights()`
  in the formula stays a frequency weight, as in brms.
- `se()` works as in brms (meta-analysis), including
  `se(x, sigma = TRUE)` and the phylogenetic version with
  `gr(g, cov = A)`.
- `mm()` ports with its `weights =` and `scale =` arguments and with
  `mmc()`; its other arguments have different spellings here (below).

## Nonlinear formulas: `nl = TRUE` and `nlf()`

`nl = TRUE` says the RESPONSE formula is a nonlinear body, so only `mu`
can be nonlinear. `nlf()` names the parameter instead, so any of them
can be, and the two spellings meet in the same place:

```r
# the same model, both ways
frm(bf(y ~ exp(b * x), b ~ 1, nl = TRUE), data = d)
frm(bf(y ~ a) + nlf(a ~ exp(b * x)) + lf(b ~ 1), data = d)

# and the model nl = TRUE cannot spell: a nonlinear sigma, linear mu
frm(bf(y ~ x) + nlf(sigma ~ a + b * z) + lf(a ~ 1, b ~ 1), data = d)
```

A parameter's link is applied to its body's value, as in brms: the
`sigma` above is `exp(a + b * z)`. Bodies chain to any depth
(`nlf(a ~ cc * x) + nlf(cc ~ exp(b))`), are computed in dependency
order, and a cycle is refused by name.

Two deliberate differences from brms, both supersets:

- brms requires `nl = TRUE` on the `bf()` whenever the response formula
  names a nonlinear parameter. Here `nlf()` has already declared the
  name, so `bf(y ~ a) + nlf(a ~ ...)` is enough; the flag still works
  and means the same thing.
- A body may name another distributional parameter of the same
  response, which reads that parameter's per-row value. brms reads such
  a name as a data column and refuses the model when no column has it.
  A column still wins here, so a ported body keeps its meaning. This is
  the spelling for a variance function of the model's own mean, which
  nlme writes `varPower(form = ~ fitted(.))`:

```r
# sd = exp(ls) * |mu|^th, one joint maximum likelihood
frm(bf(y ~ x) + nlf(sigma ~ ls + th * log(abs(mu))) +
      lf(ls ~ 1, th ~ 1), data = d)
```

## Multi-membership

`(1 | mm(g1, g2))` means what it means in brms: the observation belongs
to several levels of one grouping factor, and its design row averages
those levels' effects. The pooled level set, the `1/J` default weights
and the `scale = TRUE` normalization of a supplied weight matrix are
brms's, checked against `make_standata()`'s `J_*`, `W_*` and `Z_*`
arrays, so a ported model builds the same design matrix.

`mmc(x1, x2)` also ports: it is one random-slope coefficient of the
term whose covariate value is member specific, member `k` taking
argument `k`.

```r
# brms and frmtmb, identical spelling
frm(bf(y ~ x + (1 | mm(school1, school2))) + gaussian(), data = d)
frm(bf(y ~ x + (1 | mm(school1, school2,
                       weights = cbind(share1, share2)))) + gaussian(),
    data = d)
frm(bf(y ~ (mmc(hours1, hours2) | mm(school1, school2))) + gaussian(),
    data = d)
```

brms's remaining `mm()` arguments are spelled the way any other
grouping term spells them here, and the parser names the replacement
when it refuses one:

| brms | frmtmb |
|---|---|
| `mm(g1, g2, cor = FALSE)` | `diag(x \| mm(g1, g2))`, or `(x \|\| mm(g1, g2))` |
| `mm(g1, g2, id = "q")` | the `\|ID\|` key, `(x \| q \| g)` - not yet over `mm()` |
| `mm(g1, g2, cov = A)` | `gr(g, cov = A)` - not yet over `mm()` |
| `mm(g1, g2, by = )`, `pw =`, `dist =` | no equivalent yet |

`?frmtmb-multimembership` is the full page.

## Within-group residual correlation

`ar()`, `ma()`, `arma()`, `cosy()` and `unstr()` are written where brms
writes them, in the model formula, and mean what brms means under
`cov = TRUE`: the residuals of one group become a single correlated
draw, `y_g ~ N(mu_g, D R D)` with `D` the diagonal of that group's
`sigma` values. `?frmtmb-autocor` has the full page; four things
matter when porting.

**`cov = TRUE` is required for `ar()`, `ma()` and `arma()`.** brms
defaults to `cov = FALSE`, which is a different likelihood - a
residual regression that conditions on each group's first rows - and
that form is not implemented. The call is refused rather than
reinterpreted, so a brms model written with the default has to be
changed deliberately. `cosy()` and `unstr()` have no `cov` argument in
brms either and port unchanged.

```{r, eval = FALSE}
# brms
brm(y ~ x + ar(week, subj, cov = TRUE), data = d)
# frmtmb
frm(bf(y ~ x + ar(week, subj, cov = TRUE)) + gaussian(), data = d)
```

**`sigma` is the marginal residual SD, not the innovation SD.** brms's
`cholesky_cor_ar1()` divides by `1 - ar^2`, so its `sigma` under
`ar()`/`ma()`/`arma()` is the innovation scale, while under `cosy()`
and `unstr()` it is the marginal one. Here every structure uses the
marginal scale, so `sigma()`, pearson residuals and a `sigma ~ x` model
mean one thing throughout. The correlation parameters agree exactly;
convert the scales with `sigma_marginal = sigma_innovation /
sqrt(1 - phi^2)` for AR(1).

**Only `gaussian()` and `student()`.** These are the two families brms
gives a real residual covariance. brms accepts the same spelling for
poisson, binomial and the rest, but fits a latent Gaussian AR process
added to the linear predictor instead - which is a random effect, and
is spelled as one here:

```{r, eval = FALSE}
# brms, poisson: a latent AR process on the linear predictor
brm(cnt ~ x + ar(week, subj), data = d, family = poisson())
# frmtmb: the same idea as a random effect over the time factor
frm(bf(cnt ~ x + ar1(factor(week) + 0 | subj)) + poisson(), data = d)
```

**Higher orders are allowed, and gaps are honored.** brms limits
`cov = TRUE` to order one; `ar(p = 3)` and `arma(p = 2, q = 1)` work
here. And the lag between two rows is the distance between their
positions in the global set of time levels, so a subject missing week 3
gets `cor(week2, week4) = rho^2`; brms indexes `ar()`/`ma()`/`arma()`/
`cosy()` by position within the group and treats the missing row as no
gap. On complete balanced groups the two agree.

The parameters appear in `summary()` under "Within-group residual
correlation" and in `confint_varcorr()` under brms's names (`ar[1]`,
`cosy`, `cortime__1__2`); `autocor_matrix()` returns the fitted
correlation matrix. `weights()`, `cens()`, `trunc()`, `se()`, `mi()`,
`rescor = TRUE`, mixtures and `quadrature = TRUE` are refused, because
the likelihood no longer factorizes over rows - brms refuses the same
core set. Random effects alongside the correlated residual are the
point of the feature and are supported.

## Variance functions (nlme `weights = varFunc`)

There are no `varFunc` objects. The `sigma` formula is the variance
function: `sigma` uses a log link, so most of the nlme vocabulary is a
direct translation, and the translation was verified against
`nlme::gls()` under ML (log-likelihood agreement 1e-10 or better).

| nlme | frmtmb |
|---|---|
| `varIdent(form = ~ 1 \| g)` | `sigma ~ 0 + g` |
| `varExp(form = ~ v)` | `sigma ~ v` |
| `varPower(form = ~ v)` | `sigma ~ log(abs(v))` |
| `varFixed(~ v)` | `sigma ~ offset(0.5 * log(v))` |
| `varComb(A, B)` | the two terms in one `sigma` formula |

The coefficient of `log(abs(v))` is nlme's power `theta`. The
translation is also more general than a `varFunc`: `sigma` takes the
full predictor grammar, so random effects and smooths in the variance
are ordinary spellings.

`varPower(form = ~ fitted(.))`, variance as a function of the model's
own mean, is spelled with `nlf()` (see the nonlinear-formulas
section): `nlf(sigma ~ ls + th * log(abs(mu))) + lf(ls ~ 1, th ~ 1)`
is `sd = exp(ls) * |mu|^th`, estimated by one joint maximum
likelihood. Expect agreement with `gls` near 1e-2 rather than 1e-10:
`gls` alternates the mean fit against iteratively updated fitted
values and stops without a gradient check, while this is one joint
maximization of the same likelihood. For a positive response the
coupling often belongs in the family instead: `Gamma()` has sd
proportional to the mean, and `tweedie()` estimates the power in
`Var = phi * mu^p` directly.

## Structural objects: `data2`

Matrices that are not columns of `data` go in `data2`, the brms
spelling: the adjacency matrix of `car()`, the mesh triple of
`spde()`, and the matrices of `gr(prec = )`, `gr(cov = )` and
`equalto()`.

```{r, eval = FALSE}
frm(bf(y ~ x + car(W, gr = loc, type = "icar")) + gaussian(),
    data = d, data2 = list(W = W))
frm(bf(phen ~ 1 + (1 | gr(species, cov = A))) + gaussian(),
    data = d, data2 = list(A = A))
```

Two differences from brms. First, brms accepts a bare name in `data2`
and nothing else, while frmtmb evaluates the whole expression with
`data2` in front of the data mask, so `gr(g, cov = solve(Q))` finds
`Q` in `data2`. Second, `data2` is optional: a matrix visible from the
formula environment is still found, which is what frmtmb did before
`data2` existed.

Use `data2` anyway. Its objects are stored on the fit, so the fit is
self-contained: `saveRDS()` writes the matrices with it, and
`refit()`, `update()`, `drop1()`, `influence()` and `frm_multiple()`
re-assemble in a fresh session where the calling environment that
built the matrix is gone. Without `data2` those refits depend on the
name still resolving where the model was written.

## Method conventions

frmtmb keeps the brms spelling for functions that originate in brms and
the stats/lme4 spelling for standard generics:

- brms-origin, same names and argument spellings:
  `conditional_effects()` (includes what `conditional_smooths()`
  covers), `hypothesis()` (a delta-method Wald test here),
  `pp_check()` (simulate-based, through bayesplot's `ppc_*`
  functions), `prior_summary()`, `fixef()`, `ranef()`, `VarCorr()`,
  `ngrps()`, `variables()` (the usable parameter names, e.g.
  `sd_Subject__Days`).
- stats-origin generics follow stats/lme4/glmmTMB, not brms:
  `predict()` returns a vector (with `se.fit = TRUE`, a list), not a
  draws matrix with `Estimate`/`Q2.5` columns; `fitted()` and
  `residuals()` return vectors; `confint()` takes `level =`, not
  `probs =`; `re.form` replaces `re_formula`. The exception is an
  ordinal family, where `fitted()` follows brms and returns the `n` by
  `K` matrix of category probabilities named by the response levels,
  exactly as `predict(type = "response")` does; the latent linear
  predictor is `predict(type = "link")`.
- `conditional_effects()` on an ordinal fit draws one probability
  curve per response category, which is what brms draws under
  `categorical = TRUE`.
- `coef()` is the per-group convention shared by all three packages:
  fixed effects plus conditional modes per grouping level.
- `plot(fit)` draws residual diagnostics, not MCMC traces; for
  simulation-based residual checks use `dharma_residuals()`.

## Seeing the model code

brms users inspect the generated model with `stancode(fit)`. The
frmtmb analog is the objective function itself: an ordinary R closure
that `RTMB::MakeADFun()` tapes, rebuilt deterministically from the
fitted frame.

```{r, eval = FALSE}
nll <- frmtmb:::build_objective(fit$frame)
nll(fit$estimates)   # joint negative log density at the estimates
```

Two properties matter when reading it:

- The closure takes the named parameter list (the shape of
  `fit$frame$par_template`) and returns the JOINT negative log
  density: random-effect prior terms are inside it. The marginal
  (Laplace) likelihood the optimizer minimized is what
  `MakeADFun(random = )` makes of it, exposed as `fit$obj$fn` over
  the flat outer-parameter vector; at the optimum
  `fit$obj$fn(fit$opt$par)` equals `-logLik(fit)`.
- All data is baked into the closure's environment (the reason
  refits re-tape instead of recompiling), so `environment(nll)`
  holds the assembled frame, and the closure evaluates on plain
  numeric parameter lists without any tape.

`RTMB::MakeADFun(nll, fit$frame$par_template, random = "b",
map = fit$frame$map)` reproduces `fit$obj`. There is no generated
code to print: the closure IS the model, and `body()`,
`environment()`, and the debugger work on it like on any R function.

## When you still want brms

Genuine prior information, full posterior uncertainty for derived
quantities, models with discrete latent structure beyond
observation-level mixtures, or `loo`-based model comparison. A
practical workflow is model screening in frmtmb and a final fit in
brms; `as_tmbstan(fit)` also runs NUTS on the frmtmb objective
directly.
