---
title: "frmtmb for brms users"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{frmtmb for brms users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

frmtmb reimplements a documented subset of the brms grammar with
identical spelling, so brms model code ports mechanically: change
`brm()` to `frm()`, keep or drop the priors, and get a fit in
milliseconds-to-seconds instead of an MCMC run. Estimation is ML (or
REML) with the Laplace approximation for latent effects; inference is
frequentist (Wald and profile intervals, likelihood-ratio tests, AIC).

## What ports directly

| brms | frmtmb | notes |
|---|---|---|
| `bf(y ~ x + (x \| g))` | same | plus `us()`, `diag()`, `homdiag()`, `cs()`, `ar1()` wrappers |
| `bf(y ~ x, sigma ~ z + (1 \| g))` | same | any dpar, full grammar |
| `bf(y ~ x, sigma = 1)` | same | fixed via TMB `map` |
| `s(x)`, `t2(x, z)` | same | mgcv smooths in any dpar; `te()`/`ti()` unsupported |
| `(1 \| p \| g)` | same | cross-formula RE correlation |
| `mvbind(y1, y2) ~ x`, `mvbf()`, `set_rescor()` | same | per-response families; `rescor` gaussian-only |
| `y \| trials(n)`, `weights(w)`, `cens(c)`, `trunc(lb=, ub=)` | same | `cens`/`trunc` need a CDF-carrying family |
| `nl = TRUE` | same | a located prior places the start, else provide `start` (`par_template()` names it); se.fit on the nonlinear mu not yet |
| `lf()`, `nlf()` | same | `nlf()` on any dpar, bodies chain to any depth (below) |
| `custom_family()` | same idea | the lpdf is plain R over RTMB advectors, not Stan code |
| `cumulative()`, `multinomial()` | same | multinomial takes `K` explicitly |
| `car(M, gr = g, type =)` | same | all four types |
| `(1 \| mm(g1, g2))`, `mmc(x1, x2)` | same | multi-membership, `weights =` and `scale =` included (below) |
| `ar(t, g, cov = TRUE)`, `ma()`, `arma()`, `cosy()`, `unstr()` | same | gaussian/student only; `cov = TRUE` required (below) |
| `data2 = list(W = W)` | same | also resolves compound expressions (below) |
| `bernoulli(link = "probit")`, and every other brms link name | same | the whole brms 2.23.0 roster (below) |

### Links

Every link name brms 2.23.0 accepts is accepted here under the same
spelling, and computes the same function: `logit`, `probit`,
`probit_approx`, `cauchit`, `cloglog`, `softit`, `identity`, `log`,
`log1p`, `logm1`, `inverse`, `1/mu^2`, `sqrt`, `softplus`, `squareplus`
and `tan_half`. `power12` is frmtmb's own, for the tweedie power
parameter.

Two cautions when porting a link:

- `stats::binomial()`, `stats::poisson()` and `stats::Gamma()` validate
  their link string through `stats::make.link()` before `frm()` sees it,
  so a name `stats` has never heard of (`probit_approx`, `softit`,
  `softplus`, `squareplus`) is refused by the constructor rather than by
  frmtmb. Use frmtmb's own family constructor instead:
  `bernoulli(link = "softit")`, `Beta(link = "probit")`,
  `negbinomial(link = "softplus")`.
- brms's `probit_approx` is two different functions inside brms itself.
  Its Stan program uses `Phi_approx()`, the logistic of
  `0.07056 x^3 + 1.5976 x`, while its R-side `inv_link()` answers
  `pnorm()`. frmtmb reproduces the Stan form, because that is the
  likelihood a brms fit was actually computed with.

The ordinal families are the one place the roster is narrower:
`cumulative()`, `sratio()` and `cratio()` take `logit` and `probit`
only, and `acat()` takes `logit` only.

## What changes

- `brm(...)` becomes `frm(...)`; there are no chains or warmup, and a
  plain call is maximum likelihood rather than a posterior.
  `REML = TRUE` replaces priors as the small-sample correction for
  variance components.
- A `prior = c(prior(...), prior(...))` argument is kept as it stands,
  and so is `prior = get_prior(...)`. `frm()` spells the argument
  `prior`, as brms does, and `prior()`, `prior_()` and `prior_string()`
  build the same specification `set_prior()` does; a prior object that
  brms itself built is translated row by row, so the argument works
  whichever package's `prior()` was in scope.

  A row applies whatever its `prior` string says. That is brms's own
  rule, and it means a table edited in place, the ordinary
  `gp <- get_prior(...); gp$prior[i] <- "normal(0, 20)"` workflow, is
  honored. Earlier releases keyed the decision on the table's
  `source` column, which brms does not update after such an edit, so
  the edited row was dropped.

  What the fit MEANS changes. `frm(prior = )` is MAP: the density is a
  PENALTY on the likelihood and the answer is one mode, not a
  posterior, so the reported log likelihood, AIC and `anova()` are
  penalized quantities. The estimates land close where the priors are
  weak: a kidney model with brms's own priors gives `sd(patient)` 0.38
  against brms's posterior mean of 0.40. But "close" is a measurement,
  not a guarantee, and the two numbers are not the same quantity. Read
  a prior here as regularization.

  Where a prior LANDS is brms's placement, class by class. A density on
  `sigma`, `shape`, `phi`, `nu` or another distributional parameter is
  a density on that parameter itself, as in brms, rather than on its
  logarithm; a `class = "Intercept"` density is evaluated at the
  intercept at the MEAN of the predictors, which is the one brms
  constrains; and an ordinal family's `class = "Intercept"` reaches the
  thresholds. On four of the shapes measured against Stan, frmtmb's
  penalized objective is now brms's posterior density up to one
  `log(2)` per lower-bounded parameter, a constant that moves no mode.

  **Which mode.** The optimum is the mode in the INTERNAL coordinates,
  which is Stan's target in unconstrained space, and not the mode of
  the parameter's own density. A density on `sigma` enters the
  objective as `log p(sigma) + log|dsigma/deta|`, and what is maximized
  over `eta = log sigma` is that whole sum, log-Jacobian included.

  Measured on a gaussian `y ~ x` fit with `normal(2.5, 0.4)` on
  `sigma`, over 150 rows drawn as `set.seed(150)`,
  `x <- rnorm(150)`, `y <- rnorm(150, 1 + 0.5 * x, 2.2)`: frmtmb
  reports `sigma` 2.133344, the mode of `L * p` in the LOG-SIGMA
  coordinate is 2.133358, and the mode of `L * p` in the SIGMA
  coordinate is 2.126533. frmtmb matches the log-sigma mode, and it
  does so deliberately: that coordinate is Stan's unconstrained target,
  and matching it is what keeps the two packages' objectives one
  function, which is what everything above rests on. The gap to the
  sigma-coordinate mode is the same kind of remark as the `log(2)`,
  and about the same size.

  **Two rows frmtmb accepts and brms refuses.** A distributional class
  written without `resp` on a MULTIVARIATE model applies to EVERY
  response, following frmtmb's own convention that a class-wide prior
  is class-wide; brms refuses such a row and asks for `resp`. Write
  `resp` to address one response. Separately, `cumulative()` has no
  `disc` distributional parameter here at all, so a `disc` formula is
  refused when the model is parsed and the acceptance rule above is
  never reached for it.

  **frmtmb's own `set_prior()` takes brms's words with brms's meaning,
  and that is a BEHAVIOR CHANGE for a script written before 0.53.0.**
  A distributional parameter has two spellings, and brms accepts each
  only on the model the other does not:

  | the model | the spelling | the density is on |
  | --- | --- | --- |
  | `bf(y ~ x)` | `set_prior(..., class = "sigma")` | sigma itself |
  | `bf(y ~ x, sigma ~ 1)`, `sigma ~ x` | `set_prior(..., class = "Intercept", dpar = "sigma")` | the log-scale intercept of sigma's predictor |

  frmtmb says the same and refuses each spelling by name on the other's
  model, naming the one that applies. Before 0.53.0,
  `set_prior("student_t(3, 0, 2.5)", class = "Intercept", dpar = "sigma")`
  was accepted on a model with no `sigma` formula, where it meant a
  density on LOG sigma. To reproduce that placement, write `sigma ~ 1`
  in `bf()`, which is the model that HAS a log-scale intercept; to get
  brms's placement, drop the `dpar =` and write the class. The numbers
  move with the meaning: on a gaussian `y ~ x` with 200 rows and a
  maximum-likelihood sigma of 2.0103, `normal(0, 0.5)` gave 1.9965
  under the old link spelling and gives 1.9429 under the class
  spelling, which is what
  `brms::prior(normal(0, 0.5), class = "sigma")` gave all along.

  A prior class frmtmb keeps somewhere else (brms's mixture `theta`,
  `simo`, `sds`, `sdgp`, `lscale`, `sdcar`, `car`) is refused by name,
  with the class that reaches the same parameters where one exists,
  rather than translated into something else. So is a `coef` on a `sd`
  or `cor` row: frmtmb addresses a whole random-effect block there, and
  applying a narrowed row without its `coef` would put the density on
  every standard deviation of the block.

  Because rows now apply, such a row in a `get_prior()` table stops the
  call where it used to be dropped in silence: on a model with `s()`,
  `gp()`, `mo()` or `car()`, blank that row or write the frmtmb class
  the message names. Every refused row of a table is named in one
  message, so one edit round is enough. Of 20 shapes a brms user would
  actually write, 3 stop: `s()`, `gp()` and `mo()`. The rest translate,
  zero-inflated, hurdle and negbinomial models included.
- A prior with a location DOES place a nonlinear start. brms uses its
  priors to place its sampler, and `frm()` reads them the same way for
  nonlinear parameters: a `normal()`, `student_t()` or `cauchy()` prior
  on a nonlinear coefficient starts it at the prior location, and a
  message names what was placed. The insurance-loss growth curve of the
  brms nonlinear vignette fits from its brms priors alone.

  Without such a prior a nonlinear model still needs `start`, because
  `frm()` optimizes rather than samples and evaluates the objective AT
  the starting values, where a nonlinear body is rarely defined.
  `par_template(formula, data)` lists the parameter names before there
  is a fit; edit the result and hand it back as `start =`. brms has no
  counterpart to it, and needs none: `brm(init =)` takes Stan program
  names and brms initializes at random in a bounded range instead.
- Posterior summaries become ML estimates: `fixef()` returns point
  estimates, `confint()` gives Wald or profile intervals, `anova()`
  gives likelihood-ratio tests, and `AIC()`/`BIC()` replace `loo()`.
- `posterior_predict()` becomes `simulate()`; `posterior_epred()` over
  `newdata` becomes `predict(newdata, type = "response")`. On a
  `cens()` response both draw the LATENT uncensored value, as brms
  does; `simulate(censored = TRUE)` applies the censoring mechanism
  instead.
- `sample_prior = "only"` plus `posterior_predict()` becomes
  `frm_simulate(formula, data, prior = set_prior(...))`, which draws
  a parameter vector per simulation and returns it alongside the
  responses.
- `mo()` monotonic effects, `mi()` one-step imputation of continuous
  predictors, `mi(sdx)` measurement error (the `me()` replacement,
  as in current brms), `cs()` category-specific ordinal effects,
  `gp(x)` / `gp(x, k =)` Gaussian processes, and `mixture()`
  families (including group-level latent classes via `groups = ~g`)
  all work. `mo()`/`mi()` support two-way `:`/`*` interactions with
  numeric terms; discrete predictors cannot be imputed in-model, the
  same restriction Stan has. Multiply imputed data can instead go
  through `frm_multiple()` (Rubin's rules), the `brm_multiple()`
  analog. Being frequentist, that fit also needs pooled tests, which
  brms gets for free from the posterior: `hypothesis()` pools any
  function of the parameters, and `anova()` on the result compares two
  nested `frm_multiple()` fits with the D1, D2 and D3 rules of `mice`
  (D3 by default). Mixture fits are ML: expect multimodality, compare
  starts (`frm_allfit()`).
- `binomial()` without `trials()` is accepted and means Bernoulli, the
  `stats::glm()` convention; brms rejects it and asks for `trials()` or
  `bernoulli()`. The divergence is permissive, so brms code ports
  unchanged, but frmtmb code written this way does not port back.
- `ar()`, `ma()`, `arma()`, `cosy()` and `unstr()` are supported in
  their covariance form only; see the next section.
- glmer's proportion-response idiom (`weights = size`) becomes
  `y | trials(size)` with either proportions or counts; `weights()`
  in the formula stays a frequency weight, as in brms.
- `se()` works as in brms (meta-analysis), including
  `se(x, sigma = TRUE)` and the phylogenetic version with
  `gr(g, cov = A)`.
- `mm()` ports with its `weights =` and `scale =` arguments and with
  `mmc()`; its other arguments have different spellings here (below).

## Nonlinear formulas: `nl = TRUE` and `nlf()`

`nl = TRUE` says the RESPONSE formula is a nonlinear body, so only `mu`
can be nonlinear. `nlf()` names the parameter instead, so any of them
can be, and the two spellings meet in the same place:

```r
# the same model, both ways
frm(bf(y ~ exp(b * x), b ~ 1, nl = TRUE), data = d)
frm(bf(y ~ a) + nlf(a ~ exp(b * x)) + lf(b ~ 1), data = d)

# and the model nl = TRUE cannot spell: a nonlinear sigma, linear mu
frm(bf(y ~ x) + nlf(sigma ~ a + b * z) + lf(a ~ 1, b ~ 1), data = d)
```

A parameter's link is applied to its body's value, as in brms: the
`sigma` above is `exp(a + b * z)`. Bodies chain to any depth
(`nlf(a ~ cc * x) + nlf(cc ~ exp(b))`), are computed in dependency
order, and a cycle is refused by name.

Two deliberate differences from brms, both supersets:

- brms requires `nl = TRUE` on the `bf()` whenever the response formula
  names a nonlinear parameter. Here `nlf()` has already declared the
  name, so `bf(y ~ a) + nlf(a ~ ...)` is enough; the flag still works
  and means the same thing.
- A body may name another distributional parameter of the same
  response, which reads that parameter's per-row value. brms reads such
  a name as a data column and refuses the model when no column has it.
  A column still wins here, so a ported body keeps its meaning. This is
  the spelling for a variance function of the model's own mean, which
  nlme writes `varPower(form = ~ fitted(.))`:

```r
# sd = exp(ls) * |mu|^th, one joint maximum likelihood
frm(bf(y ~ x) + nlf(sigma ~ ls + th * log(abs(mu))) +
      lf(ls ~ 1, th ~ 1), data = d)
```

### Distribution functions in a body

brms writes a nonlinear body in Stan, where `Phi()` and the rest belong
to the language. Here the body is R evaluated onto an AD tape, and a
function the body CALLS is looked up in RTMB first:

```r
# a probit-shaped body: no RTMB:: prefix, no library(RTMB)
frm(bf(y ~ a * pnorm(b * x + c0), a ~ 1, b ~ 1, c0 ~ 1, nl = TRUE),
    data = d, start = list(beta = c(1, 1, 0)))
```

`pnorm`, `qgamma`, `plogis`, `dbeta`, `besselK`, `matrix` and their
relatives all resolve this way, and RTMB wins over a function of the
same name that you defined yourself. Write `stats::pnorm()` when you
want the other one. The rule covers only the names RTMB replaces, and
only where the body calls them, so your own helpers and the objects a
body reads are untouched. The `RTMB::` prefix keeps working and gives
the identical fit.

## Multi-membership

`(1 | mm(g1, g2))` means what it means in brms: the observation belongs
to several levels of one grouping factor, and its design row averages
those levels' effects. The pooled level set, the `1/J` default weights
and the `scale = TRUE` normalization of a supplied weight matrix are
brms's, checked against `make_standata()`'s `J_*`, `W_*` and `Z_*`
arrays, so a ported model builds the same design matrix.

`mmc(x1, x2)` also ports: it is one random-slope coefficient of the
term whose covariate value is member specific, member `k` taking
argument `k`.

```r
# brms and frmtmb, identical spelling
frm(bf(y ~ x + (1 | mm(school1, school2))) + gaussian(), data = d)
frm(bf(y ~ x + (1 | mm(school1, school2,
                       weights = cbind(share1, share2)))) + gaussian(),
    data = d)
frm(bf(y ~ (mmc(hours1, hours2) | mm(school1, school2))) + gaussian(),
    data = d)
```

brms's remaining `mm()` arguments are spelled the way any other
grouping term spells them here, and the parser names the replacement
when it refuses one:

| brms | frmtmb |
|---|---|
| `mm(g1, g2, cor = FALSE)` | `diag(x \| mm(g1, g2))`, or `(x \|\| mm(g1, g2))` |
| `mm(g1, g2, id = "q")` | the `\|ID\|` key, `(x \| q \| g)` - not yet over `mm()` |
| `mm(g1, g2, cov = A)` | `gr(g, cov = A)` - not yet over `mm()` |
| `mm(g1, g2, by = )`, `pw =`, `dist =` | no equivalent yet |

`?frmtmb-multimembership` is the full page.

## Within-group residual correlation

`ar()`, `ma()`, `arma()`, `cosy()` and `unstr()` are written where brms
writes them, in the model formula, and mean what brms means under
`cov = TRUE`: the residuals of one group become a single correlated
draw, `y_g ~ N(mu_g, D R D)` with `D` the diagonal of that group's
`sigma` values. `?frmtmb-autocor` has the full page; four things
matter when porting.

**`cov = TRUE` is required for `ar()`, `ma()` and `arma()`.** brms
defaults to `cov = FALSE`, which is a different likelihood - a
residual regression that conditions on each group's first rows - and
that form is not implemented. The call is refused rather than
reinterpreted, so a brms model written with the default has to be
changed deliberately. `cosy()` and `unstr()` have no `cov` argument in
brms either and port unchanged.

```{r, eval = FALSE}
# brms
brm(y ~ x + ar(week, subj, cov = TRUE), data = d)
# frmtmb
frm(bf(y ~ x + ar(week, subj, cov = TRUE)) + gaussian(), data = d)
```

**`sigma` is the marginal residual SD, not the innovation SD.** brms's
`cholesky_cor_ar1()` divides by `1 - ar^2`, so its `sigma` under
`ar()`/`ma()`/`arma()` is the innovation scale, while under `cosy()`
and `unstr()` it is the marginal one. Here every structure uses the
marginal scale, so `sigma()`, pearson residuals and a `sigma ~ x` model
mean one thing throughout. The correlation parameters agree exactly;
convert the scales with `sigma_marginal = sigma_innovation /
sqrt(1 - phi^2)` for AR(1).

**Only `gaussian()` and `student()`.** These are the two families brms
gives a real residual covariance. brms accepts the same spelling for
poisson, binomial and the rest, but fits a latent Gaussian AR process
added to the linear predictor instead - which is a random effect, and
is spelled as one here:

```{r, eval = FALSE}
# brms, poisson: a latent AR process on the linear predictor
brm(cnt ~ x + ar(week, subj), data = d, family = poisson())
# frmtmb: the same idea as a random effect over the time factor
frm(bf(cnt ~ x + ar1(factor(week) + 0 | subj)) + poisson(), data = d)
```

**Higher orders are allowed, and gaps are honored.** brms limits
`cov = TRUE` to order one; `ar(p = 3)` and `arma(p = 2, q = 1)` work
here. And the lag between two rows is the distance between their
positions in the global set of time levels, so a subject missing week 3
gets `cor(week2, week4) = rho^2`; brms indexes `ar()`/`ma()`/`arma()`/
`cosy()` by position within the group and treats the missing row as no
gap. On complete balanced groups the two agree.

The parameters appear in `summary()` under "Within-group residual
correlation" and in `confint_varcorr()` under brms's names (`ar[1]`,
`cosy`, `cortime__1__2`); `autocor_matrix()` returns the fitted
correlation matrix. `weights()`, `cens()`, `trunc()`, `se()`, `mi()`,
`rescor = TRUE`, mixtures and `quadrature = TRUE` are refused, because
the likelihood no longer factorizes over rows - brms refuses the same
core set. Random effects alongside the correlated residual are the
point of the feature and are supported.

## Variance functions (nlme `weights = varFunc`)

There are no `varFunc` objects. The `sigma` formula is the variance
function: `sigma` uses a log link, so most of the nlme vocabulary is a
direct translation, and the translation was verified against
`nlme::gls()` under ML (log-likelihood agreement 1e-10 or better).

| nlme | frmtmb |
|---|---|
| `varIdent(form = ~ 1 \| g)` | `sigma ~ 0 + g` |
| `varExp(form = ~ v)` | `sigma ~ v` |
| `varPower(form = ~ v)` | `sigma ~ log(abs(v))` |
| `varFixed(~ v)` | `sigma ~ offset(0.5 * log(v))` |
| `varComb(A, B)` | the two terms in one `sigma` formula |

The coefficient of `log(abs(v))` is nlme's power `theta`. The
translation is also more general than a `varFunc`: `sigma` takes the
full predictor grammar, so random effects and smooths in the variance
are ordinary spellings.

`varPower(form = ~ fitted(.))`, variance as a function of the model's
own mean, is spelled with `nlf()` (see the nonlinear-formulas
section): `nlf(sigma ~ ls + th * log(abs(mu))) + lf(ls ~ 1, th ~ 1)`
is `sd = exp(ls) * |mu|^th`, estimated by one joint maximum
likelihood. Expect agreement with `gls` near 1e-2 rather than 1e-10:
`gls` alternates the mean fit against iteratively updated fitted
values and stops without a gradient check, while this is one joint
maximization of the same likelihood. For a positive response the
coupling often belongs in the family instead: `Gamma()` has sd
proportional to the mean, and `tweedie()` estimates the power in
`Var = phi * mu^p` directly.

## Structural objects: `data2`

Matrices that are not columns of `data` go in `data2`, the brms
spelling: the adjacency matrix of `car()`, the mesh triple of
`spde()`, and the matrices of `gr(prec = )`, `gr(cov = )` and
`equalto()`.

```{r, eval = FALSE}
frm(bf(y ~ x + car(W, gr = loc, type = "icar")) + gaussian(),
    data = d, data2 = list(W = W))
frm(bf(phen ~ 1 + (1 | gr(species, cov = A))) + gaussian(),
    data = d, data2 = list(A = A))
```

Two differences from brms. First, brms accepts a bare name in `data2`
and nothing else, while frmtmb evaluates the whole expression with
`data2` in front of the data mask, so `gr(g, cov = solve(Q))` finds
`Q` in `data2`. Second, `data2` is optional: a matrix visible from the
formula environment is still found, which is what frmtmb did before
`data2` existed.

Use `data2` anyway. Its objects are stored on the fit, so the fit is
self-contained: `saveRDS()` writes the matrices with it, and
`refit()`, `update()`, `drop1()`, `influence()` and `frm_multiple()`
re-assemble in a fresh session where the calling environment that
built the matrix is gone. Without `data2` those refits depend on the
name still resolving where the model was written.

## Method conventions

frmtmb keeps the brms spelling for functions that originate in brms and
the stats/lme4 spelling for standard generics:

- brms-origin, same names and argument spellings:
  `conditional_effects()` (includes what `conditional_smooths()`
  covers), `hypothesis()` (a delta-method Wald test here),
  `pp_check()` (simulate-based, through bayesplot's `ppc_*`
  functions), `prior_summary()`, `fixef()`, `ranef()`, `VarCorr()`,
  `ngrps()`, `variables()` (the usable parameter names, e.g.
  `sd_Subject__Days`).
- stats-origin generics follow stats/lme4/glmmTMB, not brms:
  `predict()` returns a vector (with `se.fit = TRUE`, a list), not a
  draws matrix with `Estimate`/`Q2.5` columns; `fitted()` and
  `residuals()` return vectors; `confint()` takes `level =`, not
  `probs =`; `re.form` replaces `re_formula`. The exception is an
  ordinal family, where `fitted()` follows brms and returns the `n` by
  `K` matrix of category probabilities named by the response levels,
  exactly as `predict(type = "response")` does; the latent linear
  predictor is `predict(type = "link")`.
- `conditional_effects()` on an ordinal fit draws one probability
  curve per response category, which is what brms draws under
  `categorical = TRUE`, and keys the effect `"x:cats__"` as brms does
  there. It is the default here because brms's own default warns that
  it is treating an ordered factor as continuous; `categorical = FALSE`
  asks for that summary, the expected category number.
- `coef()` is the per-group convention shared by all three packages:
  fixed effects plus conditional modes per grouping level.
- `plot(fit)` draws residual diagnostics, not MCMC traces; for
  simulation-based residual checks use `dharma_residuals()`.

## Seeing the model code

brms users inspect the generated model with `stancode(fit)`. The
frmtmb analog is the objective function itself: an ordinary R closure
that `RTMB::MakeADFun()` tapes, rebuilt deterministically from the
fitted frame.

```{r, eval = FALSE}
nll <- frmtmb:::build_objective(fit$frame)
nll(fit$estimates)   # joint negative log density at the estimates
```

Two properties matter when reading it:

- The closure takes the named parameter list (the shape of
  `fit$frame$par_template`) and returns the JOINT negative log
  density: random-effect prior terms are inside it. The marginal
  (Laplace) likelihood the optimizer minimized is what
  `MakeADFun(random = )` makes of it, exposed as `fit$obj$fn` over
  the flat outer-parameter vector; at the optimum
  `fit$obj$fn(fit$opt$par)` equals `-logLik(fit)`.
- All data is baked into the closure's environment (the reason
  refits re-tape instead of recompiling), so `environment(nll)`
  holds the assembled frame, and the closure evaluates on plain
  numeric parameter lists without any tape.

`RTMB::MakeADFun(nll, fit$frame$par_template, random = "b",
map = fit$frame$map)` reproduces `fit$obj`. There is no generated
code to print: the closure IS the model, and `body()`,
`environment()`, and the debugger work on it like on any R function.

## The posterior half of the port

This page ports the model grammar and the frequentist workflow built
around it. The sampling workflow ports as well, into the companion
package **frmtmb.sample**. `frm_sample()` takes the same `bf()` object
this page describes, assembles it through the same code path, and runs
Stan's NUTS on the frmtmb objective under brms's default priors; the
draws it returns carry the brms post-processing surface, with the real
`loo()` on it. `as_tmbstan()` hands the objective to tmbstan directly.

Nothing in the grammar changes when you sample, which is why the two
halves split cleanly. What does change is everything around the draws:
the argument names of the call, the parameter spellings, and which
`brmsfit` methods have no counterpart.
`vignette("brms-posterior", package = "frmtmb.sample")` is that half,
and `vignette("posterior-diagnostics", package = "frmtmb.sample")`
covers the diagnostics of a sampled fit.

## When you still want brms

With **frmtmb.sample** installed, a posterior, brms's default priors
and `loo()`-based comparison are all available here, and
**frmtmb.latent** adds hidden Markov and latent class families, so the
reasons to reach for brms itself are narrower than they look:

- a marginal likelihood, and so a Bayes factor between models
  that are not nested. A point null INSIDE one model is now
  answered here: `hypothesis()` on a draws object reports the
  Savage-Dickey evidence ratio and posterior probability that
  brms reports, for a population-level coefficient whose prior is
  written about the coefficient itself (in practice a `class = "b"`
  prior you wrote). It refuses, by name, on a variance component, a
  correlation, a dispersion and the intercept, because each of those
  is reported on a scale its prior is not written on. What still
  needs brms is bridge sampling, which estimates the marginal
  likelihood itself;
- `reloo()` or `kfold()` on a model whose LOO approximation fails,
  since both need refits that this package does not do;
- a family or term neither package here implements, which
  `frm_compat()` answers per feature;
- Stan code you intend to read, extend or hand to somebody else.

A practical workflow is model screening with `frm()`, which costs
seconds, sampling the survivors with `frm_sample()`, and a final fit in
brms for the questions in that list.
