---
title: "frmtmb for brms users"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{frmtmb for brms users}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

frmtmb reimplements a documented subset of the brms grammar with
identical spelling, so brms model code ports mechanically: drop the
priors, change `brm()` to `frm()`, and get a maximum-likelihood fit in
milliseconds-to-seconds instead of an MCMC run. Estimation is ML (or
REML) with the Laplace approximation for latent effects; inference is
frequentist (Wald and profile intervals, likelihood-ratio tests, AIC).

## What ports directly

| brms | frmtmb | notes |
|---|---|---|
| `bf(y ~ x + (x \| g))` | same | plus `us()`, `diag()`, `homdiag()`, `cs()`, `ar1()` wrappers |
| `bf(y ~ x, sigma ~ z + (1 \| g))` | same | any dpar, full grammar |
| `bf(y ~ x, sigma = 1)` | same | fixed via TMB `map` |
| `s(x)`, `t2(x, z)` | same | mgcv smooths in any dpar; `te()`/`ti()` unsupported |
| `(1 \| p \| g)` | same | cross-formula RE correlation |
| `mvbind(y1, y2) ~ x`, `mvbf()`, `set_rescor()` | same | per-response families; `rescor` gaussian-only |
| `y \| trials(n)`, `weights(w)`, `cens(c)`, `trunc(lb=, ub=)` | same | `cens`/`trunc` need a CDF-carrying family |
| `nl = TRUE` | same | provide `start`; se.fit on the nonlinear mu not yet |
| `custom_family()` | same idea | the lpdf is plain R over RTMB advectors, not Stan code |
| `cumulative()`, `multinomial()` | same | multinomial takes `K` explicitly |

## What changes

- `brm(...)` becomes `frm(...)`; there are no priors, chains, or
  warmup. `REML = TRUE` replaces priors as the small-sample correction
  for variance components.
- Posterior summaries become ML estimates: `fixef()` returns point
  estimates, `confint()` gives Wald or profile intervals, `anova()`
  gives likelihood-ratio tests, and `AIC()`/`BIC()` replace `loo()`.
- `posterior_predict()` becomes `simulate()`; `posterior_epred()` over
  `newdata` becomes `predict(newdata, type = "response")`. On a
  `cens()` response both draw the LATENT uncensored value, as brms
  does; `simulate(censored = TRUE)` applies the censoring mechanism
  instead.
- `sample_prior = "only"` plus `posterior_predict()` becomes
  `frm_simulate(formula, data, priors = set_prior(...))`, which draws
  a parameter vector per simulation and returns it alongside the
  responses.
- `mo()` monotonic effects, `mi()` one-step imputation of continuous
  predictors, `mi(sdx)` measurement error (the `me()` replacement,
  as in current brms), `cs()` category-specific ordinal effects,
  `gp(x)` / `gp(x, k =)` Gaussian processes, and `mixture()`
  families (including group-level latent classes via `groups = ~g`)
  all work. `mo()`/`mi()` support two-way `:`/`*` interactions with
  numeric terms; discrete predictors cannot be imputed in-model, the
  same restriction Stan has. Multiply imputed data can instead go
  through `frm_multiple()` (Rubin's rules), the `brm_multiple()`
  analog. Being frequentist, that fit also needs pooled tests, which
  brms gets for free from the posterior: `hypothesis()` pools any
  function of the parameters, and `anova()` on the result compares two
  nested `frm_multiple()` fits with the D1, D2 and D3 rules of `mice`
  (D3 by default). Mixture fits are ML: expect multimodality, compare
  starts (`frm_allfit()`).
- `binomial()` without `trials()` is accepted and means Bernoulli, the
  `stats::glm()` convention; brms rejects it and asks for `trials()` or
  `bernoulli()`. The divergence is permissive, so brms code ports
  unchanged, but frmtmb code written this way does not port back.
- Not supported: `ar()/ma()` residual autocorrelation terms (use the
  `ar1()` random-effect structure).
- glmer's proportion-response idiom (`weights = size`) becomes
  `y | trials(size)` with either proportions or counts; `weights()`
  in the formula stays a frequency weight, as in brms.
- `se()` works as in brms (meta-analysis), including
  `se(x, sigma = TRUE)` and the phylogenetic version with
  `gr(g, cov = A)`.

## Method conventions

frmtmb keeps the brms spelling for functions that originate in brms and
the stats/lme4 spelling for standard generics:

- brms-origin, same names and argument spellings:
  `conditional_effects()` (includes what `conditional_smooths()`
  covers), `hypothesis()` (a delta-method Wald test here),
  `pp_check()` (simulate-based, through bayesplot's `ppc_*`
  functions), `prior_summary()`, `fixef()`, `ranef()`, `VarCorr()`,
  `ngrps()`, `variables()` (the usable parameter names, e.g.
  `sd_Subject__Days`).
- stats-origin generics follow stats/lme4/glmmTMB, not brms:
  `predict()` returns a vector (with `se.fit = TRUE`, a list), not a
  draws matrix with `Estimate`/`Q2.5` columns; `fitted()` and
  `residuals()` return vectors; `confint()` takes `level =`, not
  `probs =`; `re.form` replaces `re_formula`.
- `coef()` is the per-group convention shared by all three packages:
  fixed effects plus conditional modes per grouping level.
- `plot(fit)` draws residual diagnostics, not MCMC traces; for
  simulation-based residual checks use `dharma_residuals()`.

## Seeing the model code

brms users inspect the generated model with `stancode(fit)`. The
frmtmb analog is the objective function itself: an ordinary R closure
that `RTMB::MakeADFun()` tapes, rebuilt deterministically from the
fitted frame.

```{r, eval = FALSE}
nll <- frmtmb:::build_objective(fit$frame)
nll(fit$estimates)   # joint negative log density at the estimates
```

Two properties matter when reading it:

- The closure takes the named parameter list (the shape of
  `fit$frame$par_template`) and returns the JOINT negative log
  density: random-effect prior terms are inside it. The marginal
  (Laplace) likelihood the optimizer minimized is what
  `MakeADFun(random = )` makes of it, exposed as `fit$obj$fn` over
  the flat outer-parameter vector; at the optimum
  `fit$obj$fn(fit$opt$par)` equals `-logLik(fit)`.
- All data is baked into the closure's environment (the reason
  refits re-tape instead of recompiling), so `environment(nll)`
  holds the assembled frame, and the closure evaluates on plain
  numeric parameter lists without any tape.

`RTMB::MakeADFun(nll, fit$frame$par_template, random = "b",
map = fit$frame$map)` reproduces `fit$obj`. There is no generated
code to print: the closure IS the model, and `body()`,
`environment()`, and the debugger work on it like on any R function.

## When you still want brms

Genuine prior information, full posterior uncertainty for derived
quantities, models with discrete latent structure beyond
observation-level mixtures, or `loo`-based model comparison. A
practical workflow is model screening in frmtmb and a final fit in
brms; `as_tmbstan(fit)` also runs NUTS on the frmtmb objective
directly.
