---
title: "Case studies: models from the showcase literature"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case studies: models from the showcase literature}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb)
has_ape <- requireNamespace("ape", quietly = TRUE)
has_nlme <- requireNamespace("nlme", quietly = TRUE)
has_phy <- has_ape && has_nlme
has_metafor <- requireNamespace("metafor", quietly = TRUE)
has_tinyplot <- requireNamespace("tinyplot", quietly = TRUE)
```

Each section below takes a model that another package uses to show what
it can do, and fits it with `frm()`. Every section says where the model
comes from, gives the frmtmb call, reads the output, and then checks the
numbers against a package or a closed form that computes the same thing
a different way. The datasets are small so that this page rebuilds in
seconds.

Four sections carry a figure. The figures use
[tinyplot](https://cran.r-project.org/package=tinyplot), which is a
suggested package, and they are skipped when it is not installed.

This is not the getting-started page. See `vignette("frmtmb")` for the
grammar and `vignette("brms-migration")` for the brms feature map.

## 1. The animal model

The animal model is the flagship example of MCMCglmm (Hadfield 2010,
*Journal of Statistical Software* 33(2); see also Wilson et al. 2010,
*Journal of Animal Ecology* 79, "An ecologist's guide to the animal
model"). It splits phenotypic variance into an additive genetic part
and a residual part. Each individual gets one random effect, and those
effects are correlated through the additive relationship matrix `A`,
which the pedigree determines. Two individuals with relatedness 0.5,
such as full sibs, share half of their additive genetic deviation in
expectation.

The relatedness matrix is not a column of `data`, so it travels in
`data2`, exactly as in brms. `gr(id, cov = A)` attaches it to the
grouping factor.

```{r ped}
# a half-sib design: 12 sires, 2 dams each, 6 offspring per pair
nsire <- 12; ndam_per <- 2; noff <- 6
nfound <- nsire + nsire * ndam_per
n <- nfound + nsire * ndam_per * noff
ped <- data.frame(id = 1:n, sire = NA_integer_, dam = NA_integer_)
k <- nfound
for (s in 1:nsire) for (j in 1:ndam_per) {
  dam <- nsire + ndam_per * (s - 1) + j
  for (o in 1:noff) { k <- k + 1; ped$sire[k] <- s; ped$dam[k] <- dam }
}

# the tabular method builds A from a pedigree sorted parents-first
A <- diag(n)
for (i in seq_len(n)) {
  s <- ped$sire[i]; d <- ped$dam[i]
  if (!is.na(s)) {
    A[i, i] <- 1 + 0.5 * A[s, d]
    for (j in seq_len(i - 1)) A[i, j] <- A[j, i] <- 0.5 * (A[j, s] + A[j, d])
  }
}
dimnames(A) <- list(as.character(ped$id), as.character(ped$id))

set.seed(9)
dat <- data.frame(id = factor(ped$id, levels = ped$id),
                  sex = factor(rep(c("f", "m"), length.out = n)))
# additive genetic values with sd 1, residual sd 1, so h2 = 0.5
dat$phen <- 5 + 0.4 * (dat$sex == "m") +
  as.numeric(t(chol(A)) %*% rnorm(n)) + rnorm(n, 0, 1)
```

`REML = TRUE` is the usual choice here, because the variance components
are the estimands and the pedigree is small.

```{r animal}
fit <- frm(bf(phen ~ sex + (1 | gr(id, cov = A))) + gaussian(),
           data = dat, data2 = list(A = A), REML = TRUE,
           control = frmtmb_control(check_olre = "ignore"))
sd_a <- sqrt(VarCorr(fit)[[1]][1, 1])
c(sd_additive = sd_a, sigma = sigma(fit),
  heritability = sd_a^2 / (sd_a^2 + sigma(fit)^2))
```

The narrow-sense heritability is the intraclass correlation of the
genetic effect, `sd_a^2 / (sd_a^2 + sigma^2)`. The estimate is near the
simulated 0.5.

One warning needs a comment. Every individual appears once, so `id` has
one level per row. frmtmb warns that such a term is confounded with the
residual. That warning is a false positive when the block carries a
`cov =` matrix, because `A` is not the identity and the two variances
are then separately identified. `check_olre = "ignore"` turns it off.

### Cross-check against the closed-form REML likelihood

For a gaussian animal model the marginal distribution is known in
closed form: `y ~ N(X beta, sd_a^2 A + sigma^2 I)`. The restricted log
likelihood of that model can be written in six lines and maximized with
`optim()`. It uses no part of frmtmb.

```{r animal-check}
X <- model.matrix(~ sex, dat); y <- dat$phen
nll <- function(p) {
  V <- exp(p[1]) * A + exp(p[2]) * diag(n)
  Vi <- solve(V); M <- crossprod(X, Vi) %*% X
  b <- solve(M, crossprod(X, Vi) %*% y); r <- y - X %*% b
  as.numeric(0.5 * ((n - ncol(X)) * log(2 * pi) + determinant(V)$modulus +
                      determinant(M)$modulus + crossprod(r, Vi %*% r)))
}
op <- optim(c(0, 0), nll, method = "BFGS", control = list(reltol = 1e-13))
out <- rbind(frmtmb = c(sd_a, sigma(fit), as.numeric(logLik(fit))),
             closed_form = c(sqrt(exp(op$par)), -op$value))
colnames(out) <- c("sd_additive", "sigma", "restricted_logLik")
out
```

The variance components and the restricted log likelihood agree to six
significant digits.

### An interval for heritability

Heritability is a ratio of variance components, so an interval for it
needs more than the table of standard errors. There are two routes.
`hypothesis()` gives the Wald delta-method version, and it reads the
random-effect standard deviation under its natural-scale name
`sd_<group>__<term>`:

```{r animal-hyp, eval = FALSE}
hypothesis(fit, "sd_id__Intercept^2 / (sd_id__Intercept^2 + sigma^2)")
```

A variance ratio is bounded on `[0, 1]` and skewed, so a symmetric
delta-method interval is the weaker choice near either boundary. The
parametric bootstrap gives a percentile interval instead.
`frm_bootstrap()` takes any function of a refit.

```{r animal-boot}
bs <- frm_bootstrap(fit, nsim = 30, seed = 1, FUN = function(f) {
  v <- VarCorr(f)[[1]][1, 1]
  c(h2 = v / (v + sigma(f)^2))
})
confint(bs)
```

Thirty draws is a vignette-sized number. Use several hundred for real
work.

### Two traits at once

MCMCglmm's multi-response animal model estimates the genetic
correlation between traits. Write it in long format. Stack the traits
into one response column, and give the pedigree block one random
coefficient per trait. The covariance of that block is then the genetic
matrix `G`, and `cov = A` makes the full covariance the Kronecker
product of `G` with the relatedness matrix. `sigma ~ 0 + trait` gives
each trait its own residual standard deviation.

The `mvbf()` spelling with the `|ID|` identifier over the same
`gr(id, cov = A)` term fits the same joint density: the cross-formula
merge builds one Kronecker block, and the two spellings agree in the
log likelihood to about 1e-9. The long format shown here keeps both
traits in one formula, which is the shorter call.

```{r animal-mv-data}
set.seed(4)
G <- matrix(c(1.0, 0.6, 0.6, 0.8), 2, 2)
U <- t(chol(A)) %*% matrix(rnorm(n * 2), n, 2) %*% chol(G)
long <- data.frame(
  id = factor(rep(ped$id, times = 2), levels = ped$id),
  trait = factor(rep(c("y1", "y2"), each = n)),
  value = c(3 + U[, 1] + rnorm(n, 0, 0.7),
            1 + U[, 2] + rnorm(n, 0, 0.9)))
```

```{r animal-mv}
fmv <- frm(bf(value ~ 0 + trait + (0 + trait | gr(id, cov = A)),
              sigma ~ 0 + trait) + gaussian(),
           data = long, data2 = list(A = A))
Gh <- VarCorr(fmv)[[1]]
out <- rbind(
  estimated = c(sqrt(diag(Gh)), cov2cor(Gh)[1, 2], exp(fixef(fmv)$sigma)),
  simulated = c(sqrt(diag(G)), 0.6 / sqrt(1.0 * 0.8), 0.7, 0.9))
colnames(out) <- c("gen_sd_y1", "gen_sd_y2", "gen_cor",
                   "resid_sd_y1", "resid_sd_y2")
round(out, 4)
```

The genetic correlation comes out at 0.68 against a simulated 0.67. The
genetic standard deviations are low and the residual ones are high,
which is the usual trade in a pedigree this small: the two parts of the
variance are hard to separate, and 180 individuals is not many.

### Checking that the pedigree is doing work

A known-covariance model has one silent failure mode: the matrix is
accepted, ignored, and the fit still converges. The cheap guard is to
refit with the identity in place of `A`. If the pedigree carries
information, the log likelihood must move.

```{r animal-mv-sens}
I <- diag(n); dimnames(I) <- dimnames(A)
fmv_id <- frm(bf(value ~ 0 + trait + (0 + trait | gr(id, cov = I)),
                 sigma ~ 0 + trait) + gaussian(),
              data = long, data2 = list(I = I))
c(with_pedigree = as.numeric(logLik(fmv)),
  with_identity = as.numeric(logLik(fmv_id)),
  difference = as.numeric(logLik(fmv)) - as.numeric(logLik(fmv_id)))
stopifnot(as.numeric(logLik(fmv)) - as.numeric(logLik(fmv_id)) > 5)
```

The pedigree buys about 19 log-likelihood points over an unrelated
population. Run this check whenever you pass a matrix through `data2`.
It is two lines, and it is the only thing that separates a covariance
that is used from a covariance that is merely accepted.

## 2. Phylogenetic regression

The brms phylogenetics vignette fits the same structure with a
different matrix. Species share ancestry, so their residual departures
from a regression line are correlated by the phylogeny. Under Brownian
motion that correlation is the shared root-to-tip path length, which
`ape::vcv()` returns. The model is then a mixed model whose grouping
factor is the species and whose covariance is the phylogenetic
correlation matrix (Housworth, Martins and Lynch 2004; de Villemereuil
and Nakagawa 2014).

```{r phylo-data, eval = has_phy}
set.seed(7)
tree <- ape::rcoal(60)
tree$tip.label <- paste0("sp", 1:60)
# scale the tree to unit depth, so vcv() is a correlation matrix
tree$edge.length <- tree$edge.length / max(ape::node.depth.edgelength(tree))
A_phy <- ape::vcv(tree)

d <- data.frame(sp = factor(tree$tip.label, levels = tree$tip.label),
                x = rnorm(60))
d$y <- 0.5 + 0.8 * d$x +
  as.numeric(t(chol(A_phy)) %*% rnorm(60)) + rnorm(60, 0, 0.6)
```

```{r phylo-fit, eval = has_phy}
fphy <- frm(bf(y ~ x + (1 | gr(sp, cov = A_phy))) + gaussian(),
            data = d, data2 = list(A_phy = A_phy),
            control = frmtmb_control(check_olre = "ignore"))
sd_p <- sqrt(VarCorr(fphy)[[1]][1, 1])
c(fixef(fphy)$mu, sd_phylo = sd_p, sigma = sigma(fphy),
  phylogenetic_h2 = sd_p^2 / (sd_p^2 + sigma(fphy)^2))
```

The last quantity is the phylogenetic heritability, also called Pagel's
lambda. It is the share of the residual variance that the tree explains.

### Cross-check against PGLS

Phylogenetic generalized least squares fits the same likelihood in a
different parameterization. `ape::corPagel` gives `nlme::gls()` the
correlation matrix `lambda * C + (1 - lambda) * I`. For a tree scaled to
unit depth that matrix is the frmtmb model's marginal correlation, so
the two fits are the same model and must agree.

```{r phylo-check, eval = has_phy}
g <- nlme::gls(y ~ x, data = d, method = "ML",
               correlation = ape::corPagel(0.5, phy = tree, form = ~sp))
out <- rbind(
  frmtmb = c(fixef(fphy)$mu, sd_p^2 / (sd_p^2 + sigma(fphy)^2),
             sd_p^2 + sigma(fphy)^2, as.numeric(logLik(fphy))),
  gls = c(coef(g), coef(g$modelStruct$corStruct, unconstrained = FALSE),
          g$sigma^2, as.numeric(logLik(g))))
colnames(out) <- c("intercept", "x", "lambda", "total_var", "logLik")
out
```

```{r phylo-assert, eval = has_phy}
lam_gls <- as.numeric(coef(g$modelStruct$corStruct, unconstrained = FALSE))
stopifnot(
  max(abs(fixef(fphy)$mu - coef(g))) < 1e-4,
  abs(sd_p^2 / (sd_p^2 + sigma(fphy)^2) - lam_gls) < 1e-4,
  abs(as.numeric(logLik(fphy)) - as.numeric(logLik(g))) < 1e-4
)
```

Use `REML = TRUE` and `method = "REML"` together if you want the
restricted versions. The frmtmb form is the more general one: it takes
covariates in `sigma`, extra grouping factors, and non-gaussian
families, none of which `gls()` offers.

## 3. Meta-analysis and meta-regression

A random-effects meta-analysis is a mixed model in which every
observation carries a known standard error. The `se()` addition term
supplies it, which is the same spelling brms uses. The residual
standard deviation is fixed at zero, so the only free variance is the
between-study heterogeneity `tau`.

The data are the 13 BCG vaccine trials of Colditz et al. (1994), the
running example of the metafor package (Viechtbauer 2010, *Journal of
Statistical Software* 36(3)). The effect measure is the log risk ratio
of tuberculosis in the vaccinated group.

```{r bcg}
bcg <- data.frame(
  tpos = c(4, 6, 3, 62, 33, 180, 8, 505, 29, 17, 186, 5, 27),
  tneg = c(119, 300, 228, 13536, 5036, 1361, 2537, 87886, 7470, 1699,
           50448, 2493, 16886),
  cpos = c(11, 29, 11, 248, 47, 372, 10, 499, 45, 65, 141, 3, 29),
  cneg = c(128, 274, 209, 12619, 5761, 1079, 619, 87892, 7232, 1600,
           27197, 2338, 17825),
  ablat = c(44, 55, 42, 52, 13, 44, 19, 13, 27, 42, 18, 33, 33))
bcg$yi <- log((bcg$tpos / (bcg$tpos + bcg$tneg)) /
                (bcg$cpos / (bcg$cpos + bcg$cneg)))
bcg$vi <- 1 / bcg$tpos - 1 / (bcg$tpos + bcg$tneg) +
  1 / bcg$cpos - 1 / (bcg$cpos + bcg$cneg)
bcg$sei <- sqrt(bcg$vi)
bcg$study <- factor(seq_len(nrow(bcg)))
```

```{r meta}
fmeta <- frm(bf(yi | se(sei) ~ 1 + (1 | study)) + gaussian(),
             data = bcg, REML = TRUE)
c(pooled_logRR = unname(fixef(fmeta)$mu), se = sqrt(vcov(fmeta)[1, 1]),
  tau = sqrt(VarCorr(fmeta)[[1]][1, 1]))
```

The pooled log risk ratio is about -0.71, which is a risk ratio near
0.49. Between-study heterogeneity is large: `tau` is comparable in size
to the pooled effect itself. `confint_varcorr()` puts an interval on it.

A forest plot puts the pooled estimate next to the studies it pools.
Each study is a point with a 95 percent interval from its own known
standard error. The dashed line is the pooled log risk ratio and the
dotted line is no effect.

```{r meta-forest, eval = has_tinyplot, fig.width = 7, fig.height = 4.2, fig.alt = "Forest plot of the 13 BCG vaccine trials. Study number runs up the vertical axis and the log risk ratio runs along the horizontal axis. Each study is a black point with a horizontal interval. A dashed red vertical line marks the pooled estimate near minus 0.71. A dotted gray line marks zero. Most studies fall to the left of zero. The intervals differ widely in width, and several short intervals miss the pooled line on one side or the other."}
pooled <- unname(fixef(fmeta)$mu)
tinyplot::tinyplot(yi ~ study, data = bcg,
                   ymin = bcg$yi - 1.96 * bcg$sei,
                   ymax = bcg$yi + 1.96 * bcg$sei,
                   type = "pointrange", flip = TRUE, theme = "clean2",
                   col = "black", pch = 16,
                   xlab = "study", ylab = "log risk ratio")
abline(v = pooled, lty = 2, col = "firebrick")
abline(v = 0, lty = 3, col = "gray50")
```

The picture explains the large `tau`. Studies 8 and 11 have short
intervals that stay above the pooled line, and studies 4 and 10 have
short intervals that stay below it. No single effect size can hold all
13, so the random-effect term takes up the difference.

```{r meta-varcorr}
confint_varcorr(fmeta)
```

### Cross-check against metafor

`metafor::rma()` is the reference implementation of this model, and it
uses REML by default.

```{r meta-check, eval = has_metafor}
rr <- metafor::rma(yi, vi, data = bcg, method = "REML")
out <- rbind(frmtmb = c(unname(fixef(fmeta)$mu), sqrt(vcov(fmeta)[1, 1]),
                        sqrt(VarCorr(fmeta)[[1]][1, 1])),
             metafor = c(as.numeric(rr$beta), rr$se, sqrt(rr$tau2)))
colnames(out) <- c("pooled_logRR", "se", "tau")
out
```

```{r meta-assert, eval = has_metafor}
stopifnot(
  abs(fixef(fmeta)$mu - as.numeric(rr$beta)) < 1e-5,
  abs(sqrt(VarCorr(fmeta)[[1]][1, 1]) - sqrt(rr$tau2)) < 1e-5,
  # the standard errors come from different expressions: metafor uses
  # (X'WX)^-1 at the REML tau, frmtmb reads the joint Hessian
  abs(sqrt(vcov(fmeta)[1, 1]) / rr$se - 1) < 0.01
)
```

The pooled estimate and the heterogeneity match to five decimal places.
The standard errors agree to under one percent, because the two
packages compute them from different expressions.

### Meta-regression

Adding a moderator makes it a meta-regression. Absolute latitude is the
classic moderator for these trials: the vaccine works better away from
the equator.

```{r meta-reg}
freg <- frm(bf(yi | se(sei) ~ ablat + (1 | study)) + gaussian(),
            data = bcg, REML = TRUE)
c(fixef(freg)$mu, tau = sqrt(VarCorr(freg)[[1]][1, 1]))
```

```{r meta-reg-check, eval = has_metafor}
rr2 <- metafor::rma(yi, vi, mods = ~ablat, data = bcg, method = "REML")
out <- rbind(frmtmb = c(fixef(freg)$mu, sqrt(VarCorr(freg)[[1]][1, 1])),
             metafor = c(coef(rr2), sqrt(rr2$tau2)))
colnames(out) <- c("intercept", "ablat", "tau")
out
```

Latitude removes about half of the heterogeneity: `tau` falls from 0.56
to 0.28.

Because this is an ordinary mixed model with a fixed residual, the rest
of the grammar still applies. Nested study levels become a second
grouping factor, which is the multilevel meta-analysis. A formula on
`sigma` turns the fixed residual into a modeled one, which is
`se(sei, sigma = TRUE)` in brms.

## 4. Monotonic effects of ordinal predictors

An ordinal predictor such as an income band or a Likert item is neither
a number nor an unordered factor. Treating it as a number forces equal
steps. Treating it as a factor throws the order away. `mo()` keeps the
order and estimates the step sizes as a simplex, which is the method of
Bürkner and Charpentier (2020, *British Journal of Mathematical and
Statistical Psychology* 73), and the subject of the brms monotonic
effects vignette.

```{r mo-data}
set.seed(21)
n <- 400; L <- 6
inc <- sample(1:L, n, TRUE)
shape <- c(0, 0.15, 0.3, 0.75, 0.9, 1)   # a step between bands 3 and 4
dmo <- data.frame(income = factor(inc, ordered = TRUE), z = rnorm(n))
dmo$ls <- 5 + 2.0 * shape[inc] + 0.4 * dmo$z + rnorm(n, 0, 0.8)
```

```{r mo-fit}
fmo <- frm(bf(ls ~ mo(income) + z) + gaussian(), data = dmo)
fixef(fmo)$mu
```

`moincome` is the average step, not the total effect. The total effect
across the whole scale is `moincome * (L - 1)`, here about 1.8. To read
the shape, predict at each band with the other predictors held fixed
and rescale to the unit interval.

```{r mo-shape}
nd <- data.frame(income = factor(1:L, ordered = TRUE), z = 0)
p <- predict(fmo, newdata = nd)
rbind(estimated = round((p - p[1]) / (p[L] - p[1]), 3),
      simulated = shape)
```

The fit finds the large step between bands 3 and 4 and the small steps
elsewhere.

### Cross-check against the saturated factor model

`mo()` is the saturated factor model with an order restriction, and it
spends the same number of parameters. When the band means already
increase, the restriction is not active and the two fits are identical.

```{r mo-check}
fsat <- frm(bf(ls ~ income + z) + gaussian(), data = dmo)
out <- rbind(
  mo = c(predict(fmo, newdata = nd), logLik(fmo), attr(logLik(fmo), "df")),
  saturated = c(predict(fsat, newdata = nd), logLik(fsat),
                attr(logLik(fsat), "df")))
colnames(out) <- c(paste0("band", 1:L), "logLik", "df")
round(out, 4)
stopifnot(max(abs(predict(fmo, newdata = nd) -
                    predict(fsat, newdata = nd))) < 1e-4)
```

The step shape is easier to read as a picture than as a row of numbers.
The points are the observed band means with 95 percent intervals, the
line is the `mo()` fit at `z = 0`, and the crosses are the saturated
factor fit at the same place.

```{r mo-fig, eval = has_tinyplot, fig.width = 6, fig.height = 4, fig.alt = "Plot of the mean of ls against income band, for bands 1 to 6. Six black points with vertical intervals are the observed band means. A red line joins the monotonic model fit, and blue crosses mark the saturated factor fit. The line and the crosses sit on top of each other. The curve climbs gently from band 1 to band 3, jumps between bands 3 and 4, then flattens."}
band <- as.numeric(tapply(dmo$ls, dmo$income, mean))
band_se <- as.numeric(tapply(dmo$ls, dmo$income,
                             function(v) sd(v) / sqrt(length(v))))
tinyplot::tinyplot(x = 1:L, y = band, ymin = band - 1.96 * band_se,
                   ymax = band + 1.96 * band_se, type = "pointrange",
                   theme = "clean2", col = "gray20", pch = 16,
                   xlab = "income band", ylab = "mean of ls")
tinyplot::plt_add(x = 1:L, y = as.numeric(p), type = "l",
                  col = "firebrick", lwd = 2)
tinyplot::plt_add(x = 1:L, y = as.numeric(predict(fsat, newdata = nd)),
                  type = "p", pch = 4, cex = 1.6, col = "steelblue")
```

The big step between bands 3 and 4 is the shape the simulation put
there, and the two fits are the same curve.

The value of `mo()` is not the parameter count. It is that one
coefficient carries the effect size, that the restriction holds when
the data are noisy, and that interactions stay readable. `mo()` takes
two-way interactions with numeric terms.

```{r mo-interaction}
fixef(frm(bf(ls ~ mo(income) * z) + gaussian(), data = dmo))$mu
```

The interaction is near zero here, which is right: the simulation has
no interaction. A factor multiplier is refused, so expand a factor into
numeric indicator columns first.

## 5. Location-scale regression

GAMLSS calls it a location-scale model, mgcv calls it `gaulss`, and
brms calls it distributional regression. The mean and the residual
standard deviation both follow smooth functions of a covariate. In
frmtmb this is one `bf()` with a formula for `sigma`, and the smoothing
parameters of both smooths are estimated as variance components.

```{r ls-data}
set.seed(22)
dls <- data.frame(x = runif(400))
dls$y <- rnorm(400, sin(2 * pi * dls$x),
               exp(-1 + 1.2 * cos(2 * pi * dls$x)))
```

```{r ls-fit}
fls <- frm(bf(y ~ s(x, k = 10), sigma ~ s(x, k = 10)) + gaussian(),
           data = dls)
VarCorr(fls)
```

The two standard deviations are the smooths' wiggliness parameters. A
value near zero would mean a straight line.

### Cross-check against mgcv

`mgcv::gam()` with `family = gaulss()` and `method = "ML"` fits the same
model. It uses a `logb` link for the precision, so its scale
parameterization differs slightly from the plain log link here.

```{r ls-check}
gm <- mgcv::gam(list(y ~ s(x, k = 10), ~ s(x, k = 10)), data = dls,
                family = mgcv::gaulss(b = 0), method = "ML")
nd <- data.frame(x = seq(0.05, 0.95, length.out = 6))
pg <- predict(gm, newdata = nd, type = "response")
out <- rbind(frmtmb_mu = predict(fls, newdata = nd),
             mgcv_mu = pg[, 1],
             frmtmb_sigma = predict(fls, newdata = nd, dpar = "sigma",
                                    type = "response"),
             mgcv_sigma = 1 / pg[, 2],
             true_sigma = exp(-1 + 1.2 * cos(2 * pi * nd$x)))
colnames(out) <- paste0("x=", round(nd$x, 2))
round(out, 4)
```

```{r ls-assert}
sg <- predict(fls, newdata = nd, dpar = "sigma", type = "response")
stopifnot(
  max(abs(predict(fls, newdata = nd) - pg[, 1])) < 0.02,
  max(abs(sg * pg[, 2] - 1)) < 0.03
)
```

The two mean curves agree to about 0.01, and the two scale curves to
about two percent. They are not expected to agree exactly, because the
packages choose their smoothing parameters with different criteria.
Note that `predict()` on a distributional parameter returns the link
scale by default, so `type = "response"` is needed for a standard
deviation.

One figure holds both curves. The band is the frmtmb mean plus and
minus two fitted standard deviations, so its center is the location
model and its width is the scale model. The dashed lines are the same
band from mgcv.

```{r ls-fig, eval = has_tinyplot, fig.width = 7, fig.height = 4.4, fig.alt = "Plot of y against x on the unit interval. A shaded blue band runs across the plot with a blue center line. The band is wide at both ends and narrow near x equal to 0.6. Gray points are the 400 observations, and they crowd where the band is narrow and spread where it is wide. Dashed red lines from mgcv lie on the edges of the band."}
xg <- data.frame(x = seq(0, 1, length.out = 100))
mu_f <- as.numeric(predict(fls, newdata = xg))
sd_f <- as.numeric(predict(fls, newdata = xg, dpar = "sigma",
                           type = "response"))
pg <- predict(gm, newdata = xg, type = "response")
tinyplot::tinyplot(x = xg$x, y = mu_f, ymin = mu_f - 2 * sd_f,
                   ymax = mu_f + 2 * sd_f, type = "ribbon",
                   theme = "clean2", xlab = "x", ylab = "y",
                   ylim = range(dls$y))
tinyplot::plt_add(x = dls$x, y = dls$y, type = "p", pch = 16, cex = 0.5,
                  col = "gray45")
lines(xg$x, mu_f, lwd = 2, col = "steelblue4")
lines(xg$x, pg[, 1] - 2 / pg[, 2], lty = 2, col = "firebrick")
lines(xg$x, pg[, 1] + 2 / pg[, 2], lty = 2, col = "firebrick")
```

The band is narrow where `cos(2 pi x)` is smallest, near `x = 0.5`, and
it opens at both ends. A constant-variance smooth would draw a band of
one width and would miss both. The mgcv lines sit on the edges of the
frmtmb band, which is the agreement of the table above, drawn.

## 6. Growth mixture models

A growth mixture model says that a population contains latent classes
with different trajectories, and that individuals inside a class still
vary. It comes from the structural equation modeling literature (Muthen
and Shedden 1999, *Biometrics* 55), and brms fits it with
`mixture(..., groups = ~id)`.

The `groups` argument is what makes it a growth mixture rather than an
observation-level mixture. Without it, every measurement could belong
to a different class. With it, the class is drawn once per subject, so
all of a subject's measurements come from the same component.

```{r gmm-data}
set.seed(31)
nid <- 80; nt <- 5
cls <- rbinom(nid, 1, 0.4)              # class membership per subject
id <- rep(seq_len(nid), each = nt)
dg <- data.frame(id = factor(id), time = rep(0:(nt - 1), nid))
b0 <- 2 + rnorm(nid, 0, 0.5)            # random intercept inside class
b1 <- c(0.2, 1.2)[cls + 1]              # slow and fast classes
dg$y <- b0[id] + b1[id] * dg$time + rnorm(nid * nt, 0, 0.6)
```

```{r gmm-fit}
fgmm <- frm(bf(y ~ time + (1 | id)) +
              mixture(gaussian(), gaussian(), groups = ~id),
            data = dg)
rbind(class1 = fixef(fgmm)$mu1, class2 = fixef(fgmm)$mu2)
```

Each component keeps its own intercept, slope, residual standard
deviation, and random intercept. The two slopes recover the slow and
fast trajectories.

`mixture_probs()` gives the posterior class probability of each
subject.

```{r gmm-probs}
pr <- mixture_probs(fgmm)
head(round(pr, 3), 3)
assigned <- max.col(pr) - 1L
# label switching is inherent to a mixture, so score both labelings
max(mean(assigned == cls), mean(assigned != cls))
```

Every subject is classified correctly here, because the two
trajectories separate well over five time points. Real growth mixture
models are rarely this clean.

Split the subject trajectories by the class the model gives them, and
draw each component mean over its own panel.

```{r gmm-fig, eval = has_tinyplot, fig.width = 7, fig.height = 3.6, fig.alt = "Two panels side by side, one for each recovered class. Each panel holds the gray trajectories of the subjects assigned to that class, plotted as y against time from 0 to 4. A thick black line is the fitted component mean. The class 1 lines stay near y equal to 2 and 3 across time. The class 2 lines climb from about 2 to about 7."}
dg$class <- factor(assigned[as.integer(dg$id)] + 1L,
                   labels = c("class 1", "class 2"))
tt <- 0:(nt - 1)
b <- rbind(fixef(fgmm)$mu1, fixef(fgmm)$mu2)
comp <- data.frame(
  time = rep(tt, 2),
  yhat = c(b[1, 1] + b[1, 2] * tt, b[2, 1] + b[2, 2] * tt),
  class = factor(rep(c("class 1", "class 2"), each = length(tt))))
tinyplot::tinyplot(y ~ time | id, data = dg, facet = ~class, type = "l",
                   col = "gray60", legend = FALSE, theme = "clean2",
                   xlab = "time", ylab = "y")
tinyplot::plt_add(yhat ~ time, data = comp, facet = ~class, type = "l",
                  col = "black", lwd = 3)
```

The two panels are the model's answer to the question the mixture asks.
Inside a panel the lines are near parallel, which is the random
intercept. Between panels the lines differ in slope, which is the
class.

Two cautions carry over from any maximum likelihood mixture. The
likelihood is multimodal, so compare starting values with
`frm_allfit()`. And a mixture will always find classes, whether or not
they exist, so a model comparison against the single-class fit is the
minimum evidence for reporting them.

```{r gmm-compare}
f1 <- frm(bf(y ~ time + (1 | id)) + gaussian(), data = dg)
c(one_class = AIC(f1), two_class = AIC(fgmm))
```

## 7. Measurement error in a predictor

When a predictor is measured with error, regression on the measured
value is biased toward zero. The bias factor is the reliability ratio,
`var(true) / (var(true) + sd_error^2)`. brms corrects it by giving the
mismeasured variable its own sub-model with a known error standard
deviation, spelled `mi(sdx)` on that response and `mi(x)` in the
predictor of interest. frmtmb uses the same spelling and integrates the
latent true values out with the Laplace approximation.

```{r me-data}
set.seed(3)
n <- 300
z <- rnorm(n)
x_true <- rnorm(n, 0.5 * z, 1)
su <- 0.6                                # known measurement error sd
dme <- data.frame(x = x_true + rnorm(n, 0, su), z = z, su = su)
dme$y <- 1 + 1.0 * x_true + 0.3 * z + rnorm(n, 0, 0.7)
```

```{r me-fit}
fme <- frm(bf(y ~ mi(x) + z) + gaussian() +
             bf(x | mi(su) ~ z) + gaussian(),
           data = dme)
fixef(fme)$y_mu
```

The coefficient on the latent `x` is `mix`. Compare it with the naive
regression on the measured value, and with the attenuation that theory
predicts.

```{r me-check}
naive <- coef(lm(y ~ x + z, data = dme))[["x"]]
c(naive = naive,
  predicted_attenuation = 1.0 * var(x_true) / (var(x_true) + su^2),
  corrected = fixef(fme)$y_mu[["mix"]],
  simulated_truth = 1.0)
stopifnot(fixef(fme)$y_mu[["mix"]] > naive + 0.1)
```

The naive slope sits near the attenuated value that theory predicts.
The corrected slope is close to the truth. The correction is not free:
it widens the standard error, because the latent values are estimated
rather than observed.

```{r me-se}
c(naive_se = summary(lm(y ~ x + z, data = dme))$coefficients["x", 2],
  corrected_se = sqrt(diag(vcov(fme)))[["y_mix"]])
```

The same machinery with `mi()` and no standard deviation imputes
missing values in one step instead of correcting known error.

## 8. Sequential ordinal models with category-specific effects

The proportional odds assumption says one slope governs every category
threshold. `cs()` relaxes it and gives each threshold its own slope,
which is brms's category-specific effect. Under the cumulative
parameterization that model can produce negative probabilities, so
brms and frmtmb both refuse it there and accept it for the sequential
families `sratio`, `cratio`, and `acat`.

```{r cs-data}
set.seed(51)
n <- 500
xo <- rnorm(n)
th <- c(-1.6, 0, 1.6); eff <- c(0.3, 0.6, 0.9)
Y <- integer(n)
for (i in seq_len(n)) {
  cp <- c(plogis(th - eff * xo[i]), 1)
  Y[i] <- sample(1:4, 1, prob = diff(c(0, cp)))
}
dord <- data.frame(y = factor(Y, ordered = TRUE), x = xo)
```

```{r cs-fit}
fcs <- frm(bf(y ~ cs(x)) + sratio(), data = dord)
confint(fcs)
```

`tau_raw` holds the thresholds on an internal increasing scale.
`bcs2_k` is the effect of `x` at threshold `k`. The three effects grow,
which is what the simulation put there.

### Cross-check against a set of binomial regressions

The sequential model has an exact decomposition. It asks, at each
category in turn, whether the process stops there given that it reached
there. Those questions are independent, so the model is equivalent to
`K - 1` separate logistic regressions on shrinking subsets, and the log
likelihoods add up (Laara and Matthews 1985, *Biometrika* 72; Tutz
1991).

```{r cs-check}
glm_fits <- lapply(1:3, function(k) {
  sub <- dord[as.integer(dord$y) >= k, ]
  glm(as.integer(as.integer(sub$y) == k) ~ x, data = sub,
      family = binomial())
})
rbind(frmtmb = confint(fcs)[paste0("bcs2_", 1:3), "est"],
      # the binary question is "stop here", so the sign is reversed
      binomial_glms = -vapply(glm_fits, function(g) coef(g)[["x"]], 0))
c(frmtmb = as.numeric(logLik(fcs)),
  sum_of_glms = sum(vapply(glm_fits, function(g) as.numeric(logLik(g)), 0)))
```

```{r cs-assert}
stopifnot(
  max(abs(confint(fcs)[paste0("bcs2_", 1:3), "est"] +
            vapply(glm_fits, function(g) coef(g)[["x"]], 0))) < 1e-4,
  abs(as.numeric(logLik(fcs)) -
        sum(vapply(glm_fits, function(g) as.numeric(logLik(g)), 0))) < 1e-6
)
```

The coefficients and the log likelihood agree exactly. The frmtmb form
is the more general one, because the decomposition breaks as soon as a
random effect is shared between thresholds, and frmtmb keeps working.

A proportional odds fit is the null model, and `anova()` tests the
restriction.

```{r cs-anova}
anova(frm(bf(y ~ x) + sratio(), data = dord), fcs)
```

## Where the checks live

Every cross-check on this page runs when the page is built, and the
`stopifnot()` calls fail the build if a number moves. The strongest of
them are also pinned in `tests/testthat/test-case-studies.R`, so they
run in the test suite as well.

Two points of friction showed up while writing these case studies, and
each one is handled above:

- A `gr(cov = )` block with one row per level triggers the
  observation-level random effect warning, although the covariance
  matrix identifies the two variances. Use
  `frmtmb_control(check_olre = "ignore")`.
- A multi-trait model with a known covariance can use the long format
  of section 1 or `mvbf()` with the `|ID|` identifier over the same
  `gr(cov = )` term; the two spellings are the same joint density. One
  `|ID|` label must name one grouping specification, and the matrix
  must resolve identically in every formula.

The identity-refit check of section 1 belongs in any model that reads a
matrix from `data2`. It is the one test that a structured covariance
cannot pass by accident.
