---
title: "Feature compatibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Feature compatibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE)
library(frmtmb)
```

```{r helpers}
# Every table below is built from frm_compat(). The page cannot drift
# away from the registry, because it has no separate copy of it.
compat <- frm_compat()

SYM <- c(works = "+", conditional = "~", refused = "x",
         broken = "!", untested = "?")

# Pairs of two kinds, in a fixed row/column orientation.
slice_kinds <- function(row_kind, col_kind) {
  a <- compat[compat$kind_a == row_kind & compat$kind_b == col_kind, ]
  b <- compat[compat$kind_a == col_kind & compat$kind_b == row_kind, ]
  if (nrow(b)) {
    names(b)[1:4] <- names(b)[c(3, 4, 1, 2)]
    b <- b[, names(a)]
  }
  rbind(a, b)
}

grid_table <- function(row_kind, col_kind, rows = NULL, cols = NULL) {
  s <- slice_kinds(row_kind, col_kind)
  rn <- rows %||% unique(s$feature_a)
  cn <- cols %||% unique(s$feature_b)
  m <- matrix("", length(rn), length(cn), dimnames = list(rn, cn))
  m[cbind(match(s$feature_a, rn), match(s$feature_b, cn))] <-
    SYM[s$status]
  knitr::kable(m, align = "c")
}

# Conditions, refusals and defects, grouped by the text that explains
# them, so a rule that covers thirty pairs is written out once.
group_notes <- function(s) {
  s <- s[s$status %in% c("conditional", "refused", "broken") &
           nzchar(s$note), ]
  if (!nrow(s)) return(NULL)
  # a same-kind slice carries each pair in both orientations, to fill
  # both triangles of the grid; list it once here
  uid <- vapply(seq_len(nrow(s)), function(i) {
    paste(sort(c(s$feature_a[i], s$feature_b[i])), collapse = "\r")
  }, "")
  s <- s[!duplicated(uid), ]
  key <- paste(s$status, s$note)
  out <- do.call(rbind, lapply(unique(key), function(k) {
    g <- s[key == k, ]
    pr <- paste0(g$feature_a, " + ", g$feature_b)
    if (length(pr) > 6L) {
      pr <- c(pr[1:5], paste0("and ", length(pr) - 5L, " more"))
    }
    data.frame(Status = SYM[[g$status[1]]],
               Pairs = paste(pr, collapse = "; "),
               Note = g$note[1], stringsAsFactors = FALSE)
  }))
  out[order(match(out$Status, SYM)), ]
}

note_table <- function(row_kind, col_kind) {
  out <- group_notes(slice_kinds(row_kind, col_kind))
  if (is.null(out)) return(invisible(NULL))
  knitr::kable(out, row.names = FALSE)
}

`%||%` <- function(x, y) if (is.null(x)) y else x
fam <- frm_compat_features()
kind_of <- function(k) fam$name[fam$kind == k]
```

This package accepts a large grammar: about thirty families, a dozen
addition terms, twenty covariance structures, six predictor specials,
eight estimation modes, and several model structures. Most of the
combinations are meaningful. Some are refused. A few are accepted and
give a wrong answer.

The worst defects in this package's history were all of the third
kind. `rescor = TRUE` once accepted `cens()` and then dropped the
censoring from the likelihood. `frmtmb_control(profile = TRUE)` once
accepted parameter bounds and applied them to the wrong parameters.
Neither raised an error, so neither looked like a defect.

The lesson is that a guard which does not exist looks exactly like a
guard that passed. The absence of an error is not evidence of support.
This page therefore gives every pair of features one of five states,
and it treats "nobody has checked" as a state of its own.

## The five states

```{r legend}
knitr::kable(data.frame(
  Symbol = unname(SYM),
  Status = names(SYM),
  Meaning = c(
    "Supported and exercised by the test suite.",
    "Supported, but the combination must meet a stated condition.",
    "frm() or the post-fit method stops with an error. The refusal is deliberate.",
    "Accepted, but the result is wrong or the failure does not explain itself. Do not use the pair.",
    "Nothing checks this pair. It may work. Treat a silent success as unverified.")),
  row.names = FALSE)
```

`untested` is the state that matters most. It does not mean the
combination fails. It means that if it fails, it will fail quietly,
because nothing in the test suite would notice.

## Reading the registry from code

The tables on this page are generated from the registry, so you can
ask it the same questions directly.

```{r query, echo = TRUE}
frm_compat("rescor", "cens()")

frm_compat("quadrature", "trunc()")$status
```

Give one feature to get every pair it takes part in, and `status` to
filter:

```{r query2, echo = TRUE}
head(frm_compat("se()", status = "works"), 3)
```

`frm_compat_features()` lists the vocabulary. `frm_compat_rules()`
shows the underlying rules, which are written per kind or per feature
group rather than per pair.

## Combinations to avoid

These pairs are accepted and should not be. Each one is a defect, not
a design decision.

```{r broken}
b <- group_notes(frm_compat(status = "broken"))
knitr::kable(data.frame(Combination = b$Pairs, Evidence = b$Note),
             row.names = FALSE)
```

## Response distributions and addition terms

Addition terms are the most family-sensitive part of the grammar.
`cens()` and `trunc()` need a family that supplies a cumulative
distribution function, and `cens()` additionally refuses discrete
responses. `se()` and `mi()` need a gaussian or student model.

```{r fam-aterm}
grid_table("family", "aterm", rows = kind_of("family"),
           cols = kind_of("aterm"))
```

```{r fam-aterm-notes}
note_table("family", "aterm")
```

## Response distributions and post-fit methods

What you can do with a fit depends on what the family supplies. A
family without a simulator cannot be simulated from. A family without
a mean function has no `fitted()` value.

```{r fam-method}
grid_table("family", "method", rows = kind_of("family"),
           cols = kind_of("method"))
```

```{r fam-method-notes}
note_table("family", "method")
```

## Estimation modes

The modes constrain each other more than they constrain anything else.
`REML`, `quadrature`, and `profile` all move the fixed effects out of
the outer optimization problem, and no two of them can do that at
once. Bounds name outer parameters, so bounds on fixed effects are
refused under `REML` and under `profile`.

`autoscale`, `sparse_x`, and `verbose` are different in kind. They
change the arithmetic or the output, not the model.

```{r mode-mode}
grid_table("mode", "mode", rows = kind_of("mode"), cols = kind_of("mode"))
```

```{r mode-mode-notes}
note_table("mode", "mode")
```

### Modes and addition terms

```{r mode-aterm}
grid_table("aterm", "mode", rows = kind_of("aterm"), cols = kind_of("mode"))
```

```{r mode-aterm-notes}
note_table("aterm", "mode")
```

### Modes and covariance structures

Only `quadrature` reads the covariance structures. It marginalizes one
scalar random intercept at a time, so every block in the model must be
a one-dimensional `us`, `diag`, or `homdiag` term.

```{r mode-cov}
grid_table("covstruct", "mode", rows = kind_of("covstruct"),
           cols = kind_of("mode"))
```

```{r mode-cov-notes}
note_table("covstruct", "mode")
```

## Model structures

```{r struct-aterm}
grid_table("structure", "aterm", rows = kind_of("structure"),
           cols = kind_of("aterm"))
```

```{r struct-aterm-notes}
note_table("structure", "aterm")
```

### Structures and post-fit methods

Several post-fit methods are univariate-only. A multivariate fit
predicts one response at a time, through `predict(fit, resp = )`, and
refuses `fitted()`, `simulate()` and `residuals()`. The inference
surface is not univariate-only: `confint()`, `hypothesis()` and
`frm_sample()` read the outer parameter vector, which a multivariate
fit has like any other.

```{r struct-method}
grid_table("structure", "method", rows = kind_of("structure"),
           cols = kind_of("method"))
```

```{r struct-method-notes}
note_table("structure", "method")
```

## Predictor specials

```{r special-mode}
grid_table("special", "mode", rows = kind_of("special"),
           cols = kind_of("mode"))
```

```{r special-notes}
note_table("special", "mode")
```

The specials `mi()`, `gp()`, and `cs()` share a name with a covariance
structure. The registry writes the predictor forms as `mi_pred()`,
`gp_pred()`, and `cs_pred()` to keep them apart.

## Formula grammar

Three spellings have restrictions of their own.

```{r grammar}
spelling <- c(bar_crossing = "x * (1 | g)",
              call_group = "(1 | factor(x))",
              double_bar = "(x || g)")
# Each grammar feature carries one rule that covers all its pairs.
# Report the status and note that rule gives to most of them.
gr <- do.call(rbind, lapply(names(spelling), function(feat) {
  s <- frm_compat(feat)
  s <- s[s$kind_b != "grammar", ]
  k <- paste(s$status, s$note)
  top <- s[k == names(sort(table(k), decreasing = TRUE))[1], ][1, ]
  data.frame(Spelling = spelling[[feat]], Status = SYM[[top$status]],
             Note = top$note, stringsAsFactors = FALSE)
}))
knitr::kable(gr, row.names = FALSE)
```

## Coverage

```{r coverage}
tb <- table(factor(compat$status, levels = names(SYM)))
knitr::kable(data.frame(
  Status = names(tb), Pairs = as.integer(tb),
  Share = paste0(round(100 * as.integer(tb) / sum(tb)), "%")),
  row.names = FALSE)
```

The untested share is the honest measure of what this registry does
not yet know. It shrinks as pairs are tested, not as the code is
trusted. To close one, add a test and change the rule in `R/compat.R`.
