---
title: "Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_dharma <- requireNamespace("DHARMa", quietly = TRUE)
```

Diagnostics for a frmtmb fit answer three separate questions: did the
optimizer find a real optimum, does the fitted family describe the
data, and are the reported uncertainties trustworthy. Each has its own
tool.

```{r}
library(frmtmb)
set.seed(1)
n <- 400
dd <- data.frame(x = rnorm(n), g = factor(rep(1:20, n / 20)))
dd$y <- rpois(n, exp(0.5 + 0.4 * dd$x + rnorm(20, 0, 0.5)[dd$g]))
fit <- frm(bf(y ~ x + (1 | g)) + poisson(), data = dd)
```

## Did the optimizer converge?

`diagnose()` reports the convergence code, the largest gradient
component at the optimum, whether the Hessian is positive definite,
and covariance parameters near a boundary (a variance component
collapsing to zero is the common case):

```{r}
diagnose(fit)
```

## Convergence problems

The warning "Large maximum absolute gradient at the optimum" means
the optimizer stopped at a point where the objective still has slope
above `frmtmb_control(grad_tol = 1e-3)`. Work through these in
order; each step tells you something even when it does not fix the
warning.

1. **Identify the parameter.** `diagnose(fit)` prints the largest
   gradient component and its name. A coefficient of a continuous
   predictor points to scaling; a `theta_*` component points to a
   variance-parameter problem, and `diagnose()` flags extreme values
   (a log-SD far below zero is a variance collapsing to its boundary,
   which is a model issue, not an optimizer issue).
2. **Fix predictor scaling.** For a flagged coefficient,
   `frmtmb_control(autoscale = TRUE)` standardizes badly scaled
   columns internally and judges convergence in the rescaled units;
   rescaling the variable in the data works equally well.
3. **Restart harder.** `frmtmb_control(restarts = 3)` re-launches the
   optimizer from the current optimum until the gradient passes;
   `optCtrl = list(iter.max = 5000, eval.max = 5000)` lifts the
   iteration caps for slow problems.
4. **Compare optimizers.** `frm_allfit(fit)` refits with nlminb,
   optim, bobyqa, and nloptr and tabulates log-likelihoods. All
   optimizers agreeing on the log-likelihood to several decimals
   while one warns usually means a flat ridge near the optimum, not a
   wrong answer; disagreement means the warning is real.
5. **Look at the profile.** `plot(profile(fit, "theta_1"))` shows
   whether the flagged parameter sits in a flat valley (benign) or on
   a slope (not converged).
6. **Boundary fits.** When a random-effect SD is collapsing to zero,
   the likelihood surface is one-sided and the gradient check can
   never fully settle. Either simplify the random-effect structure,
   or keep the term and regularize with
   `prior = set_prior("exponential(1)", class = "sd")` (a MAP fit;
   the same move as `blme`'s covariance priors).
7. **Judge marginal cases.** A gradient just above the default
   `1e-3` (say 0.001-0.01) with stable estimates across restarts and
   optimizers is typically flatness at machine precision for that
   problem's scale, not non-convergence. The threshold is a
   heuristic; `frmtmb_control(grad_tol = )` tunes it, and steps 4-5
   are the evidence for deciding.

## Does the family fit the data?

Three residual types, in increasing order of statistical care:

- `residuals(fit)` is response-scale (`y - E[y]`), in the response's
  units. Good for effect-size intuition, useless as a distributional
  check for counts.
- `residuals(fit, type = "pearson")` standardizes by the family
  variance.
- `residuals(fit, type = "osa")` gives one-step-ahead quantile
  residuals via [TMB::oneStepPredict()]: each observation's CDF
  position given the previous ones, mapped through the normal quantile
  function. They are standard normal under a correctly specified model
  regardless of family, and remain valid under correlated
  observations, where pearson residuals mislead.

```{r osa-qq, fig.width = 6, fig.height = 4.5, fig.alt = "Normal quantile-quantile plot of the one-step-ahead residuals. The points follow the identity line over most of the range, with small departures in the tails."}
r <- residuals(fit, type = "osa")
qqnorm(r); abline(0, 1)
```

Simulation-based residuals through DHARMa cover the same ground with a
richer test suite (dispersion, zero-inflation, outliers) and better
plots:

```{r dharma-plot, eval = has_dharma, fig.width = 7, fig.height = 4, fig.alt = "The standard DHARMa panel pair. The left panel is a uniform quantile-quantile plot of the scaled residuals with its three goodness-of-fit tests. The right panel plots the scaled residuals against the rank-transformed model predictions."}
dh <- dharma_residuals(fit, nsim = 200, seed = 1)
plot(dh)
DHARMa::testDispersion(dh, plot = FALSE)
```

A misspecified model shows up immediately; here the same data fit
without the group effect is overdispersed:

```{r, eval = has_dharma}
fit_bad <- frm(bf(y ~ x) + poisson(), data = dd)
DHARMa::testDispersion(dharma_residuals(fit_bad, nsim = 200, seed = 1),
                       plot = FALSE)
```

Two quicker looks. `plot(fit)` draws Pearson residuals against fitted
values and their normal QQ plot. And `pp_check(fit)` overlays the
observed response with responses simulated from the fitted model,
through bayesplot's `ppc_*` functions (brms users know this display):

```{r pp-check, eval = requireNamespace("bayesplot", quietly = TRUE), fig.width = 6, fig.height = 4, fig.alt = "Posterior predictive check. A thick dark density curve is the observed count response. Twenty thin light curves are responses simulated from the fitted model. The simulated curves surround the observed one."}
bayesplot::pp_check(fit, ndraws = 20)
```

## Are the standard errors trustworthy?

Wald standard errors and the Laplace approximation share failure
modes: variance components with few groups, binary data with tiny
clusters. Four things in this package route around that:
`confint(method = "profile")` replaces the quadratic assumption,
`frm_bootstrap()` replaces it with a resampling distribution,
`frm(quadrature = TRUE)` replaces the Laplace approximation itself
with adaptive quadrature, and `frm(importance = )` reweights it
(the next section).

To measure the violation rather than route around it, install the
companion package **frmtmb.sample** and read
`vignette("posterior-diagnostics", package = "frmtmb.sample")`, which
covers `check_laplace()` and the diagnostics of a sampled fit; this
page stays with the maximum-likelihood side.

## When the Laplace approximation is the problem

The tools above ask whether the *reported uncertainty* is trustworthy.
This section asks whether the *estimate* is, because the Laplace
approximation has a regime where it is biased, and no interval method
repairs a biased point estimate.

The regime is the package's named audience: a binary response in small
clusters. The Laplace approximation replaces each group's integral by a
Gaussian centered on the conditional mode. For a Bernoulli response
with a handful of rows per group, the integrand is skewed, the Gaussian
does not match it, and the variance component comes out too small.

`frm(importance =)` corrects that. It keeps the Laplace Gaussian, but
uses it as a *proposal*: it draws from it, reweights the draws by how
much the true integrand differs from it, and averages. The result is an
unbiased estimate of the exact marginal likelihood, at every parameter
value, not an approximation that is accurate only near the optimum.
The method is Skaug and Fournier's (2006), and it is what ADMB's `-is`
option does.

```{r imp-data}
set.seed(7)
ng <- 40
per <- 5
gg <- rep(seq_len(ng), each = per)
xx <- rnorm(ng * per)
bb <- cbind(rnorm(ng, 0, 1.5), rnorm(ng, 0, 1))
bd <- data.frame(
  y = rbinom(ng * per, 1,
             plogis(-0.5 + 0.8 * xx + bb[gg, 1] + bb[gg, 2] * xx)),
  x = xx, g = factor(gg)
)
```

Fit it both ways. The correction takes a draw count, and 0 (the
default) turns it off:

```{r imp-fit}
lap <- frm(bf(y ~ x + (x | g)) + binomial(), data = bd)
imp <- frm(bf(y ~ x + (x | g)) + binomial(), data = bd,
           importance = 500)
c(laplace = as.numeric(logLik(lap)),
  importance = as.numeric(logLik(imp)))
```

The corrected log-likelihood is the honest one. `print()` says so, and
says how far it can be trusted:

```{r imp-print}
imp
```

Note what this model is. The random effect is a correlated intercept
and slope, so the per-group integral is two-dimensional.
`quadrature = TRUE` cannot marginalize it at all: the Gauss-Kronrod
transform underneath it integrates one scalar random effect at a time.
The importance correction takes any block dimension, which is the main
reason to reach for it.

### Reading the effective sample size

An importance estimate is only as good as the proposal's coverage of
the integrand. The effective sample size measures that coverage, one
number per group, as a fraction of the draws:

```{r imp-ess}
summary(imp$importance$ess)
imp$importance$mcse
```

Read it like this.

* Near 1 means the weights are nearly equal, so the proposal matches
  the integrand and almost every draw is doing work. For a gaussian
  response it is exactly 1, because the Laplace approximation is
  already exact and there is nothing to reweight.
* Below `0.25` the fit warns and names the groups. That threshold is
  placed by measurement, not by taste. Ordinary designs sit far above
  it: 0.93 on the package's probe design, 0.97 on a scalar random
  intercept, exactly 1.00 for a gaussian response, 0.64 here. A
  proposal that has genuinely stopped working sits far below it: half a
  log standard deviation of displacement drops the worst group to
  0.009. In between, the threshold is meant to fire, and does. Forty
  groups of three binary rows with a standard deviation of two report
  anywhere from 0.08 to 0.62 depending on the sample, and the warning
  is right to fire on the low half.
* `mcse` is the Monte Carlo standard error of the corrected
  log-likelihood, derived from those effective sample sizes. It is the
  right scale for judging whether two corrected fits really differ.
  Quadruple the draws to halve it.

If the warning fires, raise `importance`. The estimate stays unbiased
either way; what degrades is precision, and the effective sample size
is what tells you so.

### What it costs

The correction tapes `N` stacked copies of the likelihood, so time and
memory grow linearly in the draw count. On a 480-row, two-dimensional
block with 60 groups, the gradient costs about 0.28 ms per draw against
2.5 ms for the Laplace gradient, and the tape holds roughly 320 bytes
per row and draw. A thousand draws on that model is a few seconds and
about 180 MB. Start at a few hundred draws, read the effective sample
sizes, and raise the count only if they ask for it.

The fit proceeds in rounds. Each one freezes a proposal at the current
estimate, optimizes, and refreezes; the loop stops when the estimates
settle. `frmtmb_control()` holds the seed, the round cap and the
effective-sample-size threshold. The seed makes the fit reproducible to
the last bit, and the draws come from a private random stream, so a
fit neither reads nor disturbs the session's random state.

### Scope

The first version corrects exactly one random-effect block over a
grouping factor, of any dimension, for a single response whose family
has a rowwise density. `trunc()` and `cens()` are supported, which is a
real difference from `quadrature = TRUE`: that method has to refuse
truncation because its far quadrature nodes underflow the truncation
normalizer, while importance draws stay near the conditional mode.
Truncation does not degrade the correction either. Walking a bound up
past the response median, discarding 64 percent of the rows, still
leaves the worst group holding 0.60 of its draws.
Everything outside the scope is refused by name, with the remedy in the
message. `frm_compat("importance")` lists the rules.

## Variance components on their natural scale

`confint_varcorr()` reports random-effect SDs and correlations with
delta-method intervals on interpretable scales:

```{r}
confint_varcorr(fit)
```
