---
title: "Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_dharma <- requireNamespace("DHARMa", quietly = TRUE)
```

Diagnostics for a frmtmb fit answer three separate questions: did the
optimizer find a real optimum, does the fitted family describe the
data, and are the reported uncertainties trustworthy. Each has its own
tool.

```{r}
library(frmtmb)
set.seed(1)
n <- 400
dd <- data.frame(x = rnorm(n), g = factor(rep(1:20, n / 20)))
dd$y <- rpois(n, exp(0.5 + 0.4 * dd$x + rnorm(20, 0, 0.5)[dd$g]))
fit <- frm(bf(y ~ x + (1 | g)) + poisson(), data = dd)
```

## Did the optimizer converge?

`diagnose()` reports the convergence code, the largest gradient
component at the optimum, whether the Hessian is positive definite,
and covariance parameters near a boundary (a variance component
collapsing to zero is the common case):

```{r}
diagnose(fit)
```

## Does the family fit the data?

Three residual types, in increasing order of statistical care:

- `residuals(fit)` is response-scale (`y - E[y]`), in the response's
  units. Good for effect-size intuition, useless as a distributional
  check for counts.
- `residuals(fit, type = "pearson")` standardizes by the family
  variance.
- `residuals(fit, type = "osa")` gives one-step-ahead quantile
  residuals via [TMB::oneStepPredict()]: each observation's CDF
  position given the previous ones, mapped through the normal quantile
  function. They are standard normal under a correctly specified model
  regardless of family, and remain valid under correlated
  observations, where pearson residuals mislead.

```{r}
r <- residuals(fit, type = "osa")
qqnorm(r); abline(0, 1)
```

Simulation-based residuals through DHARMa cover the same ground with a
richer test suite (dispersion, zero-inflation, outliers) and better
plots:

```{r, eval = has_dharma}
dh <- dharma_residuals(fit, nsim = 200, seed = 1)
plot(dh)
DHARMa::testDispersion(dh, plot = FALSE)
```

A misspecified model shows up immediately; here the same data fit
without the group effect is overdispersed:

```{r, eval = has_dharma}
fit_bad <- frm(bf(y ~ x) + poisson(), data = dd)
DHARMa::testDispersion(dharma_residuals(fit_bad, nsim = 200, seed = 1),
                       plot = FALSE)
```

## Are the standard errors trustworthy?

Wald standard errors and the Laplace approximation share failure
modes: variance components with few groups, binary data with tiny
clusters. `check_laplace()` samples the fitted objective with NUTS
(initialized at the ML mode) and compares posterior means and SDs
against the ML estimates and Wald SEs; large `z_shift` or `sd_ratio`
far from 1 flags parameters where Wald intervals should not be
trusted - use `confint(method = "profile")` or the posterior itself
for those.

```{r, eval = FALSE}
check_laplace(fit, chains = 2, iter = 1000)
```

The drawn posterior is also a diagnostic object in its own right;
bayesplot works on the draws:

```{r, eval = FALSE}
ds <- frm_sample(fit, chains = 4,
                 priors = set_prior("normal(0, 5)", class = "b") +
                   set_prior("exponential(1)", class = "sd"))
bayesplot::mcmc_intervals(as.matrix(ds),
                          pars = c("(Intercept)", "x"))
```

Without priors, `frm_sample()` samples with flat improper priors on
the internal parameters; fine as a diagnostic, fragile as inference.
Supply priors (brms-style `set_prior()`) for anything more.

## Variance components on their natural scale

`confint_varcorr()` reports random-effect SDs and correlations with
delta-method intervals on interpretable scales:

```{r}
confint_varcorr(fit)
```
