---
title: "Model diagnostics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_dharma <- requireNamespace("DHARMa", quietly = TRUE)
```

Diagnostics for a frmtmb fit answer three separate questions: did the
optimizer find a real optimum, does the fitted family describe the
data, and are the reported uncertainties trustworthy. Each has its own
tool.

```{r}
library(frmtmb)
set.seed(1)
n <- 400
dd <- data.frame(x = rnorm(n), g = factor(rep(1:20, n / 20)))
dd$y <- rpois(n, exp(0.5 + 0.4 * dd$x + rnorm(20, 0, 0.5)[dd$g]))
fit <- frm(bf(y ~ x + (1 | g)) + poisson(), data = dd)
```

## Did the optimizer converge?

`diagnose()` reports the convergence code, the largest gradient
component at the optimum, whether the Hessian is positive definite,
and covariance parameters near a boundary (a variance component
collapsing to zero is the common case):

```{r}
diagnose(fit)
```

## Convergence problems

The warning "Large maximum absolute gradient at the optimum" means
two things at once, and the second one is the finding. The optimizer
stopped where the objective still has slope above
`frmtmb_control(grad_tol = 1e-3)`, AND one exact Newton step from
there would still gain more than `grad_tol` in log-likelihood. The
warning gives that second number, so it says how much the fit is
short by rather than only that a derivative is large.

A gradient alone cannot say this. The optimizer stops on a RELATIVE
change in an objective that grows with the sample size, so the
gradient at a perfectly good optimum grows with it too: the same
gaussian design warned on 0 of 20 replicates at 200 rows and 20 of 20
at 20,000. A parameter sitting on a bound from `set_prior(ub = )`
also has a nonzero gradient by construction, because the constrained
optimum is the bound. Neither is counted now.

The check has one blind spot, and it is worth knowing because the
remedy is a setting rather than a refit. The gradient is what admits a
fit to the second reading, so a fit whose gradient never reaches
`grad_tol` is never looked at. A predictor column scaled far below one
does that: it holds the gradient down while the coefficient stalls.
With `autoscale = FALSE` and a column spread of `1e-7` a gaussian fit
reports convergence at a gradient of `1.6e-5` and a log-likelihood 297
units below the standardized fit of the same data, silently. The
default `autoscale` rule is what closes it, and `diagnose()` names
badly scaled columns whether or not the gradient noticed.

Work through these in order; each step tells you something even when
it does not fix the warning.

1. **Identify the parameter.** `diagnose(fit)` prints the largest
   gradient component and its name, the parameters a bound holds in
   place, the largest gradient over the rest WITH its own parameter,
   and the log-likelihood still on the table. On a fit with an active
   bound those are two different parameters, and the one beside the
   projected gradient is the one to work on: the other is pinned and
   cannot move. A coefficient of a continuous
   predictor points to scaling; a `theta_*` component points to a
   variance-parameter problem, and `diagnose()` flags extreme values
   (a log-SD far below zero is a variance collapsing to its boundary,
   which is a model issue, not an optimizer issue).
2. **Fix predictor scaling.** For a flagged coefficient,
   `frmtmb_control(autoscale = TRUE)` standardizes badly scaled
   columns internally and judges convergence in the rescaled units;
   rescaling the variable in the data works equally well.
3. **Restart harder.** `frmtmb_control(restarts = 3)` re-launches the
   optimizer from the current optimum until the gradient passes;
   `optCtrl = list(iter.max = 5000, eval.max = 5000)` lifts the
   iteration caps for slow problems.
4. **Compare optimizers.** `frm_allfit(fit)` refits with nlminb,
   optim, bobyqa, and nloptr and tabulates log-likelihoods. All
   optimizers agreeing on the log-likelihood to several decimals
   while one warns usually means a flat ridge near the optimum, not a
   wrong answer; disagreement means the warning is real.
5. **Look at the profile.** `plot(profile(fit, "theta_1"))` shows
   whether the flagged parameter sits in a flat valley (benign) or on
   a slope (not converged).
6. **Boundary fits.** When a random-effect SD is collapsing to zero,
   the likelihood surface is one-sided and the gradient check can
   never fully settle. Either simplify the random-effect structure,
   or keep the term and regularize with
   `prior = set_prior("exponential(1)", class = "sd")` (a MAP fit;
   the same move as `blme`'s covariance priors).
7. **Judge marginal cases by the log-likelihood, not the gradient.**
   Read the second number in the warning, which `diagnose()` also
   prints. It is what a Newton step would gain, and it is measured on
   the scale answers live on: below `1e-3` nothing you report can
   move, and a warning at `1e-3` to `1e-2` is worth a look but rarely
   worth a refit. A large one is real, whatever the gradient reads: a
   near-collinear design can warn with a gradient of `5e-3` and half
   a log-likelihood unit left, which no restart recovers.
   `frmtmb_control(grad_tol = )` tunes both readings together, and
   steps 4-5 are the evidence for deciding.

## Does the family fit the data?

Three residual types, in increasing order of statistical care:

- `residuals(fit)` is response-scale (`y - E[y]`), in the response's
  units. Good for effect-size intuition, useless as a distributional
  check for counts.
- `residuals(fit, type = "pearson")` standardizes by the family
  variance.
- `residuals(fit, type = "osa")` gives one-step-ahead quantile
  residuals via [TMB::oneStepPredict()]: each observation's CDF
  position given the previous ones, mapped through the normal quantile
  function. They are standard normal under a correctly specified model
  regardless of family, and remain valid under correlated
  observations, where pearson residuals mislead.

```{r osa-qq, fig.width = 6, fig.height = 4.5, fig.alt = "Normal quantile-quantile plot of the one-step-ahead residuals. The points follow the identity line over most of the range, with small departures in the tails."}
r <- residuals(fit, type = "osa")[, "Estimate"]
qqnorm(r); abline(0, 1)
```

Simulation-based residuals through DHARMa cover the same ground with a
richer test suite (dispersion, zero-inflation, outliers) and better
plots:

```{r dharma-plot, eval = has_dharma, fig.width = 7, fig.height = 4, fig.alt = "The standard DHARMa panel pair. The left panel is a uniform quantile-quantile plot of the scaled residuals with its three goodness-of-fit tests. The right panel plots the scaled residuals against the rank-transformed model predictions."}
dh <- dharma_residuals(fit, nsim = 200, seed = 1)
plot(dh)
DHARMa::testDispersion(dh, plot = FALSE)
```

A misspecified model shows up immediately; here the same data fit
without the group effect is overdispersed:

```{r, eval = has_dharma}
fit_bad <- frm(bf(y ~ x) + poisson(), data = dd)
DHARMa::testDispersion(dharma_residuals(fit_bad, nsim = 200, seed = 1),
                       plot = FALSE)
```

Two quicker looks. `plot(fit)` draws Pearson residuals against fitted
values and their normal QQ plot. And `pp_check(fit)` overlays the
observed response with responses simulated from the fitted model,
through bayesplot's `ppc_*` functions (brms users know this display):

```{r pp-check, eval = requireNamespace("bayesplot", quietly = TRUE), fig.width = 6, fig.height = 4, fig.alt = "Posterior predictive check. A thick dark density curve is the observed count response. Twenty thin light curves are responses simulated from the fitted model. The simulated curves surround the observed one."}
bayesplot::pp_check(fit, ndraws = 20)
```

## Are the standard errors trustworthy?

Wald standard errors and the Laplace approximation share failure
modes: variance components with few groups, binary data with tiny
clusters. Four things in this package route around that:
`confint(method = "profile")` replaces the quadratic assumption,
`frm_bootstrap()` replaces it with a resampling distribution,
`frm(quadrature = TRUE)` replaces the Laplace approximation itself
with adaptive quadrature, and `frm(importance = )` reweights it
(the next section).

### NaN standard errors

The covariance is the inverse of the Hessian at the optimum. A plain
inverse fails as a whole: one unusable direction makes every entry
`NaN`. `frm()` does not stop there. It builds the Hessian when it
fits, and when the plain inverse fails it tries two repairs before it
gives up on a parameter:

1. It scales the Hessian to unit diagonal and inverts that. This is
   the same matrix, so it changes nothing where the plain inverse
   worked. It recovers the standard errors where one parameter is on a
   very different scale from the others. On a model without random
   effects it also uses the exact Hessian instead of a
   finite-difference one, which matters for a coefficient of a
   covariate in large units.
2. When that also fails, it removes the directions the likelihood does
   not determine and keeps the standard errors of the other
   parameters. A parameter goes with a direction when a measurable part
   of it lies along that direction, so a direction spread over many
   coefficients takes all of them.

The parameters it removes have no standard error. `frm()` names them
in one warning, "Standard errors are not available for k of n
parameters", with a reason for each:

* **A bound holds it.** The fit is a constrained optimum. The other
  standard errors are conditional on the bound.
* **The data separate the outcomes.** A coefficient of a binomial or
  ordinal model runs off toward infinity, and the optimizer stops
  where its tolerances do.
* **The likelihood is flat along it.** A standard deviation at 0 does
  this. So does a `mo()` simplex weight at 0, because the simplex
  coordinate has gone to minus infinity. Two parameters that enter
  only through their sum do it too, and so does a nonlinear term
  evaluated outside its own support; see the nonlinear section of
  `vignette("frmtmb")`.
* **A step along it raises the log-likelihood.** The estimates are not
  a maximum in that direction. Refit from other starting values. A
  short step along the direction confirms this before the warning
  says it; a direction where the step does not gain is called flat.
* **Its Hessian row is not finite.**

An earlier warning that names a parameter, such as the one for a
grouping factor with one level, explains that parameter only. The
other parameters still appear in this warning. A setting of `"ignore"`
on the earlier check does not explain anything.

`summary()` repeats the list under the coefficient tables, and
`confint()` and `vcov()` show `NaN` for those parameters only. A
prediction whose standard error depends on a removed direction gets
`NaN` for it, and so does an emmean or a `hypothesis()`. `fitted()`,
`predict()`, `frm_linpred()`, `conditional_effects()`, `emmeans()` and
`hypothesis()` warn once per call with the number of such rows. A
prediction that does not move along a removed
direction keeps its standard error. A fit that did not converge gets
the convergence warning instead of this one, and keeps sdreport()'s
own covariance.

The check costs time. On a model without random effects it reads the
exact Hessian, at about one gradient per coefficient: 8 to 23 percent
of the fit on small models and 19 to 24 percent on a model with 301
coefficients. With random effects it builds the finite-difference
Hessian, two gradients per outer parameter, when the model has at
most 10 outer parameters or those gradients are at most a quarter of
the evaluations the optimizer made. Otherwise the check waits for the
first use of a standard error (`summary()`, `fixef()`, `vcov()`,
`confint()`, a prediction's standard error) and warns there, so a
large mixed model can fit without the warning and give it later. The
rule counts work and does not read the clock, so a call warns in the
same place on every run. To
skip the check, use `frmtmb_control(check_se = "ignore")`.

`diagnose(fit)` also reports "Flat directions" when the likelihood
does not depend on a parameter at all: zero gradient and an empty
Hessian row, measured by perturbing the parameter and seeing whether
the gradient moves.

To measure the violation rather than route around it, install the
companion package **frmtmb.sample** and read
`vignette("posterior-diagnostics", package = "frmtmb.sample")`, which
covers `check_laplace()` and the diagnostics of a sampled fit; this
page stays with the maximum-likelihood side.

### A distributional parameter with no maximum

A standard error in the thousands on an estimate in the tens is not a
wide interval. It means the likelihood was still rising where the
optimizer stopped, so the number reported is the stopping point.

`student()`'s `nu` is the case you will meet. A student-t becomes a
gaussian only as `nu` goes to infinity, so on data with no heavy tails
the likelihood has no maximum in `nu` and the fit reports whatever the
last step reached. Values above `1e9` are ordinary, and two runs of the
same model on the same data in a different row order can differ by
orders of magnitude in `nu` while agreeing on every coefficient.

```{r}
set.seed(41)
du <- data.frame(x = rnorm(120))
# uniform errors: lighter tails than any student-t
du$y <- 1 + 0.5 * du$x + (runif(120) - 0.5) * 3.4
diagnose(frm(bf(y ~ x) + student(), data = du), quiet = TRUE)$unbounded_dpar
```

`diagnose()` reports this as "Distributional parameter at the end of
its link" and prints the natural-scale value beside the link-scale
estimate. The pair is the evidence, not either half: a parameter
legitimately far out on its link keeps a small standard error.

Read the fit as the gaussian one it has become. To get a number you can
report, refit with `gaussian()`, or hold `nu` somewhere finite with a
prior; see `set_prior()`.

The same parameter runs off the other end, and there the reading is
opposite. On tails heavier than any identified `nu` can hold, `nu` goes
down to one: `log(nu - 1)` around -20, a standard error in the
thousands, and a natural-scale value that prints as `1`.

```{r}
set.seed(77101)
dc <- data.frame(x = rnorm(200))
dc$y <- 1 + 0.5 * dc$x + rcauchy(200)      # the heaviest tails there are
diagnose(frm(bf(y ~ x) + student(), data = dc), quiet = TRUE)$unbounded_dpar
```

`diagnose()` names this under the same heading, so read the sign. A
large POSITIVE estimate means no heavy tails and `gaussian()` says so.
A large NEGATIVE one means the opposite, and `gaussian()` is then the
worst fit available: the data is the message, and a prior is the way to
hold `nu` finite.

## When the Laplace approximation is the problem

The tools above ask whether the *reported uncertainty* is trustworthy.
This section asks whether the *estimate* is, because the Laplace
approximation has a regime where it is biased, and no interval method
repairs a biased point estimate.

The regime is the package's named audience: a binary response in small
clusters. The Laplace approximation replaces each group's integral by a
Gaussian centered on the conditional mode. For a Bernoulli response
with a handful of rows per group, the integrand is skewed, the Gaussian
does not match it, and the variance component comes out too small.

`frm(importance =)` corrects that. It keeps the Laplace Gaussian, but
uses it as a *proposal*: it draws from it, reweights the draws by how
much the true integrand differs from it, and averages. The result is an
unbiased estimate of the exact marginal likelihood, at every parameter
value, not an approximation that is accurate only near the optimum.
The method is Skaug and Fournier's (2006), and it is what ADMB's `-is`
option does.

```{r imp-data}
set.seed(7)
ng <- 40
per <- 5
gg <- rep(seq_len(ng), each = per)
xx <- rnorm(ng * per)
bb <- cbind(rnorm(ng, 0, 1.5), rnorm(ng, 0, 1))
bd <- data.frame(
  y = rbinom(ng * per, 1,
             plogis(-0.5 + 0.8 * xx + bb[gg, 1] + bb[gg, 2] * xx)),
  x = xx, g = factor(gg)
)
```

Fit it both ways. The correction takes a draw count, and 0 (the
default) turns it off:

```{r imp-fit}
lap <- frm(bf(y ~ x + (x | g)) + bernoulli(), data = bd)
imp <- frm(bf(y ~ x + (x | g)) + bernoulli(), data = bd,
           importance = 500)
c(laplace = as.numeric(logLik(lap)),
  importance = as.numeric(logLik(imp)))
```

The corrected log-likelihood is the honest one. `print()` says so, and
says how far it can be trusted:

```{r imp-print}
imp
```

Note what this model is. The random effect is a correlated intercept
and slope, so the per-group integral is two-dimensional.
`quadrature = TRUE` cannot marginalize it at all: the Gauss-Kronrod
transform underneath it integrates one scalar random effect at a time.
The importance correction takes any block dimension, which is the main
reason to reach for it.

### Reading the effective sample size

An importance estimate is only as good as the proposal's coverage of
the integrand. The effective sample size measures that coverage, one
number per group, as a fraction of the draws:

```{r imp-ess}
summary(imp$importance$ess)
imp$importance$mcse
```

Read it like this.

* Near 1 means the weights are nearly equal, so the proposal matches
  the integrand and almost every draw is doing work. For a gaussian
  response it is exactly 1, because the Laplace approximation is
  already exact and there is nothing to reweight.
* Below `0.25` the fit warns and names the groups. That threshold is
  placed by measurement, not by taste. Ordinary designs sit far above
  it: 0.93 on the package's probe design, 0.97 on a scalar random
  intercept, exactly 1.00 for a gaussian response, 0.64 here. A
  proposal that has genuinely stopped working sits far below it: half a
  log standard deviation of displacement drops the worst group to
  0.009. In between, the threshold is meant to fire, and does. Forty
  groups of three binary rows with a standard deviation of two report
  anywhere from 0.08 to 0.62 depending on the sample, and the warning
  is right to fire on the low half.
* `mcse` is the Monte Carlo standard error of the corrected
  log-likelihood, derived from those effective sample sizes. It is the
  right scale for judging whether two corrected fits really differ.
  Quadruple the draws to halve it.

If the warning fires, raise `importance`. The estimate stays unbiased
either way; what degrades is precision, and the effective sample size
is what tells you so.

### What it costs

The correction tapes `N` stacked copies of the likelihood, so time and
memory grow linearly in the draw count. On a 480-row, two-dimensional
block with 60 groups, the gradient costs about 0.28 ms per draw against
2.5 ms for the Laplace gradient, and the tape holds roughly 320 bytes
per row and draw. A thousand draws on that model is a few seconds and
about 180 MB. Start at a few hundred draws, read the effective sample
sizes, and raise the count only if they ask for it.

The fit proceeds in rounds. Each one freezes a proposal at the current
estimate, optimizes, and refreezes; the loop stops when the estimates
settle. `frmtmb_control()` holds the seed, the round cap and the
effective-sample-size threshold. The seed makes the fit reproducible to
the last bit, and the draws come from a private random stream, so a
fit neither reads nor disturbs the session's random state.

### Scope

The correction takes any number of random-effect blocks, of any
dimension, if they share one grouping factor with one set of levels.
That is what distributional regression writes: `(1 | g)` in `mu` and
`(1 | g)` in `sigma` are two blocks over `g`, and a level's
coefficients from both are drawn together from one proposal. Blocks
over *different* factors, crossed or nested, are refused by name,
because the marginal likelihood is then one integral over every factor
at once and does not split into the per-group integrals the correction
resamples. The response must be a single one whose family
has a rowwise density. `trunc()` and `cens()` are supported, which is a
real difference from `quadrature = TRUE`: that method has to refuse
truncation because its far quadrature nodes underflow the truncation
normalizer, while importance draws stay near the conditional mode.
Truncation does not degrade the correction either. Walking a bound up
past the response median, discarding 64 percent of the rows, still
leaves the worst group holding 0.60 of its draws.
Everything outside the scope is refused by name, with the remedy in the
message. `frm_compat("importance")` lists the rules.

## Variance components on their natural scale

`confint_varcorr()` reports random-effect SDs and correlations with
delta-method intervals on interpretable scales:

```{r}
confint_varcorr(fit)
```
