---
title: "Getting started with frmtmb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with frmtmb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
```

frmtmb fits regression models written in the brms formula grammar by
maximum likelihood, with the Laplace approximation for random effects.
The backend is RTMB: the model formula compiles to an R objective
function that is differentiated on the TMB AD tape. There is no MCMC and
no compilation at run time.

## A mixed model

```{r, eval = has_lme4}
library(frmtmb)
data(sleepstudy, package = "lme4")

fit <- frm(bf(Reaction ~ Days + (Days | Subject)) + gaussian(),
           data = sleepstudy)
summary(fit)
```

The usual methods work: `fixef()`, `ranef()`, `VarCorr()`, `predict()`
(with `newdata`, `se.fit`, and `re.form`), `confint()` (Wald, profile,
or likelihood-root), `simulate()`, `anova()` for likelihood-ratio
tests, and `emmeans` for marginal means.

## Distributional regression

Every distributional parameter can get its own formula with the full
predictor grammar, including random effects and smooths:

```{r, eval = has_lme4}
fit2 <- frm(bf(Reaction ~ Days + (Days | Subject),
               sigma ~ Days) + gaussian(),
            data = sleepstudy)
fixef(fit2)$sigma
```

## Nonlinear formulas

With `nl = TRUE` the model formula is an arbitrary R expression over
named parameters, each of which has its own linear predictor:

```{r}
set.seed(1)
d <- data.frame(x = runif(200, 0, 5))
d$y <- 2.5 * exp(-0.7 * d$x) + rnorm(200, 0, 0.15)
nlfit <- frm(bf(y ~ a * exp(-b * x), a ~ 1, b ~ 1, nl = TRUE) +
               gaussian(),
             data = d, start = list(beta = c(1, 0.3)))
unlist(fixef(nlfit)[c("a", "b")])
```

## Custom families

A family is a plain R function returning an AD-safe log-density; no
Stan or C++ involved:

```{r}
my_poisson <- custom_family(
  "my_poisson", dpars = "mu", links = list(mu = "log"),
  lpdf = function(y, dpars, aterms) {
    y * log(dpars$mu) - dpars$mu - lgamma(y + 1)
  }
)
check_custom_family(my_poisson, y = rpois(50, 3),
                    dpars = list(mu = rep(2.5, 50)))
```

## Diagnostics

`diagnose(fit)` reports convergence forensics. Simulation-based
residuals come through DHARMa:

```{r, eval = FALSE}
plot(dharma_residuals(fit))
```

And the fitted objective can be handed to NUTS for full Bayes on the
same model, where bayesplot works on the draws:

```{r, eval = FALSE}
sf <- as_tmbstan(fit, chains = 4)
bayesplot::mcmc_trace(rstan::extract(sf, permuted = FALSE))
```
