---
title: "Getting started with frmtmb"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with frmtmb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb)
has_lme4 <- requireNamespace("lme4", quietly = TRUE)
```

frmtmb fits regression models written in the brms formula grammar by
maximum likelihood, with the Laplace approximation for random effects.
The backend is RTMB: the model formula compiles to an R objective
function that is differentiated on the TMB AD tape. There is no MCMC and
no compilation at run time.

## A mixed model

```{r, eval = has_lme4}
data(sleepstudy, package = "lme4")

fit <- frm(bf(Reaction ~ Days + (Days | Subject)) + gaussian(),
           data = sleepstudy)
summary(fit)
```

The usual methods work: `fixef()`, `ranef()`, `VarCorr()`, `predict()`
(with `newdata`, `se.fit`, and `re.form`), `confint()` (Wald, profile,
or likelihood-root), `simulate()`, `anova()` for likelihood-ratio
tests, and `emmeans` for marginal means.

## Distributional regression

Every distributional parameter can get its own formula with the full
predictor grammar, including random effects and smooths:

```{r, eval = has_lme4}
fit2 <- frm(bf(Reaction ~ Days + (Days | Subject),
               sigma ~ Days) + gaussian(),
            data = sleepstudy)
fixef(fit2)$sigma
```

## Nonlinear formulas

With `nl = TRUE` the model formula is an arbitrary R expression over
named parameters, each of which has its own linear predictor:

```{r}
set.seed(1)
d <- data.frame(x = runif(200, 0, 5))
d$y <- 2.5 * exp(-0.7 * d$x) + rnorm(200, 0, 0.15)
nlfit <- frm(bf(y ~ a * exp(-b * x), a ~ 1, b ~ 1, nl = TRUE) +
               gaussian(),
             data = d, start = list(beta = c(1, 0.3)))
unlist(fixef(nlfit)[c("a", "b")])
```

## Custom families

A family is a plain R function returning an AD-safe log-density; no
Stan or C++ involved:

```{r}
my_poisson <- custom_family(
  "my_poisson", dpars = "mu", links = list(mu = "log"),
  lpdf = function(y, dpars, aterms) {
    y * log(dpars$mu) - dpars$mu - lgamma(y + 1)
  }
)
check_custom_family(my_poisson, y = rpois(50, 3),
                    dpars = list(mu = rep(2.5, 50)))
```

## Smooths and Gaussian processes

mgcv smooths and `gp()` terms work in any linear predictor; their
smoothing and kernel parameters are estimated as variance components.
`gp()` takes up to three variables (separate lengthscales per
dimension; `iso = TRUE` shares one). The exact form predicts at new
positions by kriging; the Hilbert-space form (`k =`) interpolates
through its basis, capped at `k^D <= 1000` columns. The Hilbert-space
form uses brms's input convention: the coordinates are divided by their
largest pairwise distance, so the boundary factor `c =` (1.25 by
default, one value per variable) means the same thing here as in brms.
Reported lengthscales stay in data units.

```{r}
set.seed(2)
ds <- data.frame(x = runif(200, 0, 10))
ds$y <- sin(ds$x) + rnorm(200, 0, 0.4)
fs <- frm(bf(y ~ s(x)) + gaussian(), data = ds)
fg <- frm(bf(y ~ gp(x, k = 25)) + gaussian(), data = ds)
c(smooth = AIC(fs), gp = AIC(fg))
```

## Special terms

`mo()` fits monotonic effects of ordinal predictors, `mi()` imputes
missing continuous predictors inside the model (with `mi(sdx)` for
known measurement error), and `se()` gives meta-analysis:

```{r}
set.seed(3)
dm <- data.frame(inc = sample(0:3, 300, TRUE), z = rnorm(300))
dm$y <- 1 + c(0, 1, 1.6, 2)[dm$inc + 1] + 0.3 * dm$z + rnorm(300)
fmo <- frm(bf(y ~ mo(inc) + z) + gaussian(), data = dm)
predict(fmo, newdata = data.frame(inc = 0:3, z = 0))
```

## Mixtures

Finite mixtures keep per-component parameters and covariate-dependent
mixing weights; `groups = ~g` moves the class draw to the group level
(latent classes, combinable with random effects):

```{r}
set.seed(4)
g <- rep(1:40, each = 5)
cl <- rbinom(40, 1, 0.4)
dmix <- data.frame(y = rnorm(200, c(-1, 2)[cl + 1][g], 0.8),
                   g = factor(g))
fmix <- frm(bf(y ~ 1) + mixture(gaussian(), gaussian(), groups = ~g),
            data = dmix)
head(mixture_probs(fmix), 3)
```

## Inference beyond Wald

`hypothesis()` tests arbitrary parameter expressions (Wald, profile,
or parametric bootstrap), with natural-scale random-effect names:

```{r, eval = has_lme4}
hypothesis(fit, "sd_Subject__Days^2 / (sd_Subject__Days^2 + sigma^2)")
```

`frm_bootstrap()` runs a warm-started parametric bootstrap, and
`confint(fit, method = "profile")` profiles single parameters.

## Diagnostics

`diagnose(fit)` reports convergence forensics. Simulation-based
residuals come through DHARMa, predictive checks through
`pp_check()`, and effect displays through `conditional_effects()`:

```{r, eval = FALSE}
plot(dharma_residuals(fit))
plot(conditional_effects(fit))
```

And the fitted objective can be handed to NUTS for full Bayes on the
same model - `frm_sample()` returns draws with the full method
surface (`posterior_epred()`, `hypothesis()`, `pp_check()`), and
`check_laplace()` audits the Laplace approximation against them:

```{r, eval = FALSE}
ds <- frm_sample(fit, chains = 4)
check_laplace(fit)
```

See `vignette("brms-migration")` for the brms feature map and
`vignette("diagnostics")` for the model-checking workflow.
