---
title: "Inputs and preprocessing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Inputs and preprocessing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb)
```

This is a reference page. It says what `frm()` accepts, what it does to
what you give it, what reaches the output, and what it costs to run.
For a tour of the modeling grammar, see `vignette("frmtmb")`.

## Terminology

The package uses a small vocabulary consistently. These are the terms.

**Response.** The left side of a model formula, plus its addition terms
(`weights()`, `trials()`, `cens()`, `trunc()`, `se()`, `rate()`).
A multivariate model has more than one response.

**Distributional parameter (dpar).** Every parameter of the response
distribution, not only the mean. A Gaussian response has `mu` and
`sigma`; a negative binomial has `mu` and `shape`. Auxiliary parameters
are dpars like any other: each has a link, each takes a formula with
the full predictor grammar, and each can be fixed to a constant with
`bf(..., sigma = 1)`.

**Linear predictor.** One response paired with one dpar gives one linear
predictor, keyed `<response>.<dpar>`. Its value is
`eta = X %*% beta + Z %*% b + offset`, and the dpar is `link_inverse(eta)`.
A model has as many linear predictors as it has (response, dpar) pairs.

**Link.** The monotone function that maps a dpar onto the unbounded
scale the linear predictor lives on. Links come from an AD-safe registry,
not from `stats::make.link()`, because the C level clamping in the base
links is invisible to automatic differentiation.

**Outer and inner parameters.** The optimizer moves the *outer*
parameters: the fixed-effect coefficients (`beta`, `betad`), the
covariance and smoothing parameters (`theta`), and any family extras
such as ordinal thresholds. The *inner* parameters are the
random-effect coefficients `b`, which include the wiggly parts of
smooths and the values of Gaussian process fields. They are not
optimized in the outer loop; they are integrated out.

**Laplace approximation.** The method that integrates the inner
parameters out of the joint likelihood to leave a marginal likelihood in
the outer parameters alone. It comes from TMB and RTMB.
`frmtmb.sample::check_laplace()` audits it against NUTS draws on the
same objective.

**ML and REML.** Under maximum likelihood, only `b` is inner. `REML =
TRUE` moves the fixed effects into the inner set as well, so they are
integrated out too. This is why bounds and `vcov()` blocks that name
fixed effects are refused under REML: those coefficients are no longer
outer parameters.

**Covariance structure (covstruct).** The parameterization of one
random-effect block: `us`, `diag`, `cs`, `ar1`, `ou`, `exp`, `gau`,
`mat`, `rr`, `car`, `spde`, and the rest. It decides how many `theta`
entries the block owns and how they map to a covariance matrix.

## How a formula becomes design matrices

Model building has four stages. Each one has an inspectable output, and
`dry_run =` stops the pipeline at the first two.

```
bf(y ~ x + (1 | g), sigma ~ s(z)) + gaussian()    user grammar
      |  bf(), mvbf(), +.frmtmb_formula
      v
frmtmb_formula                                    unevaluated spec
      |  parse_spec()                             [R/parse.R]
      v
frmtmb_spec                                       data-free description
      |  assemble_frame(spec, data)               [R/frame.R]
      v
frmtmb_frame                                      X, Z, blocks, parameters
      |  build_objective(frame)                   [R/objective.R]
      v
nll(pars) closure                                 data baked in as constants
      |  RTMB::MakeADFun(nll, template, random = "b")
      v
obj  ->  nlminb(obj$par, obj$fn, obj$gr)  ->  sdreport(obj)
```

**Stage 1, parse.** `parse_spec()` reads the formula alone. It never
touches the data. It splits each linear predictor into a parametric part
(bars and special terms removed), a list of random-effect terms, a list
of smooth terms, and the special terms `mo()`, `mi()`, `cs()`, `gp()`,
`car()` and `spde()`. Because the output is data free, a spec is a pure
description of the model.

**Stage 2, assemble.** `assemble_frame()` binds the spec to a data frame
and produces the matrices.

* The parametric part goes through `stats::model.frame()` and
  `stats::model.matrix()`. The resulting `terms`, `xlevels` and
  `contrasts` are stored per linear predictor, and prediction on new
  data reapplies exactly those. `frmtmb_control(sparse_x = TRUE)`
  swaps in `Matrix::sparse.model.matrix()`; the two paths are checked
  against each other so the set of dropped aliased columns cannot
  differ.
* Random-effect terms go through `reformulas::mkReTrms()`, which returns
  lme4's transposed `Zt`. frmtmb transposes each term's slice back out
  and stores `Z` with observations in rows.
* Smooths go through `mgcv::smoothCon(absorb.cons = TRUE, modCon = 3)`
  and `mgcv::smooth2random(type = 2)`. The fixed part is appended to
  `X`; the wiggly part becomes a one-variance block in `Z`, so the
  smoothing parameter is estimated as a variance component. Prediction
  uses `mgcv::PredictMat()` with the stored transforms. `modCon = 3`
  makes that basis the fitted basis: a `t2()` smooth otherwise carries a
  second, sum-to-zero prediction constraint that `PredictMat()` honors
  and the fit does not. It changes no fitted quantity. `s()` and `t2()`
  are supported, `te()` is not.
* Gaussian processes, `car()` and `spde()` build their own sparse
  columns and carry their auxiliary data (distances, adjacency,
  finite-element matrices) on the block.
* `offset()` terms are summed per linear predictor and added to `eta`
  after `X %*% beta + Z %*% b`. `weights()` is not a design matrix at
  all: it is an addition term that multiplies the per-observation log
  density.

Every linear predictor's `Z` spans the whole random-effect coefficient
vector, so a `Z` column index *is* a coefficient index. Columns are
ordered level major and coefficient minor inside a block, at
`(level - 1) * dim + offset + coefficient`. Terms linked with an `|ID|`
key are merged into one block, which is only a matter of where their
columns land.

**Stage 3, generate.** `build_objective()` closes over the assembled
matrices and returns a plain R function of the parameter list. Family,
link and covariance dispatch all resolve before taping, so the closure
stays vectorized and free of branches on parameters.

**Stage 4, tape and optimize.** `RTMB::MakeADFun()` tapes the closure
with `random = "b"` (and `"beta"` under REML), `nlminb()` moves the outer
parameters, and `sdreport()` supplies standard errors.

You can look at the first two stages without fitting:

```{r}
set.seed(1)
d <- data.frame(y = rnorm(30), x = rnorm(30), g = factor(rep(1:6, 5)))
rownames(d) <- paste0("obs", seq_len(30))

spec <- frm(bf(y ~ x + (1 | g)) + gaussian(), data = d, dry_run = "spec")
names(spec$responses$y$dpars)
spec$responses$y$dpars$mu$fixed

frame <- frm(bf(y ~ x + (1 | g)) + gaussian(), data = d, dry_run = "frame")
names(frame$linpreds)
c(X = dim(frame$linpreds[["y.mu"]]$X),
  Z = dim(frame$linpreds[["y.mu"]]$Z))
lengths(frame$par_template)
```

`beta` holds the coefficients of the mean linear predictors, `betad`
those of the other dpars, `b` the random-effect coefficients, and
`theta` the covariance and smoothing parameters. Here the outer problem
has four parameters and the inner problem six.

## Argument types and lengths

`frm()` is the only entry point. There is no matrix-only interface: the
model is always a formula and a data frame.

| Argument | Type and length | On mismatch |
|---|---|---|
| `formula` | One `bf()` object with a family attached by `+`, or one plain two-sided formula with `family =` given separately. | A formula with no left side, or no family from either source, is an error. |
| `data` | One data frame. Its `nrow()` sets the number of observations. Every variable in every linear predictor must resolve in it. | A missing variable is an error naming the variable and the linear predictor. |
| `family` | A `frmtmb_family`, a `stats::family` object or its constructor function, or a length-one family name. Overrides any family carried by `formula`. | An unsupported name errors and lists every supported name. |
| `REML` | Length-one logical. | Moves the fixed effects into the inner set, which makes `drop1()` and fixed-effect bounds refuse. |
| `start` | Named list, one entry per parameter component. `par_template(formula, data)` returns it filled with the defaults, for a formula or for a fit. A named vector inside it addresses entries by name and leaves the rest alone; an unnamed one is positional and must be full length. | An unknown component is an error that points at the template. An unknown parameter name lists the valid ones. Naming some entries of one vector and not others is refused. |
| `control` | The list from `frmtmb_control()`. | See below. |
| `se` | Length-one logical, default `FALSE`. Defers `sdreport()`, which is roughly a quarter of the fit time. | |
| `na.action` | A function, as in `stats::lm()`. Default `stats::na.omit`. | `na.fail` errors on any `NA`. `na.exclude` drops rows for the fit and pads `fitted()`, `residuals()`, `predict()` and `simulate()` back to the input length with `NA` in the original positions. |
| `prior` | A `set_prior()` specification, or a `brmsprior` object brms built, which is translated row by row. Makes the fit penalized (MAP), so the reported log likelihood is penalized too. Spelled as brms spells it; the `priors` of releases before 0.43 is gone, and a call still using it fails as an unused argument. | `set_prior()` checks the argument count of each distribution. A brms prior class with no faithful frmtmb spelling is refused by name rather than translated into a different density. |
| `quadrature` | Length-one logical. Scalar random-intercept models only. | Any other structure, and `mi()`, `trunc()`, `REML = TRUE` or `profile = TRUE`, is an error naming the conflict. |
| `data2` | Named list of objects that the formula names and `data` cannot hold: the adjacency matrix of `car()`, the mesh triple of `spde()`, and the matrices of `gr(prec = )`, `gr(cov = )` and `equalto()`. | Anything that is not a named list is an error. A name a formula asks for and does not find is an error naming it. |
| `dry_run` | `NULL`, `"spec"`, `"frame"` or `"objective"`. | Any other value is ignored and the model fits normally. |
| `verbose` | `FALSE`, `TRUE`, or an integer level: 0 silent, 1 stage progress, 2 optimizer trace. An explicit `control$verbose` wins. | Never errors. `NA` and negative values become 0. |

`frmtmb_control()` takes `optimizer` (`"nlminb"`, `"optim"`, or a
function), `optCtrl`, `restarts`, `grad_tol`, `profile`, `sparse_x`,
`autoscale`, `check_nlev_1`, `check_olre` and `verbose`.

`predict()` takes `newdata` (a data frame), `type`, `dpar`, `resp`,
`re.form`, `se.fit` and `allow_new_levels`. Note the spelling `re.form`,
which follows lme4 rather than brms `re_formula`. Unknown arguments in
`...` produce a warning, not an error, because `allow.new.levels` is
accepted as an lme4 and glmmTMB spelling.

## Predictor classes

### Accepted

| Class | Treatment |
|---|---|
| `numeric` | Used directly. |
| `factor` | Contrasts are read from the column and stored, then reapplied to new data. The package never creates a factor for a predictor; that would change the model silently. |
| `ordered` factor | Kept ordered, with its polynomial contrasts. `mo()` requires an ordered factor with at least three levels. |
| `logical` | Handled as a two-level factor by `model.matrix()`, giving one `TRUE` column. Also accepted directly as a `mo()` or `mi()` interaction multiplier. |
| `character`, as a grouping variable | Accepted. An integer, a factor and a character grouping variable give the same fit. |
| `character`, in a fixed-effect term | Works, but by delegation: `stats::model.frame()` coerces it to a factor with the default contrasts. Level order then follows the locale collation, not your data. Convert to a factor yourself if the level order matters. |
| Matrix-valued terms | `poly()`, `ns()`, `scale()` and matrix covariates for functional regression are supported. Their basis is frozen at fit time through `predvars`, so new data is projected onto the fitted basis rather than refitted. |
| `Date`, `POSIXct`, `difftime` | Accepted as a number. `model.matrix()` reduces the column to its underlying value, so the coefficient is per day (`Date`), per second (`POSIXct`), or per unit of the `units` attribute (`difftime`). Frame assembly reports the coercion, naming the column, its unit and its origin. See "Dates and times" below. |

### Rejected

These raise an error rather than fitting something surprising.

| Input | Behavior |
|---|---|
| List column in any term | Error from `model.frame()`: `invalid type (list) for variable`. There is no frmtmb-level check, so the message names the variable but not the model. |
| `character` or `factor` multiplier in a `mo()` or `mi()` interaction | Error. `as.numeric()` on a character vector gives all `NA`, and the only symptom would be an `NA/NaN gradient evaluation` from the optimizer. The same type gate runs again on `newdata`. |
| Unordered factor in `mo()`, or fewer than three categories | Error. |
| Factor response with a family that is neither ordinal nor binomial | Error. |
| `Inf` or `NaN` in the response | Error. This is deliberately separate from `NA`, which `na.action` owns. |
| Matrix response under `mi()` | Error. |
| `cs()` grouping that is not a single factor | Error. |
| `gr(cov =)` or `gr(prec =)` that is not square, or whose dimnames miss a grouping level | Error. |
| `car(type =)` outside `"escar"`, `"icar"`, `"esicar"`, `"bym2"` | Error listing the four types. |

One case warns instead of failing: `ar1()` and `hetar1()` correlate
levels by position, so whole-number factor levels with a gap in them
(`1, 2, 4`) get a warning that the gap counts as one step, with a
pointer to `ou(num_factor(...))` for irregular spacing.

```{r, eval = FALSE}
# a list column stops at model.frame(), before any frmtmb check
d$lst <- I(lapply(seq_len(30), function(i) 1:2))
frm(bf(y ~ lst) + gaussian(), data = d)
#> Error: invalid type (list) for variable 'lst'
```

### Dates and times

A `Date`, `POSIXct` or `difftime` predictor is fitted as the number
underneath it. The model is the one you asked for, but it is expressed
in an origin you did not choose: `as.numeric()` on a `Date` counts days
from 1970-01-01, and on a `POSIXct` it counts seconds. So the slope is
per day or per second, and the intercept is the fitted value on
1970-01-01.

That intercept is the problem. A modern date is about 18000 as a
number, and a modern timestamp about 1.7e9, which puts the intercept
far outside the data and makes the objective badly conditioned. The
same model on days-since-the-first-day converges where the raw column
reports false convergence, with an identical slope:

```{r, eval = FALSE}
d <- data.frame(day = as.Date("2020-01-01") + 0:59)
d$y <- rnorm(60, 1 + 0.02 * as.numeric(d$day - min(d$day)), 0.5)

frm(bf(y ~ day) + gaussian(), data = d)
#> Date/time column used as a number: day (Date, days since 1970-01-01).
#> The coefficient is per unit of that origin and the intercept is the
#> value at it, which is far outside the data and can stop the optimizer
#> converging. Center the column, for example as.numeric(x - min(x)), to
#> put the intercept back in range.
#> Warning: Optimizer did not report convergence: false convergence (8)
#> (Intercept) = -366.16, day = 0.0201

# the same fit, with the intercept back inside the data
d$days <- as.numeric(d$day - min(d$day))
frm(bf(y ~ days) + gaussian(), data = d)
#> (Intercept) = 1.051, days = 0.0201
```

Frame assembly emits that message once per fit. It is a message and not
a warning because the coercion is deliberate and correct in meaning;
`suppressMessages()` silences it once you have centered the column or
if you want the epoch origin.

## Character options and case

All character options are matched with `match.arg()`, so matching is
case sensitive and accepts unambiguous prefixes. `"response"` and
`"resp"` both work for `predict(type =)`; `"Response"` does not.

| Function | Argument | Values |
|---|---|---|
| `predict()` | `type` | `"link"`, `"response"`, `"conditional"`, `"zprob"`, `"zlink"`, `"disp"` |
| `residuals()` | `type` | `"response"`, `"pearson"`, `"deviance"`, `"osa"` |
| `confint()` | `method` | `"wald"`, `"Wald"`, `"profile"`, `"uniroot"`, `"boot"` |
| `drop1()` | `test` | `"none"`, `"Chisq"` |
| `hypothesis()` | `method` | `"wald"`, `"profile"`, `"boot"` |
| `conditional_effects()` | `method` | `"epred"`, `"predict"` |
| `anova()` on `frm_multiple()` | `method`, `use` | `"D3"`, `"D1"`, `"D2"`; `"likelihood"`, `"wald"` |
| `frmtmb_control()` | `check_nlev_1`, `check_olre` | `"warning"`, `"ignore"`, `"stop"` |
| `set_prior()` | `class` | `"b"`, `"Intercept"`, `"sd"`, `"cor"`, `"theta"` |
| `conditional_effects()` | `band` | `"wald"`, `"profile"`, `"boot"` |
| `loo_compare()` | `criterion` | `"loo"`, `"waic"` |
| `vcov_cluster()` | `type` | `"CR0"`, `"CR1"`, `"CR1p"`, `"CR1S"` |

`confint(method =)` is the one place that accepts two spellings of the
same choice: `"Wald"` is listed alongside `"wald"` and normalized
immediately, because both spellings are common in other packages.

Options that do not use `match.arg()` are matched exactly, with no
prefix matching: `car(type =)`, a family given as a string,
`getME(name =)`, `frm(dry_run =)` and `frmtmb_control(optimizer =)`.

The single case-insensitive option in the package is the code vocabulary
of `cens()`. There, `"right"`, `"Right"` and `"R"` are the same code,
because a capitalized word reads as a spelling variant rather than as
garbage. The numeric codes `0`, `-1`, `1`, `2` mean the same thing.

## Naming collisions

Four vocabularies meet in a model: the columns of `data`, the
distributional parameters of the family, the nonlinear parameters a
formula declares, and the display names that `hypothesis()`,
`variables()` and `set_prior()` use. They namespace apart by
construction, so a column may carry any name at all. The rules below
say what happens where two of them do meet.

**Columns and coefficients never collide.** A coefficient is stored
under its dpar prefix, so a covariate named `sigma` in the mean is
`sigma` in `fixef(fit)$mu` while the residual standard deviation is
`(Intercept)` in `fixef(fit)$sigma`. Both fit, and neither reads the
other.

**A nonlinear parameter may not also be a column.** `bf(y ~ a *
exp(-b * x), a ~ 1, b ~ 1, nl = TRUE)` on data that has a column `a` is
refused: the body would use the parameter and ignore the column
silently. Rename one of them. A declared nonlinear parameter that no
body uses is refused too, and so is a dpar formula for a parameter the
family does not have (`bf(y ~ x, shape ~ 1)` on `gaussian()`, which
names `sigma` as the one available).

**In a nonlinear body, a column wins over a distributional parameter.**
A name in an `nlf()` body resolves in this order: a nonlinear parameter
of the model, then a column of `data`, then another distributional
parameter of the same response (read per row, after its link inverse).
The dpar reference is an extension of brms, which has only the first
two; putting the column ahead of it is what keeps every body brms
accepts meaning here what it means there. So
`nlf(sigma ~ ls + th * log(abs(mu)))` is a variance function of the
model's own mean, unless `data` carries a column `mu`, in which case it
is a variance function of that column.

**In a nonlinear body, a function resolves in RTMB first.** That order
is for the names a body READS. A name a body CALLS follows a different
one: RTMB, then the formula's own environment, then the search path. A
bare `pnorm()`, `qgamma()`, `dbeta()`, `besselK()` or `matrix()` in a
body is therefore RTMB's tape-capable version, with no `RTMB::` prefix
and no `library(RTMB)`, and it beats a function of the same name that
you defined yourself. Qualify with `stats::` to reach the other one.
Only the names RTMB replaces are shadowed, so a helper of your own is
found where it always was; and because only the CALLED names are
shadowed, a data frame named `df` or a vector named `order` still reads
as your object.

**The internal `.eta_<dpar>` names are unreachable from `data`.** The
objective keeps each linear predictor beside its parameter under a
reserved `.eta_` name, for the families that need the untransformed
scale. Those live in an internal list rather than in the model frame,
so a column named `.eta_mu` is an ordinary covariate and gets an
ordinary coefficient.

**In `hypothesis()`, a coefficient shadows a natural-scale name.**
`hypothesis()` evaluates expressions in one flat namespace that holds
the coefficients and the natural-scale summaries together, and the
coefficient wins:

```{r, eval = FALSE}
# a covariate literally named sigma
fit <- frm(bf(y ~ sigma + (1 | g)), family = gaussian(), data = dd)
hypothesis(fit, "sigma = 0")
#> hypothesis() reads 'sigma' as the coefficient of the model term of
#> that name, not the residual standard deviation; that quantity is
#> available as '.sigma'.
```

The message names both meanings and fires once per call. The shadowed
quantity keeps a name of its own: prefix it with a dot. The same holds
for a coefficient that spells out `sd_<group>__<term>`, `cor_...`, or
an autocorrelation name such as `ar1`; `.sd_g__Intercept` and `.ar1`
then reach the natural-scale value. The dot spelling exists only where
a collision does, and `variables()` lists both names in that case.

## What reaches the output

### Preserved

* **Row names.** The row names of `data` reach `fitted()`, `predict()`,
  `residuals()`, `model.frame()` and the `Z` matrix from
  `getME(fit, "Z")`. They are carried by the design matrix, not
  reassigned, so a row name survives a subset and a reorder.
* **Coefficient names.** `vcov()`, `confint()` and `fixef()` share one
  naming scheme.
* **Factor levels and contrasts.** The level set and the contrast matrix
  of every factor are stored per linear predictor and reapplied to
  `newdata`. A level missing from `newdata` is still a known level; a
  level that is new is an error unless `allow_new_levels = TRUE`.
* **Data-dependent bases.** `poly()`, `ns()` and `scale()` on the
  predictor side are frozen through `predvars`, so `newdata` uses the
  fitted centering, scaling and knots.
* **Row positions under `na.exclude`.** Output is padded back to the
  input length with `NA` in the dropped positions.

```{r}
fit <- frm(bf(y ~ x + (1 | g)) + gaussian(), data = d)
head(names(fitted(fit)), 3)
head(names(predict(fit)), 3)
```

### Not preserved

* **Attributes on the response.** The response reaches the objective as
  `as.numeric(as.vector(y))`, or `storage.mode(y) <- "double"` for a
  matrix response. A `scale()`d response therefore arrives without its
  `"scaled:center"` and `"scaled:scale"` attributes, and a one-column
  matrix response arrives without its `dim`. If you need to invert the
  transformation later, keep the attributes yourself.
* **Column attributes such as `units` or `haven` labels.** Nothing in
  the package inspects, preserves or warns about them. `model.frame()`
  and `model.matrix()` drop them, and the coefficient is reported in
  whatever numeric units the column held.
* **`dimnames` on a matrix covariate.** Design column names come from
  the term labels, not from the covariate's own dimnames.
* **Rows dropped by `na.omit`.** They are absent from `fitted()` and
  `residuals()`. Use `na.action = stats::na.exclude` to get them back as
  `NA`.

## Runtime and how it scales

A fit has four costs, and only one of them grows quickly.

| Stage | Behavior with data size |
|---|---|
| parse | Independent of the data. Effectively free. |
| frame | Linear in `n` and in the number of random-effect columns. |
| tape | One pass over the objective. Independent of the number of optimizer iterations. |
| optimize | (number of outer iterations) x (cost of one objective evaluation). This dominates. |
| sdreport | One Hessian and one sparse solve at the optimum. Skipped unless `se = TRUE` or a method needs it. |

One objective evaluation costs an inner Newton solve, whose work is the
sparse Cholesky factorization of the inner Hessian. That matrix is
(number of random-effect coefficients) square, and its sparsity comes
from the grouping structure, so the cost tracks the number and the
overlap of the random-effect levels rather than `n` alone. The number of
outer iterations tracks the number of outer parameters and the
conditioning of the problem, not the number of rows.

Measured on `lme4::InstEval` (n = 73421, 2972 students, 1128 lecturers)
for `y ~ service + (1 | s) + (1 | d)`, which tapes to 5 outer and 4100
inner parameters with a 4100 x 4100 inner Hessian holding 77521
nonzeros:

| Stage | Seconds |
|---|---|
| parse | 0.00 |
| frame | 0.14 |
| tape | 0.66 |
| optimize | 22.78 |
| restart | 0.44 |
| sdreport | 3.84 |
| total | 28.63 |

Everything outside `optimize` is roughly fixed: parse, frame and tape
together are about 0.8 s, and `sdreport()` about 3.8 s. `optimize` is
80 percent of the fit, and 100 percent of `optimize` is objective
evaluation: `nlminb` itself costs nothing measurable on a five parameter
problem. The first `obj$fn` call is about 1.6 s because the inner solve
starts cold; every later one is about 0.17 s because it warm starts from
the previous point, and `obj$gr` is about 0.18 s. This model takes 40
outer iterations, so halving the iteration count roughly halves the fit.

Fit times across sizes, all wall clock seconds on the same machine:

| Model | frmtmb | glmmTMB | lme4 |
|---|---|---|---|
| LMM, n = 180 (`sleepstudy`) | 0.19 | 0.08 | 0.03 |
| LMM, n = 50000, 200 groups, `(x | g)` | 4.61 | 2.86 | 0.35 |
| Poisson GLMM, n = 100000, 500 groups | 6.00 | 5.18 | 9.89 |
| Negative binomial with `dispformula`, n = 20000 | 1.31 | 0.71 | |
| Zero-inflated Poisson, n = 20000 | 0.47 | 0.58 | |
| LMM, n = 73421 (`InstEval`) | 21.50 | | 4.42 |

Read this as context, not as a race. `lme4::lmer()` is about 4.9 times
faster on `InstEval` because it profiles both the fixed effects and the
residual scale out of the deviance, leaving a two parameter problem over
a purpose-built sparse Cholesky, while frmtmb hands `nlminb` five
parameters over a generic Laplace tape. The two agree on the objective
to four decimal places (118865.3077). frmtmb is faster than `lme4` on
the large Poisson GLMM, where `glmer`'s adaptive quadrature and
penalized iteratively reweighted least squares scale worse than Laplace
with exact derivatives.

Three controls change the runtime measurably on `InstEval`:

* `frmtmb_control(profile = TRUE)` takes it from 21.50 s to 13.54 s,
  and should help more as the number of fixed effects grows.
* `frmtmb_control(optimizer = "optim")` takes it to 10.44 s, but it is
  slower than `nlminb` on the intercept-only version of the same model
  at matched tolerance, so it is a per-model option rather than a
  better default.
* `REML = TRUE` takes it to 12.06 s.

`frmtmb_control(sparse_x = TRUE)` changes nothing here, because `X` is
73421 by 2. It is for genuinely sparse and wide fixed-effect designs.

## Reproducing the timings

Every number above comes from a script in `dev/` of the source
repository at <https://github.com/aforren1/frmtmb>. `dev/` is not part
of the installed package, so clone the repository to run them.

| Script | Produces |
|---|---|
| `dev/bench-env.R` | machine, R and package versions |
| `dev/bench-setup.R` | one-time install into a scratch library |
| `dev/bench-op-baseline.R` | the per-stage table |
| `dev/bench-op-profile.R` | the split of `optimize` into `obj$fn` and `obj$gr` |
| `dev/bench-op-ceiling.R` | standalone per-call costs |
| `dev/bench-op-context.R` | the frmtmb, glmmTMB and lme4 comparison |
| `dev/bench-op-tight.R` | the same comparison at matched convergence tolerance |
| `dev/bench-rtmbp-core.R` | the parallel-backend appendix |

The comparison table is one command:

```sh
Rscript dev/bench-op-context.R
```

Run `Rscript dev/bench-setup.R` once first, and edit the scratch library
path at the top of each script to a directory you can write to. Run the
timing scripts one at a time; the numbers move by about 15 percent with
another R process on the machine.

The headline frmtmb against lme4 comparison does not need the harness.
This one line reproduces it from an installed frmtmb:

```sh
Rscript -e 'library(frmtmb); data(InstEval, package = "lme4"); print(system.time(frm(bf(y ~ service + (1 | s) + (1 | d)) + gaussian(), data = InstEval))); print(system.time(lme4::lmer(y ~ service + (1 | s) + (1 | d), data = InstEval, REML = FALSE)))'
```

The findings behind these numbers, including why parallelizing the outer
optimizer does not help and why an RTMB tape cannot be shipped to a
worker process, are written up in `dev/benchmarks.md`.

```{r srr-tags, eval = FALSE, echo = FALSE}
#' srr_stats_inputs
#'
#' Standards closed by this vignette.
#'
#' @srrstats {G1.3} The "Terminology" section defines every statistical
#'   term the package uses in its own documentation and messages:
#'   response, distributional parameter, linear predictor, link, outer
#'   and inner parameters, the Laplace approximation, ML against REML,
#'   and covariance structure.
#' @srrstats {G1.5} The performance claims of the package are reproduced
#'   by the scripts listed in "Reproducing the timings", each of which is
#'   named against the table or claim it produces.
#' @srrstats {G1.6} The comparison against alternative implementations
#'   (lme4 and glmmTMB) is reproduced by `dev/bench-op-context.R`, and
#'   the headline frmtmb against lme4 comparison by the single command
#'   given in the same section.
#' @srrstats {G2.0a} "Argument types and lengths" documents the expected
#'   length of every argument of `frm()`, including the arguments that
#'   must be length one and the named vectors whose names are checked
#'   against the outer parameter vector.
#' @srrstats {G2.1a} The same table documents the expected type of every
#'   argument of `frm()` and of `predict()`, and "Predictor classes"
#'   documents the expected type of every column of `data`.
#' @srrstats {G2.3b} "Character options and case" states that every
#'   character option is matched with `match.arg()` and is therefore case
#'   sensitive, lists the allowed values, and names the two deliberate
#'   exceptions: `confint(method = "Wald")` and the case-insensitive
#'   `cens()` codes.
#' @srrstats {G2.5} "Dates and times" documents the one class that is
#'   accepted but not used as itself: a `Date`, `POSIXct` or `difftime`
#'   predictor is fitted as the number underneath it, so the section
#'   states the unit and the origin the coefficient and the intercept
#'   come out in, shows the false convergence the epoch origin causes,
#'   and gives the centering that fixes it. Frame assembly reports the
#'   same coercion per fit, naming the column.
#' @srrstats {RE1.1} "How a formula becomes design matrices" documents
#'   how the model object is constructed, stage by stage, from the
#'   formula through the data-free spec and the assembled frame to the
#'   taped objective, and names the packages each stage reuses.
#' @srrstats {RE1.2} "Predictor classes" documents the expected format of
#'   the input data: which column classes are accepted and how each one
#'   is treated, and which are rejected and with what message.
#' @srrstats {RE1.3a} "What reaches the output" documents both halves:
#'   the aspects of the input that are transferred to the output (row
#'   names, coefficient names, factor levels and contrasts, frozen
#'   data-dependent bases, row positions under `na.exclude`) and the
#'   aspects that are not (attributes on the response, `units` and label
#'   attributes on columns, dimnames of matrix covariates, rows dropped
#'   by `na.omit`).
#' @srrstats {RE5.0} "Runtime and how it scales" documents the scaling of
#'   fit time with data size and with model structure: which stages are
#'   fixed cost, that the optimize stage dominates, that one objective
#'   evaluation is a sparse Cholesky over the random-effect coefficients
#'   rather than a pass over the rows, and measured times from n = 180 to
#'   n = 100000.
```
