---
title: "Differential equation models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential equation models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb)
has_ode <- requireNamespace("RTMBode", quietly = TRUE)
# figures are a second, independent option: the vignette renders whole
# without either package
has_plot <- has_ode && requireNamespace("tinyplot", quietly = TRUE)
```

Some models say what the *rate of change* of a quantity is, not what
the quantity is. A drug moves from a depot into the blood and is
cleared from it; a population grows and is eaten. `frm_ode()` puts such
a system inside a nonlinear formula, so the constants of the system are
ordinary nonlinear parameters with fixed effects, random effects and
covariates, and the model is fitted by maximum likelihood like any
other `frm()` model.

## Setup

`frm_ode()` needs the **RTMBode** package, which supplies a
differentiable interface to the **deSolve** integrators. It is not on
CRAN:

```{r, eval = FALSE}
install.packages("RTMBode", repos = c(
  "https://kaskr.r-universe.dev",
  "https://cloud.r-project.org"))
```

Everything below is skipped if it is not installed.

## A population pharmacokinetic model

The standard one-compartment model with first-order oral absorption has
two states: the amount `A` still in the depot, and the concentration `C`
in the central compartment.

$$\frac{dA}{dt} = -k_a A, \qquad
  \frac{dC}{dt} = \frac{k_a A}{V} - k_e C$$

A single dose `D` enters the depot at time zero, so $A(0) = D$ and
$C(0) = 0$. Three constants describe a subject: the absorption rate
$k_a$, the elimination rate $k_e$, and the apparent volume of
distribution $V$. All three must be positive, so they are estimated on
the log scale.

The system itself is a plain R function in the **deSolve** convention:
it takes the time, the state vector and the parameter vector, and it
returns the derivatives in a list. Index `y` and `p` by position.

```{r}
pk_dyn <- function(t, y, p) {
  # base c() drops the automatic-differentiation class; this restores it
  "c" <- RTMB::ADoverload("c")
  list(c(-p[1] * y[1],
         p[1] * y[1] / p[3] - p[2] * y[2]))
}
```

`datasets::Theoph` is the classic example: 12 subjects, one oral dose of
theophylline each, 11 serum concentrations over 25 hours.

```{r}
d <- datasets::Theoph
d$Subject <- factor(as.character(d$Subject))
head(d, 3)
```

The model is one `bf(nl = TRUE)` formula. The body says how a
concentration is produced; the three nonlinear parameter formulas say
how each constant varies between subjects.

```{r, eval = has_ode}
form <- bf(
  conc ~ frm_ode(pk_dyn,
                 init   = list(Dose, 0),
                 times  = Time,
                 parms  = list(exp(lka), exp(lke), exp(lV)),
                 group  = Subject,
                 states = c("depot", "central"),
                 output = "central"),
  lka ~ 1 + (1 | Subject),
  lke ~ 1 + (1 | Subject),
  lV  ~ 1,
  nl = TRUE
)

fit <- frm(form + gaussian(), data = d, se = TRUE,
           start = list(beta = c(0.5, log(0.08), log(0.5))))
summary(fit)
```

The typical subject's constants come back on the natural scale, and the
derived quantities a pharmacokineticist reads off them follow:

```{r, eval = has_ode}
ka <- exp(fixef(fit)$lka); ke <- exp(fixef(fit)$lke)
V <- exp(fixef(fit)$lV)
c(ka = ka, ke = ke, V = V,
  half_life = log(2) / ke,       # hours
  clearance = ke * V)            # L/h per kg of body weight
```

Everything downstream works as usual: `predict()` on new times,
`ranef()` for the subject deviations, `confint()`, `simulate()`,
`REML = TRUE`.

```{r, eval = has_ode}
nd <- data.frame(Subject = factor("1", levels = levels(d$Subject)),
                 Time = seq(0, 25, length.out = 5), Dose = 4.02)
predict(fit, newdata = nd)
```

The same call on a dense time grid draws the fitted curve for every
subject over that subject's own observations. Each curve carries that
subject's random effects, which is why they differ in both height and
shape.

```{r fig-theoph, eval = has_plot, fig.width = 7, fig.height = 4.5, fig.alt = "Theophylline concentration against time for twelve subjects, with the fitted curve of each subject drawn through its own points."}
grid <- do.call(rbind, lapply(split(d, d$Subject), function(s) {
  data.frame(Subject = s$Subject[1],
             Time = seq(0, max(d$Time), length.out = 100),
             Dose = s$Dose[1])
}))
grid$conc <- predict(fit, newdata = grid)

tinyplot::tinyplot(conc ~ Time | Subject, data = d, pch = 16, cex = 0.7,
                   legend = FALSE, xlab = "Time (h)",
                   ylab = "Concentration (mg/L)",
                   main = "Theophylline, fitted per subject")
tinyplot::tinyplot_add(conc ~ Time | Subject, data = grid, type = "l")
```

Each subject is observed once, so the curve and the points share a
colour. The absorption peak moves between subjects because `lka` has a
random effect.

## How the arguments work

`frm_ode()` runs **one solve per group**, over that group's own
observation times, then puts the results back in the row order of the
data.

| Argument | What it is |
|---|---|
| first argument | The derivative function, `function(t, y, parms)`. |
| `init` | Initial states, one list element per state. |
| `times` | The observation time of every row. |
| `parms` | The constants of the system, one list element per constant, in the order the derivative function indexes them. |
| `group` | The unit that owns one system: a subject, a reactor, a batch. |
| `states`, `output` | State names, and which of them the body returns. |
| `t0` | The initial time. Defaults to 0. |
| `events`, `event_scale` | A dosing table, and an estimated multiplier on its amounts. |
| `tv`, `tv_break` | Constants that change with time, and where they change. |
| `n_ss`, `ss_tol` | The length of a steady-state run-in, and when to complain about it. |
| `method`, `atol`, `rtol` | Integrator and tolerances. |

`init` and `parms` are given as lists of **columns**, not as one row of
values per state. Each element is either one value per observation or a
single value shared by every group. `init = list(Dose, 0)` is therefore
"the depot starts at this row's `Dose`, the central compartment starts
at zero". A bare vector such as `c(100, 0)` is refused, because it
cannot be told apart from one column of two observations.

With no `output` the result is a matrix with one column per state, so
`frm_ode(...)[, 2]` also works. Naming the states and selecting one by
name is clearer and is what the example above does.

## Repeated dosing

Theophylline is a single dose, and a single dose is just an initial
condition. A course of treatment is not: the depot is refilled at each
dose time, and the trajectory is the sum of what every dose so far has
contributed. `events` is the table of those doses.

```{r}
doses <- data.frame(time = c(12, 24, 36), state = "depot", value = 100)
doses
```

One row per dose. `time` is when, `state` is which compartment (by name
if `states` names them, otherwise by position), `value` is how much, and
the default `method = "add"` puts the amount into the state. Give a row
a positive `duration` and it becomes an infusion instead, delivering
`value` at a constant rate over `[time, time + duration]`. A `group`
column restricts a row to one subject; leave the column out and the
schedule applies to every subject.

A regular course does not have to be written out. `ii` is the interdose
interval and `addl` the number of doses that follow the row's own, so
these two tables are the same table:

```{r}
data.frame(time = 0, state = "depot", value = 100, ii = 12, addl = 3)
data.frame(time = c(0, 12, 24, 36), state = "depot", value = 100)
```

The expansion happens before anything else looks at the table, so the
two give bit-identical results.

In NONMEM terms an `"add"` row is a dosing record, `value` is `amt`,
`state` is `cmt`, `duration` sets `rate = amt / duration`, and `ii` and
`addl` are spelled the same way. There is no `evid` column, because
observations and doses live in two separate tables here: the rows of
`data` are the observations, the rows of `events` are the doses.

Solving directly, with the first dose as the initial condition and three
more as events:

```{r, eval = has_ode}
tt <- seq(0, 60, by = 0.25)
traj <- frm_ode(pk_dyn, init = list(100, 0), times = tt,
                parms = list(1, 0.2, 10),
                states = c("depot", "central"), output = "central",
                events = doses)
range(traj)
```

```{r fig-doses, eval = has_plot, fig.width = 7, fig.height = 4, fig.alt = "Concentration against time for four doses given twelve hours apart, rising to a plateau, with a vertical line at each dose time."}
tinyplot::tinyplot(tt, traj, type = "l", lwd = 2,
                   xlab = "Time (h)", ylab = "Concentration (mg/L)",
                   main = "Four doses, twelve hours apart")
abline(v = c(0, doses$time), lty = 3, col = "grey40")
```

Each peak is higher than the last and the troughs level off, so the
course approaches a steady state instead of repeating the first dose.
That accumulation is why a dosing schedule cannot be faked with one
larger dose.

### Inside a formula

A nonlinear body looks up every bare name in `data` first, so a column
called `doses` wins. There is no column of that name here, and a
data.frame is not something a column could hold, so the table is read
from the environment of the formula:

```{r, eval = FALSE}
conc ~ frm_ode(pk_dyn, init = list(0, 0), times = time,
               parms = list(exp(lka), exp(lke), exp(lV)),
               group = id, states = c("depot", "central"),
               output = "central", events = doses)
```

Writing the table inline says the same thing, and a function of no
arguments is the way to read the schedule at fit time:

```{r, eval = FALSE}
conc ~ frm_ode(pk_dyn, ..., events = data.frame(
                 time = c(0, 12, 24), state = "depot", value = 100))

schedule <- function() read.csv("doses.csv")
conc ~ frm_ode(pk_dyn, ..., events = schedule)
```

A fit is otherwise unchanged. Here is a small simulated course of three
doses twelve hours apart, with between-subject variability on `lka` and
`lke`:

```{r, eval = has_ode}
set.seed(4)
n_id <- 8
tt_obs <- c(0.5, 2, 6, 11.9, 14, 20, 23.9, 26, 32, 40)
dd <- data.frame(id = factor(rep(seq_len(n_id), each = length(tt_obs))),
                 time = rep(tt_obs, n_id))
ka_i <- exp(rnorm(n_id, 0, 0.3))[as.integer(dd$id)]
ke_i <- exp(rnorm(n_id, log(0.2), 0.25))[as.integer(dd$id)]
super <- function(t, ka, ke, V, amt, at) {
  u <- t - at[at <= t]
  sum(amt * ka / (V * (ka - ke)) * (exp(-ke * u) - exp(-ka * u)))
}
dd$conc <- vapply(seq_len(nrow(dd)), function(i)
  super(dd$time[i], ka_i[i], ke_i[i], 10, 100, c(0, 12, 24)), 0) +
  rnorm(nrow(dd), 0, 0.3)

dose_fit <- frm(
  bf(conc ~ frm_ode(pk_dyn, init = list(100, 0), times = time,
                    parms = list(exp(lka), exp(lke), exp(lV)),
                    group = id, states = c("depot", "central"),
                    output = "central",
                    events = data.frame(time = c(12, 24),
                                        state = "depot", value = 100)),
     lka ~ 1 + (1 | id), lke ~ 1 + (1 | id), lV ~ 1, nl = TRUE) +
    gaussian(),
  data = dd, start = list(beta = c(0, log(0.25), log(8))))
unlist(fixef(dose_fit))
```

The truth is `lka = 0`, `lke = -1.61` and `lV = 2.30`, with a residual
standard deviation of 0.3, which is `-1.20` on the log scale `sigma` is
reported on.

```{r fig-dosefit, eval = has_plot, fig.width = 7, fig.height = 4.5, fig.alt = "Observed concentrations and fitted curves for four subjects on a three dose course, with a vertical line at each dose time."}
show <- levels(dd$id)[1:4]
sub <- dd[dd$id %in% show, ]
sub$id <- droplevels(sub$id)
gr <- do.call(rbind, lapply(show, function(s)
  data.frame(id = factor(s, levels = levels(dd$id)),
             time = seq(0, 40, length.out = 200))))
gr$conc <- predict(dose_fit, newdata = gr)
gr$id <- droplevels(gr$id)

tinyplot::tinyplot(conc ~ time | id, data = sub, pch = 16,
                   xlab = "Time (h)", ylab = "Concentration (mg/L)",
                   main = "Three doses, four subjects")
tinyplot::tinyplot_add(conc ~ time | id, data = gr, type = "l")
abline(v = c(0, 12, 24), lty = 3, col = "grey40")
```

### Doses that depend on a parameter

`events$value` is a numeric column, so it cannot hold something being
estimated. `event_scale` is the way in: one value per observation,
constant within group, multiplying every amount in that group. A
bioavailability is then an ordinary nonlinear parameter, free to carry
covariates and random effects:

```{r, eval = FALSE}
bf(conc ~ frm_ode(pk_dyn, init = list(0, 0), times = time,
                  parms = list(exp(lka), exp(lke), exp(lV)),
                  group = id, states = c("depot", "central"),
                  output = "central",
                  events = data.frame(time = c(0, 12, 24),
                                      state = "depot", value = 100),
                  event_scale = 1 / (1 + exp(-logitF))),
   lka ~ 1 + (1 | id), lke ~ 1 + (1 | id), lV ~ 1, logitF ~ 1, nl = TRUE)
```

Because scaling is only meaningful for an amount, `event_scale` is
refused on a table containing `"replace"` or `"multiply"` rows.

### An observation at a dose time is the trough

An observation whose time equals a dose time reads the state **before**
that dose, which is the pre-dose sample a study protocol asks for. That
also covers an observation at `t0` when a dose is given at `t0`: it
reads `init`. If you want the post-dose value, ask for a time just
after.

### How the doses are applied

`frm_ode()` does not hand the table to **deSolve**'s own `events`
argument. It splits the integration at the event times instead, and
chains one solve per interval, carrying the state across the break
itself.

That is a correctness requirement. **RTMBode** integrates an augmented
system that carries the derivative of each state with respect to each
parameter alongside the states. A deSolve event jumps the state and
leaves those derivatives untouched, which is right for an addition and
wrong for the other two methods: measured against finite differences,
`"replace"` and `"multiply"` came out 42% and 59% off. Splitting the
solve is exact for all three, and it is what makes `event_scale`
possible at all.

The price is one solve per dosing interval per group. A twice-daily
regimen over a fortnight is 28 solves per subject, not one.

### Starting at steady state

A patient on maintenance therapy was not dosed for the first time this
morning. `ss = TRUE` on a row says the cycle it describes has already
been running long enough to settle, and the row's `ii` says how often:

```{r, eval = has_ode}
ss_ev <- data.frame(time = 0, state = "depot", value = 100,
                    ii = 12, ss = TRUE)
tt <- c(0, 1, 3, 6, 11.9)
frm_ode(pk_dyn, init = list(0, 0), times = tt, parms = list(1, 0.2, 10),
        states = c("depot", "central"), output = "central",
        events = ss_ev)
```

The steady state is reached by simulation, not by a formula, so it works
for any dynamics: every compartment is set to zero, which is how NONMEM
and rxode2 read a steady-state record and why the row overrides `init`,
and the cycle is then repeated `n_ss` times before the record's time.
The record's own dose lands afterwards, so an observation at the record's
time is the steady-state trough.

`n_ss` is where the honesty is. A real steady state is "repeat until the
state at the start of the cycle stops moving", and that test branches on
a value, which the tape cannot do: the number of solves has to be fixed
before the tape is built. So `n_ss` cycles is an **approximation**. Its
error is geometric: for linear kinetics the shortfall after `n` cycles is
the accumulation factor to the power `n`, which for a one-compartment
system is `exp(-n * k * ii)`.

A one-compartment intravenous model shows the size of it, because there
the exact steady-state trough is known:

```{r, eval = has_ode}
decay <- function(t, y, p) {
  "c" <- RTMB::ADoverload("c")
  list(c(-p[1] * y[1]))
}
iv_ss <- data.frame(time = 0, value = 100, ii = 12, ss = TRUE)
tr <- 100 * exp(-0.2 * 12) / (1 - exp(-0.2 * 12))    # the exact trough

vapply(c(2, 4, 8, 20), function(n) {
  g <- frm_ode(decay, init = list(0), times = 0, parms = list(0.2),
               events = iv_ss, n_ss = n, ss_tol = 1)
  abs(g - tr) / tr
}, 0)
```

Two cycles are 0.8% short, which is `exp(-2 * 0.2 * 12)` exactly; by
eight the run-in error is under the integrator's own tolerance, and the
last two columns are solver noise, not run-in error. The shortfall
matters only when the half-life is long against the dosing interval, and
there `frm_ode()` says so: off the tape it compares the last two cycles
and warns when they still differ by more than `ss_tol`.
During a fit that comparison cannot run at all, so read the warning from
a `predict()` or a direct call and raise `n_ss` if it fires.

The cost is `n_ss` extra solves per group, so a steady-state population
fit is several times a plain one.

### Resetting the system

`method = "reset"` sets **every** compartment to `value`. With
`value = 0` that is NONMEM's and rxode2's `evid = 3`: washout, a new
period of a crossover, a dropped-and-restarted infusion.

```{r, eval = has_ode}
frm_ode(pk_dyn, init = list(100, 0), times = c(1, 2, 3, 4),
        parms = list(1, 0.2, 10), output = 2L,
        events = data.frame(time = 2, value = 0, method = "reset"))
```

A reset and a dose at the same time is `evid = 4`, a reset dose. The
reset is always applied first, so the order of the two rows in the table
does not change the answer. A reset row names no compartment, so it needs
no `state` even in a multi-state system, and `event_scale` leaves it
alone: it sets a level, not an amount.

## Covariates that change with time

`init` and `parms` are read off each group's first row, so a covariate
that changes within a subject cannot enter them. That is the right
refusal for a constant, and the wrong answer for renal function, body
weight over a long study, or an infusion of a second drug that changes
the elimination of the first. `tv` is where those go.

```{r, eval = has_ode}
tt <- c(1, 3, 6, 9, 12)
ke_t <- ifelse(tt < 6, 0.2, 0.4)         # elimination doubles at hour 6

pk_tv <- function(t, y, p) {
  "c" <- RTMB::ADoverload("c")
  # `tv` values follow `parms`, so ka is p[1], V is p[2], and the
  # time-varying ke is p[3]
  list(c(-p[1] * y[1],
         p[1] * y[1] / p[2] - p[3] * y[2]))
}

frm_ode(pk_tv, init = list(100, 0), times = tt, parms = list(1, 10),
        tv = list(ke_t), states = c("depot", "central"),
        output = "central")
```

The solve is split at each change point, exactly as it is split at a dose
time, and each segment's dynamics see that segment's value. Nothing in
the derivative function learns that anything changed: within a segment
`p[3]` is an ordinary constant.

**The step is last observation carried forward**, which is the rule
rxode2 uses for covariates. A row's value is in force from that row's
time until the next change, and the first row's value reaches back to
`t0`. One consequence is worth stating, because it differs from a dose:
the state is *continuous* across a change point, since a covariate moves
the derivative and not the state. An observation exactly at a change
point therefore reads what the pre-change dynamics produced, and the
difference appears only afterwards. At a dose time the same observation
reads the trough, because a dose does jump the state.

### An estimated time-varying parameter

The values may be estimated. The change points may not: they decide
where the solve is split, and that is settled before the tape is built.
When a `tv` column carries estimated values there is nothing to compare,
because RTMB refuses comparison on AD types, so `tv_break` names the
data column whose changes mark them.

```{r, eval = has_ode}
set.seed(77)
n_id <- 10
tt_obs <- c(0.5, 2, 4, 8, 11.9, 13, 16, 20)
dv <- data.frame(id = factor(rep(seq_len(n_id), each = length(tt_obs))),
                 time = rep(tt_obs, n_id))
dv$phase <- factor(ifelse(dv$time < 12, "early", "late"))
b_i <- rnorm(n_id, 0, 0.25)[as.integer(dv$id)]
k_e <- exp(log(0.15) + b_i)
k_l <- exp(log(0.15) + log(2) + b_i)
dv$conc <- ifelse(dv$time < 12,
                  100 * exp(-k_e * dv$time),
                  100 * exp(-k_e * 12) * exp(-k_l * (dv$time - 12))) +
  rnorm(nrow(dv), 0, 0.6)

tv_fit <- frm(
  bf(conc ~ frm_ode(decay, init = list(100), times = time,
                    tv = list(exp(lk)), tv_break = phase,
                    group = id, output = 1L),
     lk ~ 1 + phase + (1 | id), nl = TRUE) + gaussian(),
  data = dv, start = list(beta = c(log(0.2), 0)))
unlist(fixef(tv_fit))
```

`lk` is a linear predictor that varies within a subject, which `parms`
would refuse by name. In `tv` it is the whole point. The truth is
`lk = -1.90` early and a shift of `log(2) = 0.69` late.

A `tv` column has no constant to be read off the first row, so `parms`
may be left out entirely when every input varies with time.

## Two measurements from one solve

A parent drug and its metabolite are measured on the same assay, one
column of concentrations with a compartment label beside it. Two species
of a predator-prey series stack the same way. Given one `output` value
per row, `frm_ode()` solves the system once per group and each row reads
the state its label names:

```{r, eval = FALSE}
bf(y ~ frm_ode(pm_dyn, init = list(dose, 0, 0), times = time,
               parms = list(exp(lka), exp(lkm), exp(lke)),
               group = id, output = cmt),      # cmt is 2 or 3 per row
   lka ~ 1, lkm ~ 1, lke ~ 1, nl = TRUE)
```

Without this the model is written as two calls, one per state, selected
by an indicator: the same likelihood at twice the solves. A per-row
`output` is told apart from a selection of states by length: it needs one
value per row, and more rows than the system has states.

## A combined error model

Assay error in pharmacokinetics is rarely purely additive. The usual
model is proportional plus additive,

$$\mathrm{sd} = \sqrt{a^2 + (b\,\mu)^2},$$

which `nlf()` writes directly, because an `nlf()` body may read another
parameter's per-row value. `sigma` is reported on the log scale, so the
body is half the log of the variance:

```{r, eval = FALSE}
bf(conc ~ frm_ode(pk_dyn, init = list(dose, 0), times = time,
                  parms = list(exp(lka), exp(lke), exp(lV)),
                  group = id, output = 2L),
   lka ~ 1 + (1 | id), lke ~ 1 + (1 | id), lV ~ 1, nl = TRUE) +
  nlf(sigma ~ 0.5 * log(exp(2 * ladd) + exp(2 * lprop) * mu^2)) +
  lf(ladd ~ 1, lprop ~ 1) + gaussian()
```

`exp(ladd)` is the additive component and `exp(lprop)` the proportional
one, both on the log scale so both stay positive. This needs no new
machinery: it is `nlf()` plus the ODE body, and the two components come
back separately in `fixef()`.

## Three things that will bite

### A covariate must not vary inside a group

`frm_ode()` reads each constant off the group's **first row**, because
one solve has one set of constants. A covariate that changes between an
early and a late observation of the same subject therefore describes a
model this helper cannot solve, and it is refused by name at frame
assembly:

```{r, eval = FALSE}
d$phase <- factor(ifelse(d$Time <= 2, "early", "late"))
frm(bf(conc ~ frm_ode(pk_dyn, init = list(Dose, 0), times = Time,
                      parms = list(exp(lka), exp(lke), exp(lV)),
                      group = Subject, output = 2L),
       lka ~ 1 + (1 | Subject),
       lke ~ 1 + phase + (1 | Subject),      # varies within Subject
       lV ~ 1, nl = TRUE) + gaussian(), data = d)
#> Error: Nonlinear parameter 'lke' is a dynamics input of frm_ode() but
#> is not constant within 'Subject': phaselate. ...
```

A covariate that is constant within a subject, such as weight or
treatment arm, is the intended case: `lV ~ 1 + Wt` is fine. A covariate
that really does change with time goes in `tv` instead, where the solve
is split at each change; the refusal is for `init` and `parms`, which
are read once.

Without the check the fit is quietly wrong rather than loud. The
coefficient stays at its starting value, the log-likelihood does not
move, and the only symptom is an indefinite Hessian.

### Only adaptive integrators

A fixed-step integrator run at the default number of steps does not
solve the system to the tolerance the likelihood is defined at, so it
returns a *different* likelihood, not a noisier one: on the probe data
`rk4` reported -63.93 where `lsoda`, `lsode`, `adams` and `ode45` all
reported -60.46 to seven digits. `frm_ode()` refuses `method = "rk4"`
and `method = "euler"`. The default `"lsoda"` is both the fastest and
the quietest of the adaptive methods.

### Keep the system small

Each group's system must be small. Above roughly eight states in one
system, the second-order derivative path that the Laplace approximation
needs starts returning `NaN` gradients, and with some integrators the
process crashes. `frm_ode()` warns past that point. This is why groups
are solved one at a time and never stacked into one large system, even
though stacking would look like the obvious optimization.

Compartment models live far below the ceiling, so in practice this only
means: do not try to fold the subjects into the state vector yourself.

## Cost

The tape is built once, but every gradient evaluation replays one
adjoint solve per group, so the cost is linear in the number of groups
and does not shrink with tape reuse. For the two-state model above,
eight timepoints per subject, a whole `frm()` call including
`sdreport()`:

| subjects | rows | with `frm_ode()` | with the closed form |
|---|---|---|---|
| 6 | 48 | 1.6 s | 0.05 s |
| 12 | 96 | 3.5 s | 0.06 s |
| 25 | 200 | 8.3 s | 0.08 s |
| 50 | 400 | 19.9 s | 0.11 s |

About 0.4 s per subject. A 200-subject population model is a couple of
minutes. If your system has a closed-form solution, write that instead:
it gives the same answer roughly a hundred times faster.

## What is out of scope

**Time-varying input that is not a step function.** A branch on time
inside the derivative function does not work, and it is worth knowing
why. **RTMBode** tapes that function once, so `t` is an
automatic-differentiation value inside it. Writing
`if (t < t_end) rate else 0` raises

```
Error: Comparison is generally unsafe for AD types
```

which is the good outcome. An `approxfun()` forcing table is the bad
one: it silently returns the value at the taping point, so the model you
fit is not the model you wrote. Smooth arithmetic in `t`, such as
`p[2] * exp(-p[3] * t)`, is fine. A piecewise-constant input belongs in
`tv`, or in `events` as an infusion, where it rides along as a parameter
over each interval and is differentiated exactly. deSolve's own
`forcings` argument is not reachable through **RTMBode**. Linear
interpolation between measured covariate values, which rxode2 offers, is
not available either: the value inside a segment would have to depend on
`t`.

**Estimated event times.** Lag times, estimated inter-dose intervals,
estimated observation times and estimated `tv` change points are not
supported. They decide where the solve is split, and that is settled
before the tape is built.

`predict(se.fit = TRUE)` is not available for any nonlinear predictor,
`frm_ode()` included. Ask for a nonlinear parameter instead, with
`predict(dpar = "lka")`.

## Failed solves

Mid-optimization the optimizer probes extreme rate constants, and a
solve can fail there. `on_error` and `penalty` say what to do about
that, but they reach less far than they look like they do, and the
difference is worth knowing before you rely on them.

**While fitting, most failures cannot be caught.** The body runs on the
automatic-differentiation tape. There `RTMBode::ode()` returns an AD
object and raises no R error when a trajectory goes bad, and RTMB
refuses comparison on AD types, so nothing can test the result for
`NaN`. A diverging region of the parameter space therefore reaches you
as the optimizer's own warning,

```
Warning: NA/NaN gradient evaluation
```

and `on_error = "error"` will not name the group. Two failures are still
caught on the tape: an integrator that gives up and returns fewer time
points than it was asked for, and an R error raised by your derivative
function.

**Everywhere else every check applies.** `predict()`, `simulate()`,
`residuals()`, a direct call to `frm_ode()`, and a fit whose `init` and
`parms` contain no estimated parameter are ordinary numeric arithmetic.
There a failed solve is detected, and by default its rows are filled
with `penalty`.

A penalty is never written silently. `frm_ode()` warns and names the
groups, and `frm_ode_failures()` reads the record back afterwards:

```{r, eval = FALSE}
p <- predict(fit, newdata = nd)
#> Warning: frm_ode(): the solve failed for 1 of 3 groups (2). Their rows
#> hold penalty = 1e+06, not a solution. ...

frm_ode_failures()
#> $groups
#> [1] "2"
```

Those rows are missing values dressed as numbers. Treat them as missing.

To localize an `NA/NaN gradient` from a fit, call `frm_ode()` yourself
at the suspect parameter values with `on_error = "error"`. Numerically
it will name the group that cannot be solved.

Solver warnings from **deSolve** ("corrector convergence failed
repeatedly", "exceeded maxsteps") during a fit are normal noise from
those probing steps. Judge the fit by `diagnose()` and the gradient at
the optimum, not by whether the solver complained on the way.
