---
title: "Differential equation models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential equation models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# the package load must sit in this unconditional chunk: the tangled
# script skips eval-gated chunks, so a library() inside one leaves
# every later unconditional chunk without frm()
library(frmtmb)
has_ode <- requireNamespace("RTMBode", quietly = TRUE)
```

Some models say what the *rate of change* of a quantity is, not what
the quantity is. A drug moves from a depot into the blood and is
cleared from it; a population grows and is eaten. `frm_ode()` puts such
a system inside a nonlinear formula, so the constants of the system are
ordinary nonlinear parameters with fixed effects, random effects and
covariates, and the model is fitted by maximum likelihood like any
other `frm()` model.

## Setup

`frm_ode()` needs the **RTMBode** package, which supplies a
differentiable interface to the **deSolve** integrators. It is not on
CRAN:

```{r, eval = FALSE}
install.packages("RTMBode", repos = c(
  "https://kaskr.r-universe.dev",
  "https://cloud.r-project.org"))
```

Everything below is skipped if it is not installed.

## A population pharmacokinetic model

The standard one-compartment model with first-order oral absorption has
two states: the amount `A` still in the depot, and the concentration `C`
in the central compartment.

$$\frac{dA}{dt} = -k_a A, \qquad
  \frac{dC}{dt} = \frac{k_a A}{V} - k_e C$$

A single dose `D` enters the depot at time zero, so $A(0) = D$ and
$C(0) = 0$. Three constants describe a subject: the absorption rate
$k_a$, the elimination rate $k_e$, and the apparent volume of
distribution $V$. All three must be positive, so they are estimated on
the log scale.

The system itself is a plain R function in the **deSolve** convention:
it takes the time, the state vector and the parameter vector, and it
returns the derivatives in a list. Index `y` and `p` by position.

```{r}
pk_dyn <- function(t, y, p) {
  # base c() drops the automatic-differentiation class; this restores it
  "c" <- RTMB::ADoverload("c")
  list(c(-p[1] * y[1],
         p[1] * y[1] / p[3] - p[2] * y[2]))
}
```

`datasets::Theoph` is the classic example: 12 subjects, one oral dose of
theophylline each, 11 serum concentrations over 25 hours.

```{r}
d <- datasets::Theoph
d$Subject <- factor(as.character(d$Subject))
head(d, 3)
```

The model is one `bf(nl = TRUE)` formula. The body says how a
concentration is produced; the three nonlinear parameter formulas say
how each constant varies between subjects.

```{r, eval = has_ode}
form <- bf(
  conc ~ frm_ode(pk_dyn,
                 init   = list(Dose, 0),
                 times  = Time,
                 parms  = list(exp(lka), exp(lke), exp(lV)),
                 group  = Subject,
                 states = c("depot", "central"),
                 output = "central"),
  lka ~ 1 + (1 | Subject),
  lke ~ 1 + (1 | Subject),
  lV  ~ 1,
  nl = TRUE
)

fit <- frm(form + gaussian(), data = d, se = TRUE,
           start = list(beta = c(0.5, log(0.08), log(0.5))))
summary(fit)
```

The typical subject's constants come back on the natural scale, and the
derived quantities a pharmacokineticist reads off them follow:

```{r, eval = has_ode}
ka <- exp(fixef(fit)$lka); ke <- exp(fixef(fit)$lke)
V <- exp(fixef(fit)$lV)
c(ka = ka, ke = ke, V = V,
  half_life = log(2) / ke,       # hours
  clearance = ke * V)            # L/h per kg of body weight
```

Everything downstream works as usual: `predict()` on new times,
`ranef()` for the subject deviations, `confint()`, `simulate()`,
`REML = TRUE`.

```{r, eval = has_ode}
nd <- data.frame(Subject = factor("1", levels = levels(d$Subject)),
                 Time = seq(0, 25, length.out = 5), Dose = 4.02)
predict(fit, newdata = nd)
```

## How the arguments work

`frm_ode()` runs **one solve per group**, over that group's own
observation times, then puts the results back in the row order of the
data.

| Argument | What it is |
|---|---|
| first argument | The derivative function, `function(t, y, parms)`. |
| `init` | Initial states, one list element per state. |
| `times` | The observation time of every row. |
| `parms` | The constants of the system, one list element per constant, in the order the derivative function indexes them. |
| `group` | The unit that owns one system: a subject, a reactor, a batch. |
| `states`, `output` | State names, and which of them the body returns. |
| `t0` | The initial time. Defaults to 0. |
| `method`, `atol`, `rtol` | Integrator and tolerances. |

`init` and `parms` are given as lists of **columns**, not as one row of
values per state. Each element is either one value per observation or a
single value shared by every group. `init = list(Dose, 0)` is therefore
"the depot starts at this row's `Dose`, the central compartment starts
at zero". A bare vector such as `c(100, 0)` is refused, because it
cannot be told apart from one column of two observations.

With no `output` the result is a matrix with one column per state, so
`frm_ode(...)[, 2]` also works. Naming the states and selecting one by
name is clearer and is what the example above does.

## Three things that will bite

### A covariate must not vary inside a group

`frm_ode()` reads each constant off the group's **first row**, because
one solve has one set of constants. A covariate that changes between an
early and a late observation of the same subject therefore describes a
model this helper cannot solve, and it is refused by name at frame
assembly:

```{r, eval = FALSE}
d$phase <- factor(ifelse(d$Time <= 2, "early", "late"))
frm(bf(conc ~ frm_ode(pk_dyn, init = list(Dose, 0), times = Time,
                      parms = list(exp(lka), exp(lke), exp(lV)),
                      group = Subject, output = 2L),
       lka ~ 1 + (1 | Subject),
       lke ~ 1 + phase + (1 | Subject),      # varies within Subject
       lV ~ 1, nl = TRUE) + gaussian(), data = d)
#> Error: Nonlinear parameter 'lke' is a dynamics input of frm_ode() but
#> is not constant within 'Subject': phaselate. ...
```

A covariate that is constant within a subject, such as weight or
treatment arm, is the intended case: `lV ~ 1 + Wt` is fine.

Without the check the fit is quietly wrong rather than loud. The
coefficient stays at its starting value, the log-likelihood does not
move, and the only symptom is an indefinite Hessian.

### Only adaptive integrators

A fixed-step integrator run at the default number of steps does not
solve the system to the tolerance the likelihood is defined at, so it
returns a *different* likelihood, not a noisier one: on the probe data
`rk4` reported -63.93 where `lsoda`, `lsode`, `adams` and `ode45` all
reported -60.46 to seven digits. `frm_ode()` refuses `method = "rk4"`
and `method = "euler"`. The default `"lsoda"` is both the fastest and
the quietest of the adaptive methods.

### Keep the system small

Each group's system must be small. Above roughly eight states in one
system, the second-order derivative path that the Laplace approximation
needs starts returning `NaN` gradients, and with some integrators the
process crashes. `frm_ode()` warns past that point. This is why groups
are solved one at a time and never stacked into one large system, even
though stacking would look like the obvious optimization.

Compartment models live far below the ceiling, so in practice this only
means: do not try to fold the subjects into the state vector yourself.

## Cost

The tape is built once, but every gradient evaluation replays one
adjoint solve per group, so the cost is linear in the number of groups
and does not shrink with tape reuse. For the two-state model above,
eight timepoints per subject, a whole `frm()` call including
`sdreport()`:

| subjects | rows | with `frm_ode()` | with the closed form |
|---|---|---|---|
| 6 | 48 | 1.6 s | 0.05 s |
| 12 | 96 | 3.5 s | 0.06 s |
| 25 | 200 | 8.3 s | 0.08 s |
| 50 | 400 | 19.9 s | 0.11 s |

About 0.4 s per subject. A 200-subject population model is a couple of
minutes. If your system has a closed-form solution, write that instead:
it gives the same answer roughly a hundred times faster.

## What is out of scope

Dosing **event tables** are not supported: repeated doses, infusions,
`evid`/`amt` records. `frm_ode()` covers models driven by their initial
conditions, which is the single-dose case. For event-driven population
pharmacokinetics use a dedicated tool such as **nlmixr2**.

`predict(se.fit = TRUE)` is not available for any nonlinear predictor,
`frm_ode()` included. Ask for a nonlinear parameter instead, with
`predict(dpar = "lka")`.

## Failed solves

Mid-optimization the optimizer probes extreme rate constants, and a
solve can fail there. `on_error` and `penalty` say what to do about
that, but they reach less far than they look like they do, and the
difference is worth knowing before you rely on them.

**While fitting, most failures cannot be caught.** The body runs on the
automatic-differentiation tape. There `RTMBode::ode()` returns an AD
object and raises no R error when a trajectory goes bad, and RTMB
refuses comparison on AD types, so nothing can test the result for
`NaN`. A diverging region of the parameter space therefore reaches you
as the optimizer's own warning,

```
Warning: NA/NaN gradient evaluation
```

and `on_error = "error"` will not name the group. Two failures are still
caught on the tape: an integrator that gives up and returns fewer time
points than it was asked for, and an R error raised by your derivative
function.

**Everywhere else every check applies.** `predict()`, `simulate()`,
`residuals()`, a direct call to `frm_ode()`, and a fit whose `init` and
`parms` contain no estimated parameter are ordinary numeric arithmetic.
There a failed solve is detected, and by default its rows are filled
with `penalty`.

A penalty is never written silently. `frm_ode()` warns and names the
groups, and `frm_ode_failures()` reads the record back afterwards:

```{r, eval = FALSE}
p <- predict(fit, newdata = nd)
#> Warning: frm_ode(): the solve failed for 1 of 3 groups (2). Their rows
#> hold penalty = 1e+06, not a solution. ...

frm_ode_failures()
#> $groups
#> [1] "2"
```

Those rows are missing values dressed as numbers. Treat them as missing.

To localize an `NA/NaN gradient` from a fit, call `frm_ode()` yourself
at the suspect parameter values with `on_error = "error"`. Numerically
it will name the group that cannot be solved.

Solver warnings from **deSolve** ("corrector convergence failed
repeatedly", "exceeded maxsteps") during a fit are normal noise from
those probing steps. Judge the fit by `diagnose()` and the gradient at
the optimum, not by whether the solver complained on the way.
