---
title: "Reinforcement learning: a sequential family from scratch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reinforcement learning: a sequential family from scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(frmtmb)
# Every code chunk below is read from one file, inst/rl/rw-delta.R, so
# the family shown here is the family tests/testthat/test-rl-example.R
# fits. Nothing on this page is a second copy.
knitr::read_chunk(system.file("rl", "rw-delta.R", package = "frmtmb",
                              mustWork = TRUE))
```

frmtmb has no reinforcement-learning family. This page writes one and
fits it hierarchically through the ordinary formula grammar. The whole
family is under 200 lines, the taped likelihood is 23 of them, and
about a third of the rest is the text of the messages it owes a user
when it refuses something.

The point is not the model. The point is the seam. A likelihood that
reads a whole sequence rather than one row at a time reaches the core
through `frmtmb_structure()`, and everything else keeps working: random
effects, links, standard errors, prediction.

Read `vignette("frmtmb")` first for the grammar, and
`?frmtmb_structure` for the reference description of the protocol used
here.

## The model

The task is a two-armed bandit. On each trial a subject picks one of
two arms and receives a payoff. The Rescorla-Wagner delta rule says the
subject keeps a value estimate `Q` for each arm, starts both at zero,
and after each choice moves the chosen arm's estimate toward what it
just paid:

```
Q[chosen] <- Q[chosen] + alpha * (reward - Q[chosen])
```

The choice itself is a softmax over the two values, which for two arms
is a logistic function of their difference:

```
P(arm 1) <- plogis(beta * (Q1 - Q2))
```

`alpha` is the learning rate, on `(0, 1)`. `beta` is the inverse
temperature, on `(0, Inf)`: a large `beta` means the subject almost
always takes the arm it currently values more. This is the model
hBayesDM calls `bandit2arm_delta` (Ahn, Haines and Zhang 2017,
*Computational Psychiatry* 1, 24). None of hBayesDM's code is used
here; the equations above are the published model, and the family below
is written from them.

Each of the two parameters becomes a distributional parameter with its
own linear predictor, so a learning rate can carry a condition effect
and a subject-level random intercept exactly the way a mean would.

## Why this needs the structured protocol

The likelihood does factorize over trials: given the parameters, `Q` is
a deterministic function of the subject's own earlier choices and
payoffs, so each trial contributes one Bernoulli factor.

What it is not is ROWWISE. A row's factor cannot be computed from that
row's parameters and response alone, because `Q` carries the whole
history. `frmtmb_family(lpdf =)` is handed one row's worth of
everything, so it cannot express this. `frmtmb_structure(loglik =)` is
handed the whole response and returns one number, so it can.

## The family

### The data a trial carries

```{r rl-aterm}
```

### The block: from a long data frame to subject by trial

The recursion is sequential in trials and independent across subjects,
so it should walk trials once and update every subject at each step.
That needs the rows arranged as a subject-by-trial matrix, which is
what the `frame_block` slot is for. It runs once, at frame assembly,
and everything it returns is plain data that gets saved inside the fit.

```{r rl-block}
```

Two things to notice. `idx` holds row NUMBERS, not values, so the
recursion can index any per-row quantity with it: the response, the
payoffs, and the linear predictors of `alpha` and `beta`. And subjects
with different trial counts are padded to the longest one, with `mask`
marking the padding. The mask is what makes the padding free of
branches on the tape: it multiplies the log-likelihood, so a padded
cell contributes nothing, and it multiplies the learning rate, so a
padded cell learns nothing.

Padding is not the only way. `frmtmb.latent::hmm()` sorts its sequences
by decreasing length instead, which makes the sequences still running
at step `s` a prefix of the order and needs no padded cells at all.
That is faster and harder to read. Measured on 40 subjects whose trial
counts run from 30 to 100, the padded design carries 41 percent empty
cells and builds its tape in 0.09 seconds against the balanced design's
0.08, so for a tutorial the mask wins.

### The likelihood

```{r rl-loglik}
```

The loop body is the delta rule, written for every subject at once.
`q1` and `q2` are vectors with one entry per subject, and one iteration
of the loop advances all of them. There is no conditional anywhere: the
chosen arm is selected by multiplying by the response, which is data,
and the padding is selected by multiplying by the mask, which is also
data.

`rw_terms()` stops one step short of a log-likelihood: it returns the
per-trial factors themselves, and the three slots below are three ways
of adding them up. A subject's log-likelihood is the elementwise sum
across trials, a trial's is the factor scattered back to its own row,
and the total is the sum of either. Writing them as one recursion with
three tails is not tidiness, it is the only way to be sure the three
numbers agree; the suite asserts that they do.

Why three at all. `loglik` is what the objective needs.
`loglik_group` is what `frm(importance = )` resamples, one value per
subject. `loglik_row` is what a deviance residual compares against its
saturated fit, which for a Bernoulli factor is zero, and that zero is
attached to the row values so the core does not have to guess it.

The `nrep` line is the stacking contract. The importance correction
evaluates the whole design once per draw, stacked end to end, so the
recursion has to run over subject-crossed-with-draw. Because the loop
is already vectorized across subjects, that is the same loop over a
longer vector rather than a new algorithm.

`weights` arrives and is ignored, which the family is allowed to do
only because it refuses `weights()` outright further down. A row weight
has no meaning here. A trial's factor depends on every earlier trial of
the same subject, so there is nothing separable to reweight.

### Fitted values and draws

The same recursion in plain numbers answers two more questions.

```{r rl-replay}
```

The per-trial choice probability is the natural fitted value for this
family, and its binomial variance is what pearson residuals divide by.
A forward draw is the same walk with `rbinom()` in place of the
observed choice, which is why one function serves both.

```{r rl-sim}
```

### The family object

```{r rl-family}
```

```{r rl-structure}
```

The refusals are the interesting part. A structured family starts with
every capability turned off, and each one it leaves off should say why
in its own words. `conditional_effects()` is refused not because it is
hard but because it is undefined: the expected response here is a
choice probability that depends on a whole trial history, and the
synthetic grid that function builds has no history.

`deviance` is the one flag turned back on, and it is on because of a
slot rather than in spite of one: with `loglik_row` declared the core
has each trial's log-density and its saturated value, which is what a
unit deviance is made of. `unit` is a separate declaration and keeps
its own answer. The finest factorization here is the TRIAL; the
smallest thing that can honestly be left out is a whole SUBJECT,
because dropping one trial changes every later trial's value store.

## Fitting it

```{r rl-design}
```

```{r sim-data}
set.seed(2024)
design <- rl_bandit_design(n_subj = 30, n_trial = 80)
dd <- rl_simulate(design, nsim = 1, seed = 2024)[[1]]
str(dd[1:3, ])
```

`rl_simulate()` goes through `frm_simulate()`, which draws a fresh set
of subject effects from `rl_truth` and then calls the family's own
simulator. The choices in `dd` therefore come from the same code path
the fit is about to score.

Now the fit. This is an ordinary `frm()` call:

```{r fit}
fit <- frm(bf(choice | reward(pay1, pay2) ~ condition + (1 | p | id),
              beta ~ 1 + (1 | p | id)),
           family = rw_delta(subject = id, trial = trial),
           data = dd)
fixef(fit)
```

The main right-hand side is the LEARNING RATE's predictor, because the
family declared `primary_dpars = "alpha"`. The `| p |` in both terms
correlates each subject's learning rate with its own temperature, which
is the usual hierarchical parameterization for these models.

```{r varcorr}
VarCorr(fit)
```

Compare with the simulated values:

```{r truth}
rbind(estimate = c(unlist(fixef(fit)),
                   sd_alpha = sqrt(VarCorr(fit)[[1]][1, 1]),
                   sd_beta = sqrt(VarCorr(fit)[[1]][2, 2])),
      truth = c(rl_truth$alpha_Intercept, rl_truth$alpha_conditiontrt,
                rl_truth$beta_Intercept,
                rl_truth$sd_id__choice.alphaIntercept,
                rl_truth$sd_id__choice.betaIntercept))
```

This is one dataset, so read the differences as sampling error and not
as bias. The recovery study below is the systematic version.

Everything downstream of the fit works because the structure said it
does. The fitted values are per-trial choice probabilities:

```{r fitted}
p <- fitted(fit)
summary(p)
# the model's mean predicted choice rate against the observed one
c(predicted = mean(p), observed = mean(dd$choice))
```

## Check 1: an independent reference for the recursion

The taped recursion is vectorized and masked, which is exactly the kind
of code that can be wrong in a way that still converges. So write the
model out again, the way it is stated, one subject and one trial at a
time, over quantities pulled back through the public accessors, and
compare.

```{r reference}
u <- ranef(fit)[[1]]
sid <- match(as.character(dd$id), rownames(u))
xa <- model.matrix(~ condition, dd)
alpha <- plogis(as.vector(xa %*% unname(fixef(fit)$alpha)) + u[sid, 1])
beta <- exp(unname(fixef(fit)$beta)[[1]] + u[sid, 2])

ll <- 0
for (s in unique(sid)) {
  rows <- which(sid == s)
  rows <- rows[order(dd$trial[rows])]
  q <- c(0, 0)
  for (i in rows) {
    ll <- ll + dbinom(dd$choice[i], 1, plogis(beta[i] * (q[1] - q[2])),
                      log = TRUE)
    k <- if (dd$choice[i] == 1) 1 else 2
    pay <- if (k == 1) dd$pay1[i] else dd$pay2[i]
    q[k] <- q[k] + alpha[i] * (pay - q[k])
  }
}
c(reference = ll,
  from_fitted = sum(dbinom(dd$choice, 1, p, log = TRUE)),
  difference = ll - sum(dbinom(dd$choice, 1, p, log = TRUE)))
```

`tests/testthat/test-rl-example.R` runs this check on an unbalanced
design as well, where the padding is live, and against the full joint
density rather than the data part alone.

## Check 2: an identity against Stan

The reference above shares R with the thing it checks. A second
implementation in another language and another autodiff system does
not. `tests/testthat/helper-rl.R` holds a thirty-line Stan program for
the same model, written from the equations at the top of this page, and
compares Stan's `log_prob` with frmtmb's joint log density at frmtmb's
own estimates. The Stan program declares exactly the parameters frmtmb
estimates, on the scales frmtmb estimates them on, so the map between
them is the identity and the comparison carries no Jacobian.

Measured on 30 subjects and 100 trials:

| quantity | at the estimates | at a displaced point |
|---|---|---|
| Stan `log_prob` | -1427.8453940 | -1460.3042080 |
| frmtmb joint log density | -1427.8453940 | -1460.3042080 |
| difference | -4.5e-13 | -2.5e-12 |
| largest gradient on the subject effects | 1.2e-14 | not asserted |
| largest gradient overall | 5.2 | not asserted |

The difference is zero and is expected to be exactly zero, not merely
small: core `frm()` carries no default prior, the Stan program uses
`target +=` with `_lpmf` and `_lpdf` so it drops no normalizing
constant, and every declared parameter is unbounded so
`adjust_transform = FALSE` has no Jacobian to discard. The second
column is a deliberately non-stationary point, so the agreement cannot
be an artifact of both sides sitting at an optimum.

The gradient row is the check that catches a wrong map. frmtmb puts
the subject effects at their conditional modes, so Stan's gradient with
respect to them must vanish there. The gradient with respect to the
fixed effects is NOT zero at that point and is not asserted on: the
maximum-likelihood estimate makes the MARGINAL likelihood stationary,
not the joint one.

The test is gated the same way the brms comparison tier is, on
`FRMTMB_BRMS_FIT_TESTS` and `NOT_CRAN`, because it compiles a Stan
program.

## Check 3: parameter recovery

Simulate 100 datasets at known parameter values through the family's
own simulator, fit each one, and ask whether the estimates land on the
truth and whether their intervals cover it. The design is 40 subjects
with 100 trials each and a condition effect on the learning rate. The
script is described in `dev/rl-findings.md`.

| parameter | truth | bias | Monte Carlo se | sd of estimates | coverage |
|---|---|---|---|---|---|
| `alpha_(Intercept)` | -0.619 | +0.009 | 0.018 | 0.178 | 0.94 |
| `alpha_conditiontrt` | 0.800 | -0.018 | 0.025 | 0.248 | 0.94 |
| `beta_(Intercept)` | 1.099 | +0.002 | 0.007 | 0.070 | 0.96 |
| `log sd(alpha)` | -0.693 | -0.211 | 0.057 | 0.573 | 0.93 |
| `log sd(beta)` | -0.916 | -0.062 | 0.017 | 0.167 | 0.95 |

The three fixed effects are unbiased to within their Monte Carlo error
and their 95 percent Wald intervals cover at the nominal rate. The
variance components are the weak point, and the learning-rate one
especially: its estimate is biased downward and scattered widely, which
is the usual behavior of a variance component estimated from binary
data with a small number of levels. Coverage is over the replicates
that produced a finite interval, which for `log sd(alpha)` was 98 of
the 100.

## The Laplace caveat, measured

frmtmb integrates the subject effects out with a Laplace approximation.
The approximation is a Gaussian fitted at the conditional mode, and it
is exact only when the conditional log-density really is quadratic. For
a binary response it is not, and the fewer trials a subject has, the
less quadratic it is.

The way to price that error in frmtmb is `frm(importance = )`, which
reweights draws from the Laplace Gaussian against the exact
integrand. It works here, and it works because the family declares
how its likelihood factorizes: `loglik_group` hands the correction one
value per subject, which is exactly the piece it resamples. Until that
slot existed the correction refused every family with a `loglik`,
whether or not the likelihood factorized.

```{r importance, error = TRUE}
fit_i <- frm(bf(choice | reward(pay1, pay2) ~ condition + (1 | p | id),
                beta ~ 1 + (1 | p | id)),
             family = rw_delta(subject = id, trial = trial),
             data = dd, importance = 200)
c(laplace = sqrt(VarCorr(fit)[[1]][1, 1]),
  corrected = sqrt(VarCorr(fit_i)[[1]][1, 1]))
fit_i$importance[c("draws", "mcse")]
```

The chunk is left with `error = TRUE` on purpose. The correction can
decline, and on this model it sometimes does; the next section is
about when.

## Short sessions, and what the correction says about them

Run the same recovery study at 20 trials per subject instead of 100,
and read the bias and the coverage:

| parameter | 100 trials, bias | coverage | 20 trials, bias | coverage |
|---|---|---|---|---|
| `alpha_(Intercept)` | +0.009 | 0.94 | +0.064 | 0.95 |
| `alpha_conditiontrt` | -0.018 | 0.94 | -0.005 | 0.95 |
| `beta_(Intercept)` | +0.002 | 0.96 | -0.000 | 0.92 |
| `log sd(alpha)` | -0.211 | 0.93 | -0.628 | 0.73 |
| `log sd(beta)` | -0.062 | 0.95 | -0.247 | 0.89 |

100 replicates each, 40 subjects, the same true parameter values.
Coverage is over the replicates that produced a finite interval: at 20
trials that was 84 of 100 for `log sd(alpha)` and 96 or more for
everything else.

Two readings. The fixed effects survive. At 20 trials per subject their
bias is still inside Monte Carlo error and their intervals still cover
close to the nominal rate, so a study that wants the condition effect
on the learning rate can work with short sessions.

The variance components do not survive. The learning rate's log
standard deviation comes out 0.63 too low, which is a standard
deviation about half the true one, and its interval covers 73 times in
100. Sixteen of the hundred short-session fits gave no usable interval
for it at all.

Two causes could produce that, and they are not the same thing. A
variance component estimated by maximum likelihood from binary data
with 40 levels is biased downward whether or not the integral is
approximated; and the Laplace approximation is itself worse the fewer
trials a subject has. The recovery table cannot separate them.
`frm(importance = )` can, and now that this family declares its
pieces, it can be asked. Six seeds per design, the same 40 subjects
and the same truth, 100 draws:

| | 40 x 100 trials | 40 x 20 trials |
|---|---|---|
| replicates the correction completed | 4 of 6 | 5 of 6 |
| mean shift, `alpha_(Intercept)` | -0.004 | +0.034 |
| mean shift, `alpha_conditiontrt` | +0.010 | +0.014 |
| mean shift, `beta_(Intercept)` | -0.002 | +0.048 |
| mean shift, `log sd(alpha)` | +0.357 | +0.480 |
| largest single shift, `log sd(alpha)` | 0.545 | 1.386 |
| Monte Carlo standard error | 0.13 to 0.18 | 0.02 to 0.70 |
| smallest effective sample size per draw | 0.53 | 0.03 |

The fixed effects do not move. That is the recovery table's first
reading confirmed by an independent route: their bias is not Laplace
error, because removing the Laplace error leaves them where they were.

The variance component moves up, a long way, and the diagnostics say
how far to trust it. At 20 trials the correction is often not
answering at all: one replicate in six does not complete, the largest
shift is 1.39 on the log scale, and the effective sample size falls to
three draws in a hundred.

One dataset, named so it can be checked, shows what that looks like.
Take 40 subjects at 20 trials from
`set.seed(4206); rl_simulate(rl_bandit_design(40, 20), seed = 4206)`.
Its Laplace `log sd(alpha)` is -1.2024. At 100 draws the correction
returns +0.1831, a standard deviation of 1.20 against a truth of 0.5,
with a Monte Carlo standard error of 0.336 and a smallest effective
sample size of 0.33 per draw. At 400 draws, with enough draws to see
its own weights, it REFUSES, reporting that the corrected objective
rose rather than fell and pointing at the `[ID]` block's two variance
components and their correlation.

That is one dataset and not a law. Other 20-trial datasets are fine:
a reviewer running two seeds of this design got one that completed at
400 draws and one whose 100-draw correction was already clean
(shift 0.025, mcse 0.070, smallest effective sample size 0.90). The
honest summary is the table above, where one replicate in six failed
and the spread of shifts was an order of magnitude: at this design the
correction is unreliable rather than uniformly broken, and a number it
returns needs its own diagnostics read before it is believed.

So the honest reading is not that the variance-component bias is
Laplace error. It is that at 100 trials the correction runs and moves
the variance component up by about a third of a log unit, the same
order as the whole bias, so some real part of that bias is the
approximation; and that at 20 trials the correction will not stand
behind an answer at all, which is a second reason to treat a
subject-level standard deviation from 20 binary trials as a lower
bound rather than as an estimate. The operational advice is unchanged:
report fixed effects from short sessions, and do not report a
variance component from them.

## Shape and cost

The family loops over TRIALS and vectorizes over subjects. It is worth
seeing what the other spelling costs. Here is the same likelihood
written the way the model is usually stated, one row at a time, with
the value store updated by sub-assignment:

```{r slow-shapes, eval = FALSE}
for (t in seq_len(ncol(idx))) {
  for (s in seq_len(nrow(idx))) {          # the extra loop
    i <- idx[s, t]
    eta <- beta[i] * (q1[s] - q2[s])
    ll <- ll + y[i] * eta - RTMB::logspace_add(0 * eta, eta)
    if (y[i] == 1) q1[s] <- q1[s] + alpha[i] * (pay1[i] - q1[s])
    else q2[s] <- q2[s] + alpha[i] * (pay2[i] - q2[s])
  }
}
```

All three spellings give the same objective value to fourteen
significant digits. On 40 subjects and 100 trials, 4000 rows, with the
three builds interleaved and each gradient timed over a batch of 100
calls:

| `loglik` shape | tape build | one gradient |
|---|---|---|
| vectorized over subjects | 0.08 s | 7.3 ms |
| row loop, scalar value store | 0.57 s | 7.2 ms |
| row loop, length-n log-density vector | 0.65 s | 7.2 ms |

The whole penalty is paid at TAPE CONSTRUCTION, and none of it per
evaluation. That is what it has to be: once the tape exists it is a
node list, and the R code that built it is gone. The gradient column is
flat to within measurement noise, and an optimization runs hundreds of
gradients against one build.

Do not read the build column as a fixed multiplier either. RTMB's `[<-`
copies the vector it writes into, so the sub-assigning form gets
relatively worse as the response grows:

| rows | vectorized | length-n sub-assignment | factor |
|---|---|---|---|
| 1000 | 0.03 s | 0.16 s | 5.3 |
| 4000 | 0.08 s | 0.60 s | 7.5 |
| 10000 | 0.14 s | 1.67 s | 11.9 |
| 20000 | 0.34 s | 4.33 s | 12.7 |

Values agree to 1e-11 across the whole sweep. The rule to take away is
the shape, not a number: vectorize the tape-building loop, and expect
the cost of not doing so to grow with the data.

Scaling of the vectorized family, interleaved, `frm(dry_run =
"objective")` with the frame-assembly time subtracted, median of 7
builds and the best of 3 gradient batches:

| subjects | trials | rows | tape build | one gradient |
|---|---|---|---|---|
| 10 | 100 | 1000 | 0.04 s | 10.0 ms |
| 40 | 25 | 1000 | 0.03 s | 8.8 ms |
| 40 | 100 | 4000 | 0.08 s | 39.6 ms |
| 40 | 400 | 16000 | 0.28 s | 153.4 ms |
| 160 | 100 | 16000 | 0.22 s | 160.6 ms |
| 40 | 800 | 32000 | 0.61 s | 296.2 ms |

Both costs track the ROW count, which is the node count. Read the two
16000-row rows against each other for the part that does not: more
trials at a fixed row count costs more to tape, because the R loop runs
more iterations, and costs the same per gradient, because the tape is
the same size either way. The size of that build gap is not stable
enough to quote. Two interleaved runs of this table on one machine put
it at 1.27x and 1.70x.

## What a built-in family would add

The example above is a complete family. Three things would still change
if it moved into a package.

**A pointwise log-likelihood for `loo()`.** The family declares its
factorization, so `frm(importance = )` and deviance residuals work
above. `loo()` and `waic()` would work too, on a sampled fit: they
live in frmtmb.sample and read the same slots, and their columns would
be SUBJECTS rather than trials, because `unit` says a trial cannot
honestly be left out. Nothing here exercises that path, because these
are maximum-likelihood fits and an elpd is a posterior quantity.

**Newdata.** `predict(type = "response")` on new data would need a
block built without a response, and the protocol defers that case on
purpose. For this family the meaning is clear enough (replay the new
subject's payoff schedule against the fitted parameters), so a built-in
version would be worth the work.

**More of the literature.** A real package would carry the family of
models this one is the simplest member of: separate learning rates for
gains and losses, a decay term, more than two arms. All of them share
this block and this loop shape, and differ only in the body.

None of that is a limitation of the protocol as a place to put a
sequential likelihood. The core knows nothing about reinforcement
learning, and the model above fits, predicts, simulates and recovers
its parameters through the ordinary grammar.
