---
title: "Drift-diffusion models of two-choice response times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drift-diffusion models of two-choice response times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_rw <- requireNamespace("RWiener", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  # every fit and every reference value needs the simulator or the
  # reference density, so the whole vignette is gated on one flag
  eval = has_rw
)
library(frmtmb.ddm)
```

```{r, echo = FALSE, eval = !has_rw, comment = NA}
cat("RWiener is not installed, so this vignette was built without\n",
    "running its examples. Install RWiener to reproduce them.\n")
```

## What the model says

A participant sees a letter string and decides whether it is a word.
The drift-diffusion model says that evidence accumulates noisily from a
starting point until it reaches one of two boundaries, and the response
time is when it gets there. Four quantities describe that:

- `mu`, the **drift rate**: how fast evidence arrives, and in which
  direction. This is the one that usually carries the experimental
  effect.
- `bs`, the **boundary separation**: how much evidence is required.
  Wide boundaries mean slow, accurate responses.
- `ndt`, the **non-decision time**: encoding the stimulus and moving the
  finger, which no amount of deciding accounts for.
- `bias`, the **relative start point**: where accumulation begins
  between the boundaries. 0.5 is unbiased.

The names are brms's, so a model written for `brms::wiener()` reads the
same here.

## Simulating an experiment

`ddm_simulate()` draws response times and boundary choices jointly, the
way an experiment produces them. Here, words and non-words differ in
drift rate and in nothing else, which is the standard claim about a
lexical decision task.

```{r simulate}
set.seed(20240903)
n <- 1200
word <- rep(c(0, 1), each = n / 2)
dat <- ddm_simulate(n,
                    mu = 0.3 + 1.0 * word,   # words drift faster
                    bs = 1.5,
                    ndt = 0.30,
                    bias = 0.5)
dat$lex <- factor(word, labels = c("nonword", "word"))
head(dat)
```

`upper` is the decision: 1 for a response at the upper boundary, 0 for
the lower one.

```{r describe}
with(dat, tapply(rt, list(lex, upper), mean))
with(dat, tapply(rt, lex, length))
prop.table(table(dat$lex, dat$upper), margin = 1)
```

## The decision indicator

The boundary a trial ended at is data, and it reaches the density as an
addition term spelled the way brms spells it:

```{r spelling, eval = FALSE}
# brms
brm(rt | dec(decision) ~ lex, family = wiener(), data = dat)

# here
frm(bf(rt | dec(decision) ~ lex), family = wiener(), data = dat)
```

`dec()` takes what brms takes: a factor or a character vector, read on
its levels with the second one as the upper boundary, or a 0/1 column.
The package contributes the term to frmtmb's addition-term registry when
it loads, so nothing has to be enabled.

`vint()` carries the same thing as a plain 0/1 integer and also works.
It was the only route before frmtmb had a registry to contribute to, and
models written that way keep running:

```{r spelling2, eval = FALSE}
frm(bf(rt | vint(upper) ~ lex), family = wiener(), data = dat)
```

Supplying neither is refused. An absent addition term reaches a density
as `NULL`, `NULL` in arithmetic is a zero-length result, and the fit
would otherwise return with a log likelihood summed over no rows.

## Fitting

Drift depends on lexicality; everything else is constant across trials.
`bias = 0.5` fixes the start point on the response scale, which is the
usual choice when nothing in the design would bias it.

```{r fit}
fit <- frm(bf(rt | vint(upper) ~ lex, bias = 0.5),
           family = wiener(), data = dat)
summary(fit)
```

Read the coefficients on their own scales. Drift is on the identity
scale, so the `lexword` coefficient is the change in drift rate
directly.
Boundary separation is on the log scale. The non-decision time is on a
logit scaled onto `(0, min(rt))`, which is what keeps it inside the
support:

```{r coefs}
e <- unlist(fixef(fit))
c(drift_nonword = e[["mu.(Intercept)"]],
  drift_word_gain = e[["mu.lexword"]],
  boundary = exp(e[["bs.(Intercept)"]]),
  ndt = min(dat$rt) / (1 + exp(-e[["ndt.(Intercept)"]])))
```

Against the values the data were generated from -- drift 0.3 and 1.3,
boundary 1.5, non-decision time 0.30 -- that is the recovery this design
supports at `r n` trials.

## Why the non-decision time needs a bounded link

The density is exactly zero for a response time at or below `ndt`. A log
link would let the optimizer propose a non-decision time above the
fastest observed response, where the likelihood is not merely small but
undefined, and it would find out only by stepping there.

So `ndt` gets a logit scaled onto `(0, max_ndt)`, with `max_ndt`
defaulting to the smallest response time in the data. The constraint
becomes structural: no finite linear predictor can violate it.

```{r ndt-link}
lk <- wiener(max_ndt = 0.4)$links[["ndt"]]
lk$linkinv(c(-10, -2, 0, 2, 10))   # always strictly inside (0, 0.4)
```

Pass `max_ndt` explicitly when you intend to `predict()` on new data
whose fastest response differs from the training set's.

## Predictions

`fitted()` and `predict(type = "response")` give the mean response time,
conditional on the boundary that row ended at, in closed form.

```{r predict}
nd <- data.frame(lex = factor(c("nonword", "word")), upper = c(1, 1))
predict(fit, newdata = nd, type = "response")
```

`vint()` is required on `newdata` as well; there is no default boundary,
because the conditional mean genuinely depends on it.

One caution that looks like a bug and is not. With `bias` fixed at 0.5,
the conditional mean decision time is the **same** at both boundaries,
for any drift rate. That is a theorem about the diffusion, not an
oversight: an unbiased start gives fast and slow responses the same
average time whichever way they go. Free the bias and the two separate:

```{r bias-free}
set.seed(7)
d2 <- ddm_simulate(900, mu = 0.7, bs = 1.4, ndt = 0.25, bias = 0.35)
f2 <- frm(bf(rt | vint(upper) ~ 1), family = wiener(), data = d2)
c(pred_upper = predict(f2, newdata = data.frame(upper = 1),
                       type = "response"),
  pred_lower = predict(f2, newdata = data.frame(upper = 0),
                       type = "response"),
  obs_upper = mean(d2$rt[d2$upper == 1]),
  obs_lower = mean(d2$rt[d2$upper == 0]))
```

## Simulating from a fit

`simulate()` draws response times at each row's own boundary, by
inverse transform through the first-passage quantile function. Comparing
those draws with the observed times is the usual posterior-predictive
sanity check.

```{r simulate-fit}
set.seed(3)
sims <- as.matrix(simulate(fit, nsim = 100))
c(observed = mean(dat$rt), simulated = mean(sims))
quantile(dat$rt, c(0.1, 0.5, 0.9))
quantile(sims,   c(0.1, 0.5, 0.9))
```

## What this family does not do

- **`cens()` and `trunc()` are refused.** They need a distribution
  function, and the family declares none. The Wiener first-passage CDF
  is a third series with its own truncation problem, and it is not
  written here.
- **`residuals(type = "pearson")` and `"deviance"` are refused.** No
  variance function and no unit deviance. `type = "response"` works.
- **The decision is data, not a modelled outcome.** This family models
  the response time given the boundary. It does not jointly model which
  boundary was reached, so it is not a substitute for fitting choice and
  time together.
- **The start-point range is not held inside the boundaries by
  construction.** `wiener(variability = "sz")` gives `sz` a logit link,
  so the width is below 1 and the uniform start point stays inside the
  boundaries whenever `bias` is 0.5, which is the usual case. At a
  strongly biased start a wide `sz` can push the range past a boundary.
  Nor is `ndt - st / 2` held above zero. Both are joint constraints on
  two distributional parameters, and a link is a property of one, so
  neither can be made structural from outside frmtmb.

`frm_compat("wiener")` states each of these, along with what was tested
and what merely was not.

```{r compat}
frm_compat("wiener")[, c("feature_b", "status")]
```

## The full model: across-trial variability

Everything above holds the four parameters fixed within a condition.
Ratcliff's full diffusion model does not: it draws the drift rate, the
start point and the non-decision time afresh on every trial, which is
what lets one model account for both the fast errors a start-point range
produces and the slow errors a drift range produces.

`variability` turns those on. Each one it names becomes an ordinary
distributional parameter and takes its own formula like any other:

```{r fullddm}
set.seed(4)
full <- ddm_simulate(600, mu = 1.2, bs = 1.5, ndt = 0.30,
                     sv = 1.0, st = 0.10)
ffit <- frm(bf(rt | dec(upper) ~ 1, bias = 0.5),
            family = wiener(variability = c("sv", "st")), data = full)
e <- unlist(fixef(ffit))
ub <- min(full$rt)
c(mu  = e[["mu.(Intercept)"]],
  bs  = exp(e[["bs.(Intercept)"]]),
  ndt = ub / (1 + exp(-e[["ndt.(Intercept)"]])),
  sv  = exp(e[["sv.(Intercept)"]]),
  st  = 2 * ub / (1 + exp(-e[["st.(Intercept)"]])))
```

### Three integrals, three answers

The likelihood is the analytic Wiener density averaged over those
distributions. They are not averaged the same way, because they are not
the same integral.

**The drift rate has a closed form.** The drift enters the log density
only as `-v a w - v^2 t / 2`: the series that carries the shape of the
distribution is the driftless one, and the drift multiplies it by an
exponential-quadratic. Averaging an exponential-quadratic against a
normal completes the square, so

```
mean of exp(-v a w - v^2 t / 2)
  = (1 + t sv^2)^(-1/2)
    exp( ((a w)^2 sv^2 - 2 v a w - v^2 t) / (2 (1 + t sv^2)) )
```

That is exact. There is no node count for `sv` and no accuracy to
trade, and estimating it is free relative to the plain density: over
three seeds a fit with `sv` took 0.97 to 1.07 times a fit without it,
and at the density level both cost 0.0002 s per call. The two
quadratures below are the ones that cost.

**The other two are uniform, and get fixed Gauss-Legendre nodes.** The
node positions and counts are decided when the family object is built.
Nothing may branch on a parameter on an automatic-differentiation tape,
and a node count that moved with `st` would be exactly that; a parameter
only rescales the interval the fixed nodes are mapped onto.

The two defaults differ because the two integrands do. The start-point
integrand is analytic across its whole range. The non-decision-time
integrand is not free to use its whole range: the density is zero for a
non-decision time at or past the response time, so on a fast trial the
range is cut and the integrand turns on sharply at the cut. Measured
against a far finer rule on such a trial:

```{r nodes}
lp <- function(nsz, nst) {
  nd <- frmtmb.ddm:::ddm_nodes(c("sz", "st"), c(sz = nsz, st = nst))
  frmtmb.ddm:::ddm_lpdf_var(0.36, 1.0, 1.5, 0.45, 0.30, 1.0, 0.2, 0.2,
                            0, nd, TRUE, 1e-9)
}
best <- lp(31L, 81L)
n <- c(3, 5, 7, 11, 15, 21, 31)
data.frame(nodes = n,
           sz_err = abs(vapply(n, function(k) lp(k, 81L), 0) - best),
           st_err = abs(vapply(n, function(k) lp(31L, k), 0) - best))
```

The start-point column is at machine precision by 7 nodes. The
non-decision-time column takes 21 to reach 1e-9, which is where the
default sits. Lower it with `wiener(nodes = c(st = 11))` if the fit is
slow and the extra digits are not worth the time.

### A contaminant mixture

Fast guesses are not diffusion trials, and a model with no component for
them pays by dragging the non-decision time down. The standard treatment
is a mixture, which needs the Wiener component's non-decision time to
sit above the fastest response time. That is refused by default, for a
bare model correctly, and `allow_unreachable = TRUE` says the other
component covers those trials:

```{r mixture, eval = FALSE}
frm(bf(rt | dec(decision) ~ 1, bias1 = 0.5),
    family = mixture(wiener(max_ndt = 0.4, allow_unreachable = TRUE),
                     lognormal()),
    data = dat)
```

A trial below the non-decision time then gets a log density that
exponentiates to exactly zero, which is the right likelihood for that
component, and differentiates to exactly zero, which a true `-Inf` would
not: `-Inf` gives the correct value and a `NaN` gradient, and one `NaN`
stops the fit.

### Why this is not `quadrature = TRUE`

frmtmb has a `quadrature` argument, and it is not this. It marginalizes
RANDOM EFFECTS by adaptive Gauss-Kronrod, is wired to the random-effect
coefficient vector by name, and refuses a model that has no
random-effect block at all. Across-trial variability shares nothing
between trials, has no level to estimate, and gives every row its own
integral whether or not the model has a grouping factor. It belongs
inside the density, and that is where it is.

## The density

The likelihood is the Wiener first-passage time density of Navarro and
Fuss (2009). They give two series for it: one that converges quickly at
small normalized times `u = t / a^2`, one that converges quickly at
large ones, and a criterion for choosing between them per evaluation.

An automatic-differentiation tape cannot make that choice. `u` depends
on the boundary separation, which is a parameter, so a comparison on `u`
is a branch on a parameter -- and RTMB advectors carry no comparison
operators, so it does not even tape. Branching on the response alone
would tape and would be wrong, because the same response time falls on
either side of the criterion depending on the boundary separation.

The two BOUNDARIES are the case where that rule works in your favor.
Only the lower-boundary density is written, because reflecting the
process maps an upper crossing onto a lower one with `bias -> 1 - bias`
and `mu -> -mu`, and the reflection is driven by the `vint()` column,
which is data. Arithmetic on a data column tapes.

This package evaluates **both** series at a fixed truncation and blends
their logs with a logistic weight in `log(u)`: Navarro and Fuss's
criterion with the step smoothed. The weight saturates to exactly 0 and
exactly 1 outside a narrow band, and inside that band both series are
accurate, so any convex combination of them is accurate too.

The cost is that every evaluation pays for both series. What it buys is
a density that holds over the whole range of normalized times rather
than the part a single truncation happens to cover:

```{r accuracy}
u <- c(0.005, 0.05, 0.4, 2, 8, 30)
ref <- vapply(u, function(uu) {
  log(RWiener::dwiener(uu + 1e-12, 1, 1e-12, 0.45, 1, resp = "lower"))
}, 0)
mine <- vapply(u, function(uu) {
  frmtmb.ddm:::ddm_lpdf_lower(uu, 1, 1, 0.45)
}, 0)
small_only <- vapply(u, function(uu) {
  frmtmb.ddm:::ddm_log_gs(uu, 0.45) - 1 * 1 * 0.45 - 1 * 1 * uu / 2
}, 0)
data.frame(u,
           blended_rel_err = abs(mine - ref) / abs(ref),
           small_time_only_rel_err = abs(small_only - ref) / abs(ref))
```

The last column is what a fixed truncation of the small-time series
alone would give. It is fine until it is not.

## Reference

Navarro, D. J. and Fuss, I. G. (2009). Fast and accurate calculations
for first-passage times in Wiener diffusion models. *Journal of
Mathematical Psychology*, 53(4), 222-230.

Ratcliff, R. and Tuerlinckx, F. (2002). Estimating parameters of the
diffusion model: approaches to dealing with contaminant reaction times
and parameter variability. *Psychonomic Bulletin & Review*, 9(3),
438-481.
