---
title: "Drift-diffusion models of two-choice response times"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drift-diffusion models of two-choice response times}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
has_rw <- requireNamespace("RWiener", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  # every fit and every reference value needs the simulator or the
  # reference density, so the whole vignette is gated on one flag
  eval = has_rw
)
library(frmtmb.ddm)
```

```{r, echo = FALSE, eval = !has_rw, comment = NA}
cat("RWiener is not installed, so this vignette was built without\n",
    "running its examples. Install RWiener to reproduce them.\n")
```

## What the model says

A participant sees a letter string and decides whether it is a word.
The drift-diffusion model says that evidence accumulates noisily from a
starting point until it reaches one of two boundaries, and the response
time is when it gets there. Four quantities describe that:

- `mu`, the **drift rate**: how fast evidence arrives, and in which
  direction. This is the one that usually carries the experimental
  effect.
- `bs`, the **boundary separation**: how much evidence is required.
  Wide boundaries mean slow, accurate responses.
- `ndt`, the **non-decision time**: encoding the stimulus and moving the
  finger, which no amount of deciding accounts for.
- `bias`, the **relative start point**: where accumulation begins
  between the boundaries. 0.5 is unbiased.

The names are brms's, so a model written for `brms::wiener()` reads the
same here.

## Simulating an experiment

`ddm_simulate()` draws response times and boundary choices jointly, the
way an experiment produces them. Here, words and non-words differ in
drift rate and in nothing else, which is the standard claim about a
lexical decision task.

```{r simulate}
set.seed(20240903)
n <- 1200
word <- rep(c(0, 1), each = n / 2)
dat <- ddm_simulate(n,
                    mu = 0.3 + 1.0 * word,   # words drift faster
                    bs = 1.5,
                    ndt = 0.30,
                    bias = 0.5)
dat$lex <- factor(word, labels = c("nonword", "word"))
head(dat)
```

`upper` is the decision: 1 for a response at the upper boundary, 0 for
the lower one.

```{r describe}
with(dat, tapply(rt, list(lex, upper), mean))
with(dat, tapply(rt, lex, length))
prop.table(table(dat$lex, dat$upper), margin = 1)
```

## The decision indicator

brms spells the boundary a trial ended at as `rt | dec(decision)`.
frmtmb's addition terms are a closed set and `dec()` is not one of them,
so this family reads the indicator from `vint()` instead:

```{r spelling, eval = FALSE}
# brms
brm(rt | dec(decision) ~ lex, family = wiener(), data = dat)

# here
frm(bf(rt | vint(upper) ~ lex), family = wiener(), data = dat)
```

`upper` must already be coded 0/1. Unlike `dec()`, `vint()` will not
take a factor, and passing anything else is refused rather than
silently coerced.

## Fitting

Drift depends on lexicality; everything else is constant across trials.
`bias = 0.5` fixes the start point on the response scale, which is the
usual choice when nothing in the design would bias it.

```{r fit}
fit <- frm(bf(rt | vint(upper) ~ lex, bias = 0.5),
           family = wiener(), data = dat)
summary(fit)
```

Read the coefficients on their own scales. Drift is on the identity
scale, so the `lexword` coefficient is the change in drift rate
directly.
Boundary separation is on the log scale. The non-decision time is on a
logit scaled onto `(0, min(rt))`, which is what keeps it inside the
support:

```{r coefs}
e <- unlist(fixef(fit))
c(drift_nonword = e[["mu.(Intercept)"]],
  drift_word_gain = e[["mu.lexword"]],
  boundary = exp(e[["bs.(Intercept)"]]),
  ndt = min(dat$rt) / (1 + exp(-e[["ndt.(Intercept)"]])))
```

Against the values the data were generated from -- drift 0.3 and 1.3,
boundary 1.5, non-decision time 0.30 -- that is the recovery this design
supports at `r n` trials.

## Why the non-decision time needs a bounded link

The density is exactly zero for a response time at or below `ndt`. A log
link would let the optimizer propose a non-decision time above the
fastest observed response, where the likelihood is not merely small but
undefined, and it would find out only by stepping there.

So `ndt` gets a logit scaled onto `(0, max_ndt)`, with `max_ndt`
defaulting to the smallest response time in the data. The constraint
becomes structural: no finite linear predictor can violate it.

```{r ndt-link}
lk <- wiener(max_ndt = 0.4)$links[["ndt"]]
lk$linkinv(c(-10, -2, 0, 2, 10))   # always strictly inside (0, 0.4)
```

Pass `max_ndt` explicitly when you intend to `predict()` on new data
whose fastest response differs from the training set's.

## Predictions

`fitted()` and `predict(type = "response")` give the mean response time,
conditional on the boundary that row ended at, in closed form.

```{r predict}
nd <- data.frame(lex = factor(c("nonword", "word")), upper = c(1, 1))
predict(fit, newdata = nd, type = "response")
```

`vint()` is required on `newdata` as well; there is no default boundary,
because the conditional mean genuinely depends on it.

One caution that looks like a bug and is not. With `bias` fixed at 0.5,
the conditional mean decision time is the **same** at both boundaries,
for any drift rate. That is a theorem about the diffusion, not an
oversight: an unbiased start gives fast and slow responses the same
average time whichever way they go. Free the bias and the two separate:

```{r bias-free}
set.seed(7)
d2 <- ddm_simulate(900, mu = 0.7, bs = 1.4, ndt = 0.25, bias = 0.35)
f2 <- frm(bf(rt | vint(upper) ~ 1), family = wiener(), data = d2)
c(pred_upper = predict(f2, newdata = data.frame(upper = 1),
                       type = "response"),
  pred_lower = predict(f2, newdata = data.frame(upper = 0),
                       type = "response"),
  obs_upper = mean(d2$rt[d2$upper == 1]),
  obs_lower = mean(d2$rt[d2$upper == 0]))
```

## Simulating from a fit

`simulate()` draws response times at each row's own boundary, by
inverse transform through the first-passage quantile function. Comparing
those draws with the observed times is the usual posterior-predictive
sanity check.

```{r simulate-fit}
set.seed(3)
sims <- as.matrix(simulate(fit, nsim = 100))
c(observed = mean(dat$rt), simulated = mean(sims))
quantile(dat$rt, c(0.1, 0.5, 0.9))
quantile(sims,   c(0.1, 0.5, 0.9))
```

## What this family does not do

- **`cens()` and `trunc()` are refused.** They need a distribution
  function, and the family declares none. The Wiener first-passage CDF
  is a third series with its own truncation problem, and it is not
  written here.
- **`residuals(type = "pearson")` and `"deviance"` are refused.** No
  variance function and no unit deviance. `type = "response"` works.
- **The decision is data, not a modelled outcome.** This family models
  the response time given the boundary. It does not jointly model which
  boundary was reached, so it is not a substitute for fitting choice and
  time together.

`frm_compat("wiener")` states each of these, along with what was tested
and what merely was not.

```{r compat}
frm_compat("wiener")[, c("feature_b", "status")]
```

## The density

The likelihood is the Wiener first-passage time density of Navarro and
Fuss (2009). They give two series for it: one that converges quickly at
small normalized times `u = t / a^2`, one that converges quickly at
large ones, and a criterion for choosing between them per evaluation.

An automatic-differentiation tape cannot make that choice. `u` depends
on the boundary separation, which is a parameter, so a comparison on `u`
is a branch on a parameter -- and RTMB advectors carry no comparison
operators, so it does not even tape. Branching on the response alone
would tape and would be wrong, because the same response time falls on
either side of the criterion depending on the boundary separation.

This package evaluates **both** series at a fixed truncation and blends
their logs with a logistic weight in `log(u)`: Navarro and Fuss's
criterion with the step smoothed. The weight saturates to exactly 0 and
exactly 1 outside a narrow band, and inside that band both series are
accurate, so any convex combination of them is accurate too.

The cost is that every evaluation pays for both series. What it buys is
a density that holds over the whole range of normalized times rather
than the part a single truncation happens to cover:

```{r accuracy}
u <- c(0.005, 0.05, 0.4, 2, 8, 30)
ref <- vapply(u, function(uu) {
  log(RWiener::dwiener(uu + 1e-12, 1, 1e-12, 0.45, 1, resp = "lower"))
}, 0)
mine <- vapply(u, function(uu) {
  frmtmb.ddm:::ddm_lpdf_lower(uu, 1, 1, 0.45)
}, 0)
small_only <- vapply(u, function(uu) {
  frmtmb.ddm:::ddm_log_gs(uu, 0.45) - 1 * 1 * 0.45 - 1 * 1 * uu / 2
}, 0)
data.frame(u,
           blended_rel_err = abs(mine - ref) / abs(ref),
           small_time_only_rel_err = abs(small_only - ref) / abs(ref))
```

The last column is what a fixed truncation of the small-time series
alone would give. It is fine until it is not.

## Reference

Navarro, D. J. and Fuss, I. G. (2009). Fast and accurate calculations
for first-passage times in Wiener diffusion models. *Journal of
Mathematical Psychology*, 53(4), 222-230.
