---
title: "Generalized drift-diffusion models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalized drift-diffusion models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(frmtmb)
library(frmtmb.ddm)
set.seed(20)
```

`vignette("ddm")` fits the drift-diffusion model whose density is known
in closed form. This one is about the models whose density is not.

## What the generalized model adds

The Wiener model of `wiener()` holds three things fixed. Evidence
accumulates at a constant rate, the two decision boundaries stay where
they are, and the accumulator has no memory of its own level. Drop any
of them and the first-passage density stops having a closed form.

`gddm()` is the generalized drift-diffusion model of Shinn, Lam and
Murray (2020). The accumulator follows

```
dx = a(x, t) dt + dW
```

from a starting distribution, and is absorbed at moving boundaries at
plus and minus `B(t)`. The response time is the absorption time plus a
non-decision time. Three things are now yours to choose:

- **the drift** `a(x, t)`, which may depend on a covariate and on the
  accumulator's own level. A drift that pulls the accumulator back
  toward zero is leaky integration; one that pushes it away is
  unstable.
- **the boundary** `B(t)`, which may collapse over the course of a
  trial, so that a decision gets easier to reach the longer it takes.
- **the starting distribution**, a point or an interval.

Each choice brings its own free parameters, and each of those takes a
formula like any other distributional parameter.

Two responses is the whole of it. One accumulator in one dimension
between two absorbing boundaries can end a trial at the upper wall or
the lower wall and nowhere else, so multi-alternative choice is a
different architecture rather than another parameter, and a third level
in the decision indicator is refused rather than folded into one of the
two.

## An experiment

The design is the one the paper fits: a motion discrimination task at
several coherences, where the drift grows with coherence and the
boundaries collapse as the trial runs on.

`gddm_simulate()` draws from the model's own density rather than from a
forward simulation of the diffusion, whose first passages arrive late by
however much the step misses excursions between monitoring times.

```{r simulate}
coh <- c(0, 0.128, 0.512)
ctl <- gddm_control(t_max = 2, dt = 0.02, ny = 101)
dat <- do.call(rbind, lapply(coh, function(cc) {
  gddm_simulate(300, mu = 6, alpha = 0.8, leak = 1, bs = 3, tau = 1.2,
                ndt = 0.25, coh = cc,
                drift = list(gddm_drift_coherence(cmax = 0.512),
                             gddm_drift_leak()),
                bound = gddm_bound_exponential(), control = ctl)
}))
dat$coh <- rep(coh, each = 300)
str(dat)
```

Accuracy rises with coherence, and responses get faster:

```{r describe}
data.frame(coh = coh,
           accuracy = round(tapply(dat$upper, dat$coh, mean), 3),
           mean_rt = round(tapply(dat$rt, dat$coh, mean), 3))
```

## What the data must carry

Two things beyond the response time, and both are data.

Which boundary a trial ended at reaches the family through `dec()`, as
it does for `wiener()`, or through `vint()`, which is the
general-purpose route and the one used below. The **condition** a trial
belongs to reaches it through `vint()`, and that one is particular to
this family: one solve of the Fokker-Planck equation serves every trial
that shares a parameter vector, and the family finds those trials
through an index it is given rather than by comparing parameter values,
which it cannot do on a tape.

`vint()` numbers its values positionally, so the two spellings put the
condition in different places, and the family reads whichever it is:

```
rt | dec(response) + vint(cond)   # boundary in dec(), condition first
rt | vint(upper, cond)            # boundary first, condition second
```

`gddm_conditions()` builds the index. Name every variable that appears
on the right-hand side of any formula in the model, and every covariate
a drift term reads.

```{r conditions}
dat$cond <- gddm_conditions(dat, coh)
table(dat$cond)
```

Naming more variables than you need is safe and only costs solves.
Naming fewer is wrong, and wrong in a way nothing downstream can
detect: the family checks that the covariates a drift reads are constant
within a condition, but it cannot check the rest.

## Fitting

```{r fit}
fit <- frm(bf(rt | vint(upper, cond) + vreal(coh) ~ 1, bias = 0.5),
           family = gddm(drift = list(gddm_drift_coherence(cmax = 0.512),
                                      gddm_drift_leak()),
                         bound = gddm_bound_exponential(),
                         control = ctl),
           data = dat)
summary(fit)
```

The estimates come back on the link scale, so put them back:

```{r estimates}
e <- unlist(fixef(fit))
data.frame(
  truth = c(mu = 6, alpha = 0.8, leak = 1, bs = 3, tau = 1.2, ndt = 0.25),
  estimate = round(c(
    mu = unname(e[["mu.(Intercept)"]]),
    alpha = exp(unname(e[["alpha.(Intercept)"]])),
    leak = unname(e[["leak.(Intercept)"]]),
    bs = exp(unname(e[["bs.(Intercept)"]])),
    tau = exp(unname(e[["tau.(Intercept)"]])),
    ndt = min(dat$rt) / (1 + exp(-unname(e[["ndt.(Intercept)"]])))), 3))
```

`bs` is the separation between the boundaries, not the distance from the
start to one of them, so it means what it means in `wiener()` and the
two families' estimates are directly comparable.

`leak` is positive for leaky integration and negative for unstable
integration. It is the paper's `l`, which is the negative of PyDDM's
`leak`.

## The numerical scheme, and why it is this one

There is no density to evaluate, so every likelihood evaluation solves
the Fokker-Planck equation forward in time and reads the probability
flux through each boundary. Three things about how.

**The boundaries are pinned, not chased.** Substituting `y = x / B(t)`
puts the walls at plus and minus one for all `t`. The spatial grid is
then fixed while the boundary collapses, and no grid index depends on a
parameter. This is what makes the model differentiable at all: the usual
treatment sandwiches a moving bound between two integer grid indices, so
the objective depends on where the bound falls between nodes, which is
why the reference implementation fits by differential evolution and
takes no derivatives.

**The scheme is Crank-Nicolson.** Once the walls are stationary it is
available, and it is second order in the step and unconditionally
stable. It is not available to a solver that chases the bound in the
original coordinate: the discretized operator is stiff and its
eigenvalues grow as the boundary collapses, so an explicit method would
need a step orders of magnitude smaller.

**The defective density is renormalized.** The discretized solve loses a
little probability mass, and how much it loses depends on the
parameters, because a configuration that absorbs faster loses less. A
likelihood that does not divide the loss out therefore pays a hidden
bonus for absorbing quickly, and the bonus lands squarely on the leak
and the boundary height. This is not a small effect: turning it off
moves the leak by more than a factor of two on data simulated from the
model itself. `gddm_control(renormalize = FALSE)` exists so the size of
that bias can be measured, not because it is ever the better model.

Renormalizing also makes the density explicitly conditional on a
response inside the modeled window, which is why `t_max` has to contain
every response time and is worth setting to the experiment's deadline
when there is one.

## How accurate, and how expensive

With a constant drift and boundaries that do not move, the generalized
model **is** the Wiener model, so the solver can be checked against the
closed form. At the shipped grid, over decision times from 0.2 s on and
across a range of drifts, separations and starting points, the two agree
to better than 0.01 in the log density. A coarser grid is worse and a
finer one is better; the package's own test suite pins all three.

Two honest limits. The density at very short decision times, a few steps
above zero, is much larger than the truth: an implicit scheme spreads a
little mass everywhere immediately, where the true first-passage density
is exponentially small. It is small in absolute terms, and adding a
lapse component floors it, but a fit should not be asked to read it.
And refining `ny` alone helps only because the starting mass is
deliberately spread over two grid cells rather than one; a sharper start
would ring, because Crank-Nicolson does not damp the highest frequencies
a grid carries.

The cost scales with the number of **conditions**, not the number of
trials, because one solve serves a whole condition. It also scales with
`t_max / dt`. Both the tape build and the evaluation grow with them, and
`gddm_control(tridiagonal =)` trades one against the other: the default
records the solve into the tape, which is slow to build and fast to run,
while `"atomic"` collapses it into a single node with a hand-written
derivative, which is quick to build and slower to run. The default suits
a fit, which evaluates the objective many times.

## When a row falls off the grid

The density is floored before it is logged. Where the solved density at
a trial's own response time underflows, the log density is a large
finite negative number rather than `NaN`. That matters twice: `NaN` is
not a value a line search can use, and inside `mixture()` one `NaN`
takes every other component with it through the log-sum-exp. The floored
row is flat, so its gradient is exactly zero, which is what makes it
harmless rather than merely quiet.

Being quiet is the cost, so the count is available to read:

```{r floored}
gddm_floored(fit)
```

Zero means the grid represented every observation. A few rows means a
few trials sit within a few time steps of the fitted non-decision time,
where a fixed grid cannot resolve a density that is climbing through
orders of magnitude, and those rows contributed a constant rather than
information. Many rows means the fit is not to be trusted: shrink `dt`,
or give the model a lapse component, which floors the density in the
model rather than in the arithmetic.

## When not to use this

Use `wiener()` whenever it applies. That is whenever the drift is
constant in the state and in time and the boundaries do not move: there
the first-passage density is a known pair of series and costs
arithmetic, while here every evaluation solves a partial differential
equation once per condition. The difference is orders of magnitude.

What `gddm()` buys is the class of models the closed form cannot express
at all. If the scientific question is whether the boundaries collapse,
or whether integration is leaky, there is no analytic alternative to
compare against.

## Writing your own component

The catalogue is meant to be extended. A drift term is the value of
`gddm_drift_term()`, a boundary is `gddm_bound_term()` and a starting
distribution is `gddm_start_term()`. Each carries its own free
parameters, each of which then takes a formula like any other.

A boundary that decays toward a floor rather than toward zero, say:

```{r custom}
bound_floor <- function() {
  gddm_bound_term(
    "floored",
    dpars = list(
      bs = list(link = "log", init = function(y, aterms) 1.5),
      tau = list(link = "log", init = function(y, aterms) max(y)),
      floor = list(link = "logit", init = function(y, aterms) 0.3)),
    fn = function(t, p, ctl) {
      f <- p$floor + (1 - p$floor) * exp(-t / p$tau)
      list(B = 0.5 * p$bs * f,
           dlogB = -((1 - p$floor) / p$tau) * exp(-t / p$tau) / f)
    })
}
gddm(bound = bound_floor())[["dpars"]]
```

Two rules. Nothing in a component may compare a parameter against
anything, because a tape records no comparisons. And a boundary returns
its logarithmic derivative alongside its value, because the change of
variable that pins the walls needs it and differencing the boundary
would cost the scheme its accuracy.

## Reference

Shinn, M., Lam, N. H. and Murray, J. D. (2020). A flexible framework for
simulating and fitting generalized drift-diffusion models. *eLife*, 9,
e56938.
