---
title: "How to report a fitted curve"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to report a fitted curve}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
has_tinyplot <- requireNamespace("tinyplot", quietly = TRUE)
library(frmtmb)
library(frmtmb.spline)
```

You fitted a smooth. Now you have to report it, and a plot of the
fitted line with a shaded band is not yet a claim. This article shows
how to make three claims that are, on a movement dataset where the
truth is known:

* the curve lies inside a band, everywhere at once;
* the curve is rising here and falling there;
* the peak is at this time, plus or minus this much.

## The data

A reaching movement has a bell-shaped speed profile. Twenty subjects
each make twelve reaches; every subject's profile has its own height
and its own timing, and the measured speed is noisy.

```{r data}
set.seed(4)
n_sub <- 20
n_rep <- 12
n_t <- 30
peak <- function(t, h, s) h * exp(-0.5 * ((t - s) / 0.16)^2)
sub <- rep(seq_len(n_sub), each = n_rep * n_t)
d <- data.frame(
  subject = factor(sub),
  trial = rep(seq_len(n_sub * n_rep), each = n_t),
  t = rep(seq(0, 1, length.out = n_t), times = n_sub * n_rep))
h_sub <- rnorm(n_sub, 1, 0.12)
s_sub <- rnorm(n_sub, 0.5, 0.04)
d$v <- peak(d$t, h_sub[sub], s_sub[sub]) + rnorm(nrow(d), 0, 0.06)
str(d, max.level = 1)
```

What is the population curve here? Not one subject's. Each subject
peaks at its own time, so the AVERAGE of the twenty curves is a wider,
slightly lower bump than any of them. That average is what a population
smooth estimates, and it is the thing to compare against:

```{r truth}
pop_curve <- function(tt) {
  rowMeans(vapply(seq_len(n_sub),
                  function(i) peak(tt, h_sub[i], s_sub[i]),
                  numeric(length(tt))))
}
c(peak_of_one_subject = max(peak(seq(0, 1, 0.001), 1, 0.5)),
  peak_of_the_average = max(pop_curve(seq(0, 1, 0.001))))
```

Nothing below is told any of it.

## Fit it

`s(t)` is the population curve and `s(t, subject, bs = "fs")` gives
every subject its own departure from it, with one shared smoothing
parameter. This is an ordinary `frm()` call; nothing in this package
changes how a model is fitted.

```{r fit}
fit <- frm(bf(v ~ s(t, k = 12) + s(t, subject, bs = "fs", k = 5)),
           family = gaussian(), data = d)
VarCorr(fit)
```

## 1. The curve, with a band that covers all of it

`frm_curve()` evaluates the fitted linear predictor on a grid.
`re.form = NA` drops the per-subject curves, so what comes back is the
POPULATION curve.

```{r curve}
grid <- data.frame(t = seq(0, 1, length.out = 80))
cv <- frm_curve(fit, newdata = grid, re.form = NA, nsim = 20000, seed = 1)
cv
```

Two intervals come back. `.lower_ci` and `.upper_ci` are the ordinary
pointwise interval: each point of the curve, taken on its own, is
inside its own interval 95 percent of the time. `.lower_sim` and
`.upper_sim` are the SIMULTANEOUS band: the WHOLE curve is inside it 95
percent of the time.

Those are different claims and the second is the one a figure invites.
A 95 percent pointwise band drawn over 80 points does not cover the
whole curve 95 percent of the time; it covers it far less often,
because there are 80 chances to escape. The simultaneous critical value
is the correction, and here it is about 2.7 rather than 1.96:

```{r crit}
c(pointwise = cv$.crit[1], simultaneous = cv$.crit_sim[1],
  ratio = cv$.crit_sim[1] / cv$.crit[1])
```

The simultaneous critical value is simulated, so it comes with its own
Monte Carlo standard error. Report it, or a reader cannot tell your
2.72 from someone else's 2.70:

```{r mcse}
attr(cv, "check")$crit_mcse
```

The `"check"` attribute carries one more number, and it is the one that
licenses everything above. frmtmb exports no route to the joint
covariance of a grid prediction, so this package rebuilds it and then
checks its own answer against `predict(se.fit = TRUE)`, which is
exported. The two agree to about twelve significant figures:

```{r check}
attr(cv, "check")$cov_rel_error
```

If they ever did not, `frm_curve()` would refuse rather than report a
band. The count of `predict()` calls the rebuild cost is there too, and
it is small: at `re.form = NA` the per-subject coefficients contribute
nothing and are skipped in blocks rather than one at a time.

```{r ncall}
attr(cv, "check")$n_predict
```

```{r fig-curve, eval = has_tinyplot, fig.alt = "The fitted population speed profile against time, a bell shape peaking near 1.0 at t = 0.5. A wide pale band is the simultaneous interval and a narrower darker band the pointwise interval. A dashed line, the truth, lies inside both across the whole range."}
tt <- cv$t
est <- cv$.estimate
tinyplot::tinyplot(x = tt, y = est, type = "l", lwd = 2,
                   col = "steelblue4", theme = "clean2",
                   ylim = range(cv$.lower_sim, cv$.upper_sim),
                   xlab = "time", ylab = "speed",
                   main = "Population speed profile")
tinyplot::tinyplot_add(x = c(tt, rev(tt)),
                       y = c(cv$.lower_sim, rev(cv$.upper_sim)),
                       type = "polygon", col = "#4682b422", border = NA)
tinyplot::tinyplot_add(x = c(tt, rev(tt)),
                       y = c(cv$.lower_ci, rev(cv$.upper_ci)),
                       type = "polygon", col = "#4682b444", border = NA)
tinyplot::tinyplot_add(x = tt, y = est, type = "l", lwd = 2,
                       col = "steelblue4")
tinyplot::tinyplot_add(x = grid$t, y = pop_curve(grid$t), type = "l",
                       lty = 2, col = "grey30")
```

The population curve is inside the simultaneous band at every grid
point, which is what the band claims:

```{r cover}
truth <- pop_curve(grid$t)
c(inside_simultaneous = mean(truth >= cv$.lower_sim & truth <= cv$.upper_sim),
  inside_pointwise = mean(truth >= cv$.lower_ci & truth <= cv$.upper_ci))
```

Both bands cover here, and that is not a contradiction: the difference
between them is a statement about REPEATED datasets, not about this one.
Over many datasets the pointwise band misses the whole curve far more
often than 5 percent of the time and the simultaneous band misses it 5
percent of the time.

Two cautions about what "the truth" means. A band covers the curve the
MODEL is estimating, so with 7200 observations the sampling error is
small and the penalty's bias is most of what is left; and the estimand
is the average of the twenty subject curves, not any one of them.
Comparing against one subject's profile would put the "truth" outside
the band over part of the range, and the band would not be at fault.

## 2. Where the curve is rising

The slope of the curve is a curve of its own, and it has the same two
intervals. Where its band excludes zero, the speed is changing.

```{r deriv}
g2 <- data.frame(t = seq(0.05, 0.95, length.out = 40))
d1 <- frm_curve_deriv(fit, var = "t", order = 1, newdata = g2,
                      re.form = NA, nsim = 20000, seed = 2)
rising <- d1$.lower_sim > 0
falling <- d1$.upper_sim < 0
c(rising_from = min(g2$t[rising]), rising_to = max(g2$t[rising]),
  falling_from = min(g2$t[falling]), falling_to = max(g2$t[falling]))
```

Read that as a sentence: the speed is increasing over the first
interval and decreasing over the second, and both statements hold
simultaneously at the stated level.

The step size is worth one paragraph. The derivative is a central
difference of the DESIGN, and its step defaults to a fraction of the
covariate's range, set per order: `1e-6` of the range at order 1 and
`1e-4` at order 2. A single fixed step cannot serve both, because the
second difference divides by the square of the step and turns
floating-point cancellation into the dominant error long before the
first difference notices. Pass `eps =` to override it.

```{r eps}
c(order_1 = attr(d1, "eps"),
  order_2 = attr(frm_curve_deriv(fit, var = "t", order = 2, newdata = g2,
                                 re.form = NA, simultaneous = FALSE),
                 "eps"))
```

```{r fig-deriv, eval = has_tinyplot, fig.alt = "The first derivative of the fitted speed profile against time. It rises from zero, peaks near 3 at t = 0.35, crosses zero at t = 0.5 and falls to about minus 3 at t = 0.65 before returning toward zero. A shaded band around it excludes zero on both sides of the crossing."}
t2 <- d1$t
tinyplot::tinyplot(x = t2, y = d1$.estimate, type = "l", lwd = 2,
                   col = "firebrick", theme = "clean2",
                   ylim = range(d1$.lower_sim, d1$.upper_sim),
                   xlab = "time", ylab = "d speed / d t",
                   main = "Slope of the population profile")
tinyplot::tinyplot_add(x = c(t2, rev(t2)),
                       y = c(d1$.lower_sim, rev(d1$.upper_sim)),
                       type = "polygon", col = "#b2222222", border = NA)
tinyplot::tinyplot_add(x = t2, y = d1$.estimate, type = "l", lwd = 2,
                       col = "firebrick")
abline(h = 0, lty = 3)
```

## 3. Where the peak is

A movement paper reports the TIME of peak speed, and usually reports it
without an interval, because nothing hands one over. `frm_curve_feature()`
does. The peak solves `f'(t) = 0`, so perturbing the fit moves it, and
the implicit-function theorem turns the curve's covariance into the
variance of its location:

`var(t*) = var(f'(t*)) / f''(t*)^2`

```{r peak}
pk <- frm_curve_feature(fit, var = "t", type = "maximum", newdata = g2,
                        re.form = NA)
pk
```

The average curve peaks near `t = 0.5`, and the estimate is inside a
few hundredths of it. Note that `.value_se` is NOT inflated by the uncertainty in
where the peak is: at a stationary point the derivative of the height
with respect to the coefficients loses its location term, because
`f'(t*)` is zero there. The height's standard error is the ordinary
pointwise one at `t*`.

Movement onset is the same calculation against a threshold instead of
against zero. Every crossing the grid brackets comes back, so a
threshold the profile passes twice gives two rows:

```{r onset}
frm_curve_feature(fit, var = "t", type = "crossing", at = 0.2,
                  newdata = g2, re.form = NA)
```

## Per-subject curves

Drop `re.form = NA` and supply the grouping column, and the same three
functions describe THAT subject's curve, with the subject's own
deviation and its uncertainty included.

```{r subject}
gs <- data.frame(t = seq(0.05, 0.95, length.out = 40),
                 subject = factor(3, levels = levels(d$subject)))
pk3 <- frm_curve_feature(fit, var = "t", type = "maximum", newdata = gs,
                         re.form = NULL)
pk3[, c(".estimate", ".se", ".value", ".value_se")]
```

Subject 3's own peak time, against the population peak above. Repeat
over subjects and you have per-subject peak latencies with standard
errors, which is what a second-level analysis wants as input.

## What to report

* The simultaneous band, not the pointwise one, whenever the figure
  invites a statement about the shape of the curve.
* The simultaneous critical value WITH its Monte Carlo standard error.
* For a feature, the estimate and its interval, and say which
  derivative the implicit-function delta method was applied to.
