---
title: "How to fit a flexible parametric survival model"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to fit a flexible parametric survival model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2)
has_tinyplot <- requireNamespace("tinyplot", quietly = TRUE)
has_flexsurv <- requireNamespace("flexsurv", quietly = TRUE)
knitr::opts_chunk$set(eval = has_flexsurv)
library(frmtmb)
library(frmtmb.spline)
```

A Cox model leaves the baseline hazard unspecified, which is what makes
it robust and also what stops it from telling you the hazard. A
Royston-Parmar model writes the log cumulative hazard as a natural
cubic spline in log time. You get proportional hazards when you want
them, a smooth estimated baseline you can plot, and time-varying
effects when the proportional-hazards assumption fails.

`royston_parmar()` is that family, parameterized exactly as
`flexsurv::flexsurvspline()` parameterizes it.

## The data

`flexsurv`'s `bc` data: 686 breast cancer patients in three prognostic
groups, recurrence-free survival in years, 299 events and 387
right-censored rows.

```{r data}
data(bc, package = "flexsurv")
table(bc$group, event = bc$censrec)
```

One thing to get right before anything else. frmtmb's `cens()` reads
`0` as an observed event and `1` as right censored. A `Surv()` status
column is the other way round. Convert it, once, and name the column
for what it holds:

```{r cens}
bc$censored <- 1 - bc$censrec
```

## Fit it

`df` is the degrees of freedom of the spline: the number of interior
knots plus one. `df = 3` gives two interior knots and four coefficients,
which is Royston and Parmar's usual starting point. The knots go at
equally spaced quantiles of the log UNCENSORED times, which needs the
response, so they are chosen when `frm()` assembles the model frame.

```{r fit}
fit <- frm(bf(recyrs | cens(censored) ~ group),
           family = royston_parmar(df = 3), data = bc)
fixef(fit)
```

`mu` is `gamma0`, the spline intercept, and the `group` coefficients sit
on it. Because `gamma0` shifts the whole log cumulative hazard, those
are LOG HAZARD RATIOS: the model is proportional hazards.

```{r hr}
exp(unlist(fixef(fit)$mu)[-1])
```

`gamma1` and upward are the other spline coefficients, and together with
`gamma0` they are the baseline.

## It is the same model flexsurv fits

Not "close to". The same. Both packages evaluate one function, and the
proof is to hand flexsurv's fitted coefficients to frmtmb's objective:

```{r identity}
fs <- flexsurv::flexsurvspline(survival::Surv(recyrs, censrec) ~ group,
                               data = bc, k = 2, scale = "hazard")
kn <- fs$knots
o <- frm(bf(recyrs | cens(censored) ~ group),
         family = royston_parmar(knots = kn[2:(length(kn) - 1)],
                                 bknots = kn[c(1, length(kn))]),
         data = bc, dry_run = "objective")
cf <- fs$res[, "est"]
gam <- cf[grep("^gamma", names(cf))]
par <- c(gam[1], cf[!grepl("^gamma", names(cf))], gam[-1])
c(flexsurv = fs$loglik, frmtmb_at_flexsurv_par = -o$obj$fn(par))
```

Those two numbers agree to about ten units in the last place of a
number near 800. Note `flexsurv`'s `k` is this package's `df - 1`: `k`
counts the interior knots and `df` counts them plus one.

The two optima then agree as well:

```{r optima}
c(flexsurv = fs$loglik, frmtmb = as.numeric(logLik(fit)))
```

## Read the baseline off

Every spline coefficient is a distributional parameter, so
`predict(type = "link", dpar = )` reaches each of them and the fitted
baseline is the basis times those coefficients. There is no penalized
smooth in this model, so `frm_curve()` has no random-effect block to
read and refuses; add an `s()` term and it applies here as it does
anywhere else.

```{r baseline}
tg <- exp(seq(log(0.3), log(max(bc$recyrs)), length.out = 60))
nd <- data.frame(recyrs = tg,
                 group = factor("Good", levels = levels(bc$group)))
lp <- lapply(c("mu", "gamma1", "gamma2", "gamma3"), function(dp) {
  predict(fit, newdata = nd, type = "link", dpar = dp)
})
kn <- environment(family(fit)[["lpdf"]])$allknots
x <- log(tg)
# the Royston-Parmar basis, written out: a constant, x itself, and one
# natural cubic term per interior knot, each linear beyond the boundary
rp_basis <- function(knots, x) {
  nk <- length(knots)
  out <- list(rep(1, length(x)), x)
  for (j in seq_len(nk - 2)) {
    kj <- knots[j + 1]
    lam <- (knots[nk] - kj) / (knots[nk] - knots[1])
    out[[j + 2]] <- pmax(x - kj, 0)^3 - lam * pmax(x - knots[1], 0)^3 -
      (1 - lam) * pmax(x - knots[nk], 0)^3
  }
  out
}
bas <- rp_basis(kn, x)
logH <- Reduce(`+`, Map(function(b, g) b * as.numeric(g), bas, lp))
S <- exp(-exp(logH))
head(data.frame(t = round(tg, 3), logH = round(logH, 3),
                survival = round(S, 3)), 4)
```

```{r fig-surv, eval = has_tinyplot, fig.alt = "Fitted survival curves for the three prognostic groups against time in years, each falling from 1. The good-prognosis curve stays above 0.8 at eight years, the medium curve reaches about 0.6 and the poor curve about 0.3. Step functions of the same three colors, the Kaplan-Meier estimates, track each fitted curve closely."}
cols <- c("steelblue4", "goldenrod3", "firebrick")
surv_of <- function(g) {
  ndg <- data.frame(recyrs = tg,
                    group = factor(g, levels = levels(bc$group)))
  lps <- lapply(c("mu", "gamma1", "gamma2", "gamma3"), function(dp) {
    as.numeric(predict(fit, newdata = ndg, type = "link", dpar = dp))
  })
  exp(-exp(Reduce(`+`, Map(`*`, bas, lps))))
}
tinyplot::tinyplot(x = tg, y = surv_of("Good"), type = "l", lwd = 2,
                   col = cols[1], theme = "clean2", ylim = c(0, 1),
                   xlab = "years", ylab = "survival",
                   main = "Fitted survival against Kaplan-Meier")
for (i in 2:3) {
  tinyplot::tinyplot_add(x = tg, y = surv_of(levels(bc$group)[i]),
                         type = "l", lwd = 2, col = cols[i])
}
km <- survival::survfit(survival::Surv(recyrs, censrec) ~ group, data = bc)
km_s <- summary(km)
for (i in seq_along(levels(bc$group))) {
  k <- km_s$strata == levels(km_s$strata)[i]
  lines(km_s$time[k], km_s$surv[k], type = "s", col = cols[i], lty = 3)
}
legend("bottomleft", levels(bc$group), col = cols, lwd = 2, bty = "n")
```

The fitted curves track the Kaplan-Meier estimates, which the model was
never shown.

## When hazards are not proportional

A formula on `gamma0` shifts the whole curve, which is proportional
hazards. A formula on any OTHER coefficient multiplies a function of log
time, so it is a TIME-VARYING effect: `flexsurv` spells this `anc =`,
and here it is an ordinary second formula.

```{r tvc}
ftv <- frm(bf(recyrs | cens(censored) ~ group, gamma1 ~ group),
           family = royston_parmar(df = 3), data = bc)
c(proportional = as.numeric(logLik(fit)),
  time_varying = as.numeric(logLik(ftv)),
  AIC_proportional = AIC(fit), AIC_time_varying = AIC(ftv))
```

`gamma1` multiplies `log(t)`, so letting it vary by group lets each
group's hazard ratio drift with time rather than holding still.

## What this family will not do

`fitted()` and `predict(type = "response")` refuse, and say why. The
mean of a survival time under this model is an integral of the fitted
survival function with no closed form, and the censored rows do not
identify its upper tail. `mu` is a spline coefficient, not a fitted
value, and the identity link would let it pass for one:

```{r refuse, error = TRUE}
fitted(fit)
```

## The check you must run

Two things in this family are floors rather than answers, and both are
silent in the fitted object. `rp_floored()` is where they go to be read,
and it REFUSES rather than reports:

```{r floored}
rp_floored(fit, action = "report")
```

Zero on both counts here. Run it on every fit whose data carry
censoring, because neither `logLik()` nor `AIC()` can tell you.

**Why it matters.** frmtmb forms a right-censored contribution as
`log(1 - F(y))` on the probability scale, and core gives a family no
complementary log-CDF slot to hand back `log S` directly. So the scored
`log S` carries absolute error about `eps / S`: exact at `-log S` of 10,
wrong by 1.7e-04 at 30, and floored at -35.127363 past 36. Past 30 its
GRADIENT is exactly zero, so the optimizer prices such a row at a
constant and fits the others as if it were free.

That is not hypothetical. A 600-subject dataset with one subject
censored far beyond every event time converges without a warning and
reports a log likelihood thousands of units away from the model's, with
a treatment coefficient wrong by tens of percent; `flexsurv` declines to
fit the same data at all. `rp_floored()` refuses it:

```{r floored-bad, error = TRUE}
set.seed(20260905)
n <- 600
bad <- data.frame(grp = factor(rep(c("A", "B"), each = n / 2)))
bad$t <- rweibull(n, shape = 1.6,
                  scale = ifelse(bad$grp == "A", 0.30, 3.0))
bad$censored <- 0L
bad$t[1] <- 50
bad$censored[1] <- 1L
bad_fit <- frm(bf(t | cens(censored) ~ grp),
               family = royston_parmar(df = 3), data = bad)
c(converged = bad_fit$opt$convergence, logLik = as.numeric(logLik(bad_fit)))
rp_floored(bad_fit)
```

The refusal is POST-FIT and cannot be otherwise: `logLik()` reads the
optimizer's own value and the family protocol has no hook that runs when
a fit finishes. `frm_curve()` and its two companions call `rp_floored()`
for you, so the documented way to inspect this family will not draw a
curve off a fit like that one.

Monotonicity of the cumulative hazard is the second floor, and it is not
enforced; `flexsurv` does not enforce it either, and a spline that turns
over inside the data range is an over-parameterized fit rather than
something the software should have prevented. Where it turns over there
is no hazard and the true log density is `-Inf`, so this family floors
it to keep the optimizer alive, which makes `logLik()` a
pseudo-likelihood. `rp_floored()` counts those rows too. You can see the
same thing directly in the slope of the fitted log cumulative hazard,
which must stay positive:

```{r monotone}
dlogH <- diff(logH) / diff(x)
c(min_slope = min(dlogH), all_increasing = all(dlogH > 0))
```

## How many knots

More knots buy a more flexible baseline and cost degrees of freedom.
Compare on AIC, as Royston and Parmar do:

```{r df}
sapply(1:5, function(k) {
  AIC(frm(bf(recyrs | cens(censored) ~ group),
          family = royston_parmar(df = k), data = bc))
})
```
