---
title: "Learning models with covariates on their parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Learning models with covariates on their parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(frmtmb.learn)
```

A reinforcement-learning model is usually fitted one subject at a time,
and the group question is then asked of the estimates: are the learning
rates different between conditions? That two-step answer throws away the
uncertainty in each estimate before comparing them.

This package writes those models as frmtmb families, so every parameter
is an ordinary distributional parameter with its own linear predictor. A
learning rate takes a condition effect the way a mean does, and the
effect comes back with a standard error from one fit.

## The task

A probabilistic reversal task. One arm pays with probability 0.7 and the
other with 0.3, and halfway through the session they swap. The design
carries the payoff schedule of BOTH arms on every trial, which is what
makes a draw from the model coherent: a simulated subject that takes the
arm the real one did not still has to be paid.

```{r design}
set.seed(1)
d <- frm_task_design("reversal", n_subject = 20, n_trial = 60, seed = 1)
head(d)
table(d$after_reversal)
```

Now draw choices from the model, with a learning rate that really does
differ before and after the reversal, and real spread between subjects.
`frm_task_simulate()` takes parameters on their natural scales, one
value per subject.

```{r simulate}
subj <- levels(d$id)
u <- stats::rnorm(length(subj), 0, 0.5)
after <- d$after_reversal == "after"
eta <- stats::qlogis(0.30) + 1.1 * after + u[as.integer(d$id)]
d$choice <- frm_task_simulate(
  bandit2arm_delta(subject = id, trial = trial), d,
  pars = list(alpha = stats::plogis(eta), tau = 3), seed = 1)[[1]]$choice
mean(d$choice == 1)
```

## The fit

This is an ordinary `frm()` call. The main right-hand side is the
LEARNING RATE's predictor, because the family declares `alpha` as its
primary distributional parameter; `tau`, the softmax sensitivity, gets
its own.

```{r fit}
fit <- frm(bf(choice | reward(pay1, pay2) ~ after_reversal + (1 | id),
              tau ~ 1),
           family = bandit2arm_delta(subject = id, trial = trial),
           data = d)
fixef(fit)
```

The second coefficient is the thing the study is about, and it has an
interval:

```{r effect}
ci <- confint(fit)
ci[grepl("after_reversalafter", rownames(ci)), , drop = FALSE]
```

On the probability scale, the fitted learning rate before and after:

```{r scale}
b <- unlist(fixef(fit))
c(before = stats::plogis(b[["alpha.(Intercept)"]]),
  after  = stats::plogis(b[["alpha.(Intercept)"]] +
                           b[["alpha.after_reversalafter"]]))
```

Nothing about that formula is special to a reversal. `after_reversal`
could be a clinical group, a drug dose, `s(trial)` for a rate that
drifts smoothly through the session, or a random slope. A separate `prl`
family would fit one rate before and one after and leave you to compare
them; this fits the DIFFERENCE.

## Reading the trajectory

The value estimates and prediction errors are not in the fitted object.
They are recomputed by replaying the recursion at the estimates.

```{r trace}
tr <- frm_value_trace(fit)
head(tr)
```

`q1` and `q2` are the values the choice on that trial was made ON,
before the outcome moved them, which is the ordering a plot of learning
needs. `pe` is the prediction error and `p` is the fitted probability of
the choice that was actually made.

```{r plot, fig.width = 7, fig.height = 4, fig.alt = "Mean fitted value of each arm across trials, with the reversal marked"}
agg <- aggregate(cbind(q1, q2) ~ trial, tr, mean)
long <- rbind(data.frame(trial = agg$trial, value = agg$q1, arm = "arm 1"),
              data.frame(trial = agg$trial, value = agg$q2, arm = "arm 2"))
if (requireNamespace("tinyplot", quietly = TRUE)) {
  tinyplot::tinyplot(value ~ trial | arm, data = long, type = "l", lwd = 2,
                     main = "Mean fitted arm value", ylab = "fitted value")
} else {
  plot(agg$trial, agg$q1, type = "l", ylim = range(agg$q1, agg$q2),
       xlab = "trial", ylab = "fitted value", lwd = 2,
       main = "Mean fitted arm value")
  lines(agg$trial, agg$q2, lty = 2, lwd = 2)
  legend("topright", c("arm 1", "arm 2"), lty = c(1, 2), lwd = 2, bty = "n")
}
abline(v = 30, col = "grey60")
```

The two curves cross after the reversal, which is the model tracking the
contingency change.

`p` is the per-trial factor of the likelihood, so `sum(log(p))` is the
CONDITIONAL data log-likelihood given the fitted subject effects. In a
fit with no random effects that is `logLik(fit)` exactly; here it is
not, because `logLik()` reports the Laplace-approximated MARGINAL
likelihood, which carries the random-effect density and a curvature term
besides.

```{r identity}
c(conditional = sum(log(tr$p)), marginal = as.numeric(logLik(fit)))
```

## What refuses, and why

These families declare no mean on the response scale. The response is
the option a subject took, coded 1 to K, and it is nominal: arm 2 is not
twice arm 1, and the Iowa gambling task's four decks have no order at
all. Core forms a residual as `y - mean`, so a mean declared here would
make `fitted()`, `predict(type = "response")` and every residual return
arithmetic on a category code.

```{r refuse-fitted, error = TRUE}
fitted(fit)
```

`frm_value_trace()` above returns everything a mean would have carried,
and more. `frm_compat()` is the place to ask what else refuses:

```{r compat}
frm_compat("bandit2arm_delta", "residuals")$status
frm_compat("bandit2arm_delta", "s()")$status
frm_compat("bandit2arm_delta", "importance")$status
```

That last one refused in the first release and works now. The
correction reweights draws from the Laplace Gaussian and needs one
log-likelihood value per subject; the families declare them, so it runs.
The next section says what it is worth.

## The Laplace caveat

frmtmb integrates the subject effects out with a Laplace approximation,
which is exact only when the conditional log-density is quadratic. For
binary choices it is not, and the fewer trials a subject has the less
quadratic it is. The consequence was measured by simulation, 100
replicates at 40 subjects, the full tables in `?frmtmb.learn`:

| parameter | bias at 100 trials | coverage | bias at 20 trials | coverage |
|---|---|---|---|---|
| `alpha_(Intercept)` | +0.009 | 0.97 | -0.026 | 0.92 |
| `alpha_after` | -0.015 | 0.94 | +0.009 | 0.91 |
| `tau_(Intercept)` | +0.003 | 0.95 | +0.016 | 0.96 |
| `log sd(alpha)` | -0.266 | 0.99 | -1.596 | 0.95 |

The fixed effects survive short sessions: at a fifth of the trials their
bias is still inside Monte Carlo error and their intervals still cover
near the nominal rate. Report a condition effect on a learning rate from
a short session.

The variance component's point estimate does not survive, and its
interval does. At 20 trials `log sd(alpha)` comes out 1.6 too low, but
the spread of the estimates rises from 0.88 to 3.01, so the interval
widens at least as fast as the estimate degrades and still covers 0.95.
Treat a subject-level standard deviation from twenty binary trials as a
lower bound, and read its interval rather than its point estimate.

`frm(importance =)` is what separates Laplace error from the ordinary
downward bias of a variance component estimated from binary data with
few levels, and it now runs. Measured on this design, six datasets per
cell: at 100 trials, where the Laplace fit has kept a real variance
component, the correction moves `sd(alpha)` up by 0.11 to 0.18 log
units and toward the truth, with a Monte Carlo standard error a third
of that and effective sample sizes near 1. So part of the downward bias
at 100 trials IS Laplace error, and it is a modest part.

At 20 trials the correction has nothing to work with. Five of six
Laplace fits have already collapsed `sd(alpha)` to 1e-4, and the
correction then walks at its own step cap for every round while warning
that it did not converge. A shift reported alongside that warning is
the step cap rather than an estimate, so check `sqrt(VarCorr(fit))` and
read the warning before believing one. The warning advises raising
`frmtmb_control(importance_rounds =)`, and on a collapsed component
that makes the artifact bigger in exact proportion: five rounds of a
0.3645 cap give 1.822 and ten give 3.645. `?frmtmb.learn` carries the
per-dataset table.

## The families

```{r families}
fams <- frm_learn_families()
knitr::kable(fams[, c("family", "hbayesdm", "task", "options")])
```

Each family's parameters, and the map to hBayesDM's spelling of them:

```{r pars}
knitr::kable(fams[, c("family", "pars", "hbayesdm_pars")])
```

`alpha` is always a learning rate on (0, 1) and `tau` is always the
softmax sensitivity on (0, Inf), so what a formula does to `alpha` means
the same thing whichever family it is written against. Two cautions the
map does not remove. Core refuses a parameter name containing a dot or
an underscore, which is why the dual-rate family spells its rates `Arew`
and `Apun`. And `lambda` is used by the literature for three unrelated
quantities: the arm decay in the Kalman filter, the eligibility trace in
the two-step model, and loss aversion in the Iowa gambling task. Each
family's help says which. In particular, hBayesDM's `alpha` for
`igt_pvl_delta` is this package's `shape`, not this package's `alpha`.

All eight share one recursion, which walks trials once and updates
every subject at each step. A family supplies a choice rule and a
learning rule and is a few dozen lines; the walk, the padding for
unequal trial counts, the softmax, and the three modes it runs in
(likelihood, fitted trajectory, simulation) are written once.

`rlddm()` is the one that does not softmax. Its value difference drives
the drift rate of a two-boundary diffusion, so a trial contributes the
joint density of which boundary was reached and when, and choices and
response times are modeled together. Its response is the response TIME
and the boundary travels in `dec()`; the density comes from
`frmtmb.eam::wiener_lpdf()`.

A second family, to show how little changes. The dual-rate learner
absorbs good news and bad news at different speeds, and the asymmetry is
the parameter most of the clinical literature is about:

```{r dual}
fit2 <- frm(bf(choice | reward(pay1, pay2) ~ 1, Apun ~ 1, tau ~ 1),
            family = bandit2arm_dual(subject = id, trial = trial),
            data = d)
b2 <- unlist(fixef(fit2))
c(rate_after_reward = stats::plogis(b2[["Arew.(Intercept)"]]),
  rate_after_loss   = stats::plogis(b2[["Apun.(Intercept)"]]))
```

On BINARY payoffs the two settings of `split` are the same model, which
is worth knowing before reading anything into the choice: the value
store stays inside the payoff range, so the sign of the prediction error
and the sign of the outcome agree on every trial. They differ only when
the payoff is graded. See `?bandit2arm_dual`.

## Where to go next

* `?bandit2arm_delta` and the other seven family pages for the
  equations and each family's own cautions.
* `?frm_value_trace` for the trajectory, and `?frm_task_simulate` for
  drawing datasets from the generative process.
* `?frmtmb.learn` for the Laplace numbers, the recovery tables, and
  what `frm(importance =)` is worth on which designs.
* `vignette("reinforcement-learning", package = "frmtmb")` for the
  structured-family protocol these families are built on, written out
  longhand for one model.
